---
title: "Barcode Estimation"
format:
  html:
        embed-resources: false
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(ggrepel)
    library(dplyr)
    library(patchwork)
})
```

```{r}
#| label: load-data

# Discover all knee_data and nuclear_fraction CSVs staged into the work dir
knee_data_files <- list.files(".", pattern = "_knee_data\\.csv$", full.names = TRUE)
nf_files        <- list.files(".", pattern = "_nuclear_fraction\\.csv$", full.names = TRUE)

# Extract sample IDs from file names
.sample_id_from_file <- function(paths, suffix) {
    sub(suffix, "", basename(paths))
}

knee_data_ids <- .sample_id_from_file(knee_data_files, "_knee_data.csv")
nf_ids        <- .sample_id_from_file(nf_files,        "_nuclear_fraction.csv")

# Named lists keyed by sample ID (matches scprocess pattern)
knee_fs <- setNames(as.list(knee_data_files), knee_data_ids)
nf_fs   <- setNames(as.list(nf_files),        nf_ids)

# Aggregate knee params into a single data.frame (one row per sample)
# Select only the summary columns — avoids loading per-barcode data
rna_knee_params_df <- lapply(names(knee_fs), function(s) {
    dt <- fread(knee_fs[[s]], select = c("sample_id", "knee1", "shin1", "knee2", "shin2",
                                          "total_droplets_included", "expected_cells"))
    unique(dt)
}) |> rbindlist()

sample_var <- "sample_id"
all_samples <- names(knee_fs)

# Chunk samples into groups of 8 for faceted plots (matches scprocess chunking)
chunk_size <- 8
slope_ratio_split_l <- split(
    all_samples,
    ceiling(seq_along(all_samples) / chunk_size)
)
```

## Barcode rank plots

Each panel shows the UMI rank plot (log–log scale) for one sample.
Horizontal lines mark the four knee/shin estimates:

- **knee1** / **shin1**: upper boundary of cell-containing droplets (first knee and inflection)
- **knee2** / **shin2**: lower boundary of cells / upper boundary of empty droplets (second knee and inflection)

Vertical lines show:

- **expected_cells**: rank at the first inflection (starting point for cell calling)
- **empty_plateau_middle**: geometric midpoint between shin1 and knee2 (total droplets included threshold)

```{r}
#| label: plot-functions

# Ported from scprocess/scripts/mapping.R: plot_barcode_ranks_w_params()
plot_barcode_ranks_w_params <- function(knee_fs, ambient_knees_df,
                                         sample_var = "sample_id",
                                         show_lines = TRUE) {
    s_ord <- names(knee_fs)

    # Build rank data for each sample (unique UMI -> cumulative barcode rank)
    # Only load the `total` column — all other columns are unused for plotting
    knee_data <- lapply(s_ord, function(s) {
        dt <- fread(knee_fs[[s]], select = "total")
        dt[, .(n_bc = .N), by = .(lib_size = total)] |>
            setorder(-lib_size) |>
            _[, bc_rank := cumsum(n_bc)] |>
            _[, (sample_var) := s]
    }) |> rbindlist()

    knee_vars <- c(sample_var, "knee1", "shin1", "knee2", "shin2",
                   "total_droplets_included", "expected_cells")

    lines_knees <- as.data.table(ambient_knees_df)[
        get(sample_var) %in% s_ord, ..knee_vars
    ] |>
        setnames("total_droplets_included", "empty_plateau_middle") |>
        melt(id.vars = sample_var) |>
        _[, `:=`(
            axis = fifelse(
                variable %in% c("knee1", "knee2", "shin1", "shin2"), "y", "x"
            ),
            type = fifelse(
                variable %in% c("knee1", "knee2", "shin1", "shin2"),
                "knee/shin estimate",
                "cellbender parameter"
            )
        )]

    hlines <- lines_knees[axis == "y"]
    vlines <- lines_knees[axis == "x"]

    p_labels <- c("1", "10", "100", "1k", "10k", "100k", "1M")
    p_breaks <- c(1e0, 1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

    knee_data[[sample_var]] <- factor(knee_data[[sample_var]], levels = s_ord)
    if (show_lines) {
        hlines[[sample_var]] <- factor(hlines[[sample_var]], levels = s_ord)
        vlines[[sample_var]] <- factor(vlines[[sample_var]], levels = s_ord)
    }

    p <- ggplot(knee_data) +
        aes(x = bc_rank, y = lib_size) +
        geom_line(linewidth = 0.3, color = "#283747") +
        facet_wrap(~get(sample_var), ncol = 4) +
        scale_x_log10(labels = p_labels, breaks = p_breaks) +
        scale_y_log10(labels = p_labels, breaks = p_breaks) +
        scale_color_manual(
            values = c("#7c4b73", "#88a0dc"),
            breaks = c("knee/shin estimate", "cellbender parameter")
        ) +
        theme_classic(base_size = 9) +
        theme(legend.position = "none") +
        labs(x = "barcode rank", y = "library size", color = NULL)

    if (show_lines) {
        p <- p +
            geom_hline(
                data    = hlines,
                mapping = aes(yintercept = value, color = type)
            ) +
            geom_vline(
                data    = vlines,
                mapping = aes(xintercept = value, color = type)
            ) +
            geom_text_repel(
                data    = hlines,
                mapping = aes(y = value, x = 10, label = variable),
                size    = 2.5
            ) +
            geom_text_repel(
                data    = vlines,
                mapping = aes(x = value, y = 100, label = variable),
                size    = 2.5, angle = 90
            )
    }

    p
}
```

```{r}
#| label: knee-plots
#| results: asis
#| fig-width: 10
#| fig-height: 4

for (i in seq_along(slope_ratio_split_l)) {
    samples  <- slope_ratio_split_l[[i]]
    name     <- sprintf("sample chunk #%d", i)
    knee_fs_chunk <- knee_fs[samples]

    cat("####", name, "\n")
    suppressWarnings(print(
        plot_barcode_ranks_w_params(
            knee_fs_chunk,
            rna_knee_params_df,
            sample_var = sample_var
        )
    ))
    cat("\n\n")
}
```

## QC metrics: cells vs empties

Per-sample violin plots comparing UMI counts and spliced fraction between
barcodes called as **cells** (rank ≤ `expected_cells`) and **empty** droplets
(`in_empty_plateau == TRUE`). Cells should have higher UMI counts and — for
nuclear/single-nucleus data — a higher spliced fraction than empties.

```{r}
#| label: qc-metrics-cells-empties
#| fig-width: 12
#| fig-height: 5

# Build long-format table: one row per (sample, barcode, qc_metric)
# mirroring scprocess/scripts/mapping.R plot_qc_metrics_split_by_cells_empties()
umis_brks   <- log10(c(1, 10, 30, 100, 300, 1e3, 3e3, 1e4, 3e4, 1e5, 3e5, 1e6))
umis_labs   <- c("1", "10", "30", "100", "300", "1k", "3k", "10k", "30k", "100k", "300k", "1M")
splice_brks <- qlogis(c(0.01, 0.03, 0.1, 0.3, 0.5, 0.7, 0.9, 0.97, 0.99))
splice_labs <- c("1%", "3%", "10%", "30%", "50%", "70%", "90%", "97%", "99%")

cells_empties_dt <- lapply(all_samples, function(s) {
    dt <- fread(knee_fs[[s]],
                select = c("sample_id", "barcode", "rank", "total",
                           "spliced", "unspliced",
                           "expected_cells", "in_empty_plateau"))
    dt[rank <= expected_cells | in_empty_plateau == TRUE][
        , .(
            sample_id  = s,
            barcode,
            umis       = log10(total),
            splice_pct = qlogis((spliced + 1) / (spliced + unspliced + 2)),
            what       = fifelse(rank <= expected_cells, "cell", "empty")
        )
    ]
}) |> rbindlist()

cells_empties_dt[, sample_id := factor(sample_id, levels = all_samples)]

g_umis <- ggplot(cells_empties_dt[!is.na(umis)]) +
    aes(x = sample_id, y = umis, fill = what) +
    geom_violin(colour = NA, kernel = "rectangular",
                adjust = 0.1, scale = "width", width = 0.8) +
    scale_y_continuous(breaks = umis_brks, labels = umis_labs) +
    scale_fill_manual(values = c(cell = "#1965B0", empty = "grey70")) +
    coord_flip() +
    theme_classic(base_size = 9) +
    theme(legend.position = "none") +
    labs(x = NULL, y = "no. of UMIs", fill = NULL)

g_spl <- ggplot(cells_empties_dt[!is.na(splice_pct)]) +
    aes(x = sample_id, y = splice_pct, fill = what) +
    geom_violin(colour = NA, kernel = "rectangular",
                adjust = 0.1, scale = "width", width = 0.8) +
    scale_y_continuous(breaks = splice_brks, labels = splice_labs) +
    scale_fill_manual(values = c(cell = "#1965B0", empty = "grey70")) +
    coord_flip() +
    theme_classic(base_size = 9) +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    labs(x = NULL, y = "spliced pct.", fill = "barcode type")

g_umis + g_spl + plot_layout(widths = c(1.2, 1))
```

## Knee parameter summary

```{r}
#| label: knee-summary-table

knitr::kable(
    rna_knee_params_df[order(rna_knee_params_df$sample_id), ],
    caption = "Knee and shin estimates per sample",
    format.args = list(big.mark = ",")
)
```

## Nuclear fraction

Per-barcode nuclear fraction (U / (S+U)) across all samples.
High nuclear fraction indicates predominantly unspliced / nuclear RNA.

```{r}
#| label: nuclear-fraction-plot
#| fig-width: 10
#| fig-height: 4

nf_dt <- lapply(names(nf_fs), function(s) {
    dt <- fread(nf_fs[[s]], select = "nuclear_fraction")
    dt[, (sample_var) := s]
}) |> rbindlist()

nf_dt[[sample_var]] <- factor(nf_dt[[sample_var]], levels = all_samples)

ggplot(nf_dt, aes(x = nuclear_fraction, fill = get(sample_var))) +
    geom_density(alpha = 0.5, color = NA) +
    facet_wrap(~get(sample_var), ncol = 4, scales = "free_y") +
    scale_fill_viridis_d(option = "D") +
    theme_classic(base_size = 9) +
    theme(legend.position = "none") +
    labs(
        x = "nuclear fraction  [U / (S+U)]",
        y = "density",
        title = "Per-barcode nuclear fraction distribution"
    )
```

```{r}
#| label: nuclear-fraction-summary

nf_summary <- nf_dt[, .(
    n_barcodes = .N,
    median_nf  = round(median(nuclear_fraction), 3),
    mean_nf    = round(mean(nuclear_fraction), 3),
    pct_high   = round(mean(nuclear_fraction > 0.8) * 100, 1)
), by = sample_var]

knitr::kable(
    nf_summary[order(nf_summary$sample_id), ],
    col.names = c("Sample", "Barcodes", "Median NF", "Mean NF", "% NF > 0.8"),
    caption   = "Nuclear fraction summary per sample"
)
```
