---
title: "Ambient RNA removal diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  warning: false
  message: false
  cache: false
---

```{=html}
<style>
/* keep report figures from rendering oversized */
.quarto-figure img, .cell-output-display img, p > img, img.img-fluid {
  max-width: min(100%, 820px);
  height: auto;
}
</style>
```

This report summarizes decontX ambient RNA correction for each sample, including
the number of cells retained, the amount of background used for contamination
estimation, and how read composition changes before and after decontamination.

```{r setup}
knitr::opts_chunk$set(fig.path = 'plots/', dev = 'png', dpi = 110)
suppressPackageStartupMessages({
  library(data.table)
  library(magrittr)
  library(Matrix)
  library(ggplot2)
  library(patchwork)
  library(strex)
  library(DT)
})

# Discover per-sample files written by DECONTX into the Nextflow work dir
qc_fs      <- list.files(".", pattern = "^barcodes_qc_metrics_.+\\.csv\\.gz$", full.names = TRUE)
bcs_fs     <- list.files(".", pattern = "^cell_barcodes_.+\\.csv$",            full.names = TRUE)
dcx_fs     <- list.files(".", pattern = "^dcx_params_.+\\.csv$",               full.names = TRUE)
summary_fs <- list.files(".", pattern = "^dcx_summary_.+\\.csv$",              full.names = TRUE)

# Extract sample IDs from filenames
.sid <- function(fs, prefix, suffix) {
  sub(suffix, "", sub(paste0("^.*/", prefix), "", fs))
}

s_lvls <- .sid(summary_fs, "dcx_summary_", "\\.csv$")
names(qc_fs)      <- .sid(qc_fs,      "barcodes_qc_metrics_", "\\.csv\\.gz$")
names(bcs_fs)     <- .sid(bcs_fs,     "cell_barcodes_",       "\\.csv$")
names(dcx_fs)     <- .sid(dcx_fs,     "dcx_params_",          "\\.csv$")
names(summary_fs) <- .sid(summary_fs, "dcx_summary_",         "\\.csv$")

# Align to same sample order
qc_fs      <- qc_fs[s_lvls]
bcs_fs     <- bcs_fs[s_lvls]
dcx_fs     <- dcx_fs[s_lvls]
summary_fs <- summary_fs[s_lvls]

# Load data
summary_dt <- lapply(summary_fs, fread) %>% rbindlist()
ok_bcs_ls  <- lapply(bcs_fs,     function(f) fread(f, header = FALSE)$V1)
dcx_ls     <- lapply(dcx_fs,     fread)

# Filter QC metrics to barcodes with >= 10 total pre-UMIs (mirrors scprocess get_usa_dt)
usa_dt_ls  <- lapply(qc_fs, function(f) {
  dt <- fread(f)
  dt[pre_S + pre_U + pre_A >= 10]
})

n_samples  <- length(s_lvls)
chunk_size <- 10
idx_ls     <- seq_along(s_lvls)
chunks_ls  <- split(idx_ls, ceiling(idx_ls / chunk_size))
```

## Summary table

```{r summary_table}
summary_tbl <- summary_dt[, .(
  sample_id, n_cells, n_empties, expected_cells,
  total_droplets_included,
  mean_pct_contamination
)]
setnames(summary_tbl, c(
  "Sample", "Cells (post-decontX)", "Empties used",
  "Expected cells", "Total droplets included", "Mean % contamination"
))

DT::datatable(
  summary_tbl,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(
    pageLength = 10,
    lengthMenu = list(c(10), c("10")),
    dom = "tip",
    searching = TRUE,
    autoWidth = TRUE,
    order = list(list(0, "asc"))
  )
)
```

## Spliced % vs UMIs

Relationship between total UMIs and spliced read percentage, before and after
decontX. Ambient RNA is enriched for spliced reads, so effective ambient removal
shifts the spliced % of cells downward. Barcodes are coloured by whether they
were called as cells (grey), used as ambient background (red), or excluded
entirely (dark grey).

```{r plot_spliced_vs_umis, results="asis"}
cat('::: {.panel-tabset}\n\n')
.plot_spliced_vs_umis <- function(ss, usa_dt, ok_bcs, total_inc) {
  pscount  <- 10
  PCT_BRKS <- c(0.001, 0.003, 0.01, 0.03, 0.1, 0.5, 0.9, 0.97, 0.99, 0.997, 0.999) %>% qlogis()
  PCT_LABS <- c("0.1%", "0.3%", "1%", "3%", "10%", "50%", "90%", "97%", "99%", "99.7%", "99.9%")
  LOG_BRKS <- log10(pscount + c(0, 10, 30, 100, 300, 1000, 3000, 1e4, 3e4, 1e5, 3e5))
  LOG_LABS <- c("0", "10", "30", "100", "300", "1k", "3k", "10k", "30k", "100k", "300k")
  COL_VALS <- c(
    "cell"                       = "grey80",
    "no, used for\nambient"      = "#FB8072",
    "no, excluded\ncompletely"   = "grey30"
  )

  plot_dt <- usa_dt[, .(
    barcode,
    pre_total    = pre_S + pre_U + pre_A,
    post_total   = post_S + post_U + post_A,
    pre_logit_S  = qlogis((pre_S  + 1) / (pre_S  + pre_U  + 2)),
    post_logit_S = qlogis((post_S + 1) / (post_S + post_U + 2))
  )] %>%
    melt(measure.vars = measure(stage, value.name,
                                 pattern = "(pre|post)_(.+)")) %>%
    .[, stage := factor(stage, levels = c("pre", "post"))] %>%
    .[total > 0] %>%
    .[order(total)]

  # Determine which barcodes are "included" (top total_inc by pre UMI)
  inc_bcs <- usa_dt[order(-pre_S - pre_U - pre_A), barcode][seq_len(total_inc)]

  plot_dt[, status := fifelse(
    barcode %in% ok_bcs,  names(COL_VALS)[1],
    fifelse(barcode %in% inc_bcs, names(COL_VALS)[2], names(COL_VALS)[3])
  ) %>% factor(levels = names(COL_VALS))]

  g_dens <- ggplot(plot_dt) +
    aes(y = logit_S, x = log10(total + pscount)) +
    geom_bin2d(bins = c(80, 50)) +
    scale_x_continuous(breaks = LOG_BRKS, labels = LOG_LABS) +
    expand_limits(x = log10(pscount)) +
    scale_y_continuous(breaks = PCT_BRKS, labels = PCT_LABS) +
    scale_fill_distiller(palette = "RdBu", trans = "log10") +
    facet_grid(. ~ stage) +
    theme_classic() +
    labs(y = "spliced pct.", x = "total UMIs", title = ss)

  g_dots <- ggplot(plot_dt) +
    aes(y = logit_S, x = log10(total + pscount), colour = status) +
    geom_point(size = 0.1) +
    scale_x_continuous(breaks = LOG_BRKS, labels = LOG_LABS) +
    expand_limits(x = log10(pscount)) +
    scale_y_continuous(breaks = PCT_BRKS, labels = PCT_LABS) +
    scale_color_manual(values = COL_VALS,
      guide = guide_legend(override.aes = list(size = 3, alpha = 1))) +
    facet_grid(. ~ stage) +
    theme_classic() +
    labs(y = "spliced pct.", x = "total UMIs", colour = "called as cell?")

  g_dens / g_dots + plot_layout(axes = "collect")
}

for (ss in s_lvls) {
  cat("### ", ss, "\n\n")
  tot_inc <- summary_dt[sample_id == ss, total_droplets_included]
  print(.plot_spliced_vs_umis(ss, usa_dt_ls[[ss]], ok_bcs_ls[[ss]], tot_inc))
  cat("\n\n")
}
cat(':::\n\n')
```

## Reads removed as ambient

Proportion of reads removed from barcodes called as cells. A logit scale is used
so that small proportions remain readable. Values near 0% indicate little ambient
removal; values toward 99%+ may flag problematic samples.

```{r plot_reads_removed_as_ambient, results="asis"}
cat('::: {.panel-tabset}\n\n')
.plot_reads_removed_as_ambient <- function(usa_dt_ls_chunk, ok_bcs_ls_chunk) {
  logit_brks <- c(1e-4, 1e-3, 1e-2, 0.10, 0.50, 0.90, 0.99, 0.999) %>% qlogis()
  logit_labs <- c("0.01%", "0.1%", "1%", "10%", "50%", "90%", "99%", "99.9%")

  plot_dt <- lapply(names(usa_dt_ls_chunk), function(nn) {
    usa_tmp  <- usa_dt_ls_chunk[[nn]]
    bcs_tmp  <- ok_bcs_ls_chunk[[nn]]
    usa_tmp[barcode %in% bcs_tmp] %>%
      .[, .(total_pre  = pre_S  + pre_U  + pre_A,
             total_post = post_S + post_U + post_A), by = barcode] %>%
      .[!is.na(total_post)] %>%
      .[, logit_removed := qlogis(1 - total_post / total_pre)] %>%
      .[, sample_id := nn]
  }) %>% rbindlist()

  ggplot(plot_dt) +
    aes(x = sample_id,
        y = logit_removed %>%
          pmin(tail(logit_brks, 1)) %>%
          pmax(head(logit_brks, 1))) +
    geom_violin(colour = NA, fill = "grey60",
                kernel = "rectangular", adjust = 0.1,
                scale = "width", width = 0.8) +
    scale_y_continuous(breaks = logit_brks, labels = logit_labs) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0.5),
          plot.margin = unit(c(1, 4, 1, 1), "lines")) +
    labs(y = "pct. reads removed as ambient", x = NULL)
}

w_per_s <- 0.8
for (ii in seq_along(chunks_ls)) {
  chunk     <- chunks_ls[[ii]]
  chunk_ids <- s_lvls[chunk]
  cat("### samples ", min(chunk), "-", max(chunk), "\n\n")
  fig_w <- 1 + w_per_s * max(length(chunk), 4)
  cat(sprintf("<div style='width:%spx'>\n", round(fig_w * 96)))
  print(.plot_reads_removed_as_ambient(
    usa_dt_ls[chunk_ids],
    ok_bcs_ls[chunk_ids]
  ))
  cat("</div>\n\n")
}
cat(':::\n\n')
```

## R session info

<details>
```{r session_info}
sessionInfo()
```
</details>
