---
title: "Annotation diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
source("annotation_utils.R")
source("integration_plots.R")

sel_res <- Sys.getenv("ANNOTATION_SEL_RES", "0.2")
min_cpm_mkr <- as.numeric(Sys.getenv("ANNOTATION_MIN_CPM_MKR", "50"))
not_ok_re <- Sys.getenv("ANNOTATION_NOT_OK_RE", "(lincRNA|lncRNA|pseudogene|antisense)")
top_n <- as.integer(Sys.getenv("ANNOTATION_TOP_N", "10"))
fdr_cut <- as.numeric(Sys.getenv("ANNOTATION_FDR_CUT", "0.01"))
max_zero_p <- as.numeric(Sys.getenv("ANNOTATION_MAX_ZERO_P", "0.5"))
```

```{r load_data}
int_dt <- fread("integration_dt.csv.gz")
mkrs_dt <- fread("annotation_markers.csv.gz")
cpms_dt <- fread("annotation_logcpms.csv.gz")
panel_dt <- fread("annotation_marker_panel.csv.gz")
marker_expr_rds <- "annotation_marker_expression.rds"
cell_labels_f <- "annotation_cell_labels.csv.gz"

cl_col <- paste0("RNA_snn_res.", sel_res)
stopifnot("Selected annotation resolution not found in integration output" = cl_col %in% names(int_dt))

int_dt <- int_dt[!is.na(get(cl_col)) & sample_id != "", .(
  sample_id,
  cell_id,
  UMAP1,
  UMAP2,
  cluster = get(cl_col)
)]

cl_ord <- order_cluster_labels(unique(int_dt$cluster))
int_dt[, cluster := factor(as.character(cluster), levels = cl_ord)]
panel_label_levels <- unique(as.character(panel_dt$label))
panel_dt[, label := factor(as.character(label), levels = panel_label_levels)]

top_min_dt <- mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][
  logcpm.sel >= log(min_cpm_mkr + 1)
] |> get_top_markers(fdr_cut = fdr_cut, n_top = top_n, max_zero_p = max_zero_p)

top_heatmap_panel_dt <- make_top_marker_heatmap_panel(top_min_dt, cl_order = cl_ord)

top50_dt <- mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][
  logFC > 0 & logcpm.sel >= log(min_cpm_mkr + 1)
][order(cluster, FDR, -logFC, -logcpm.sel, symbol)]
top50_dt[, prop_zero := n_zero / n_cl]
top50_dt[, marker_rank := seq_len(.N), by = cluster]
top50_dt <- top50_dt[marker_rank <= 50]

marker_coverage_dt <- panel_dt[, .(
  markers = paste(unique(symbol), collapse = ", ")
), by = label]

if (file.exists(cell_labels_f)) {
  cell_labels_dt <- fread(cell_labels_f)
  cluster_label_dt <- unique(cell_labels_dt[, .(cluster, label, label_score, n_markers)])
  cell_label_lu <- unique(cell_labels_dt[, .(cell_id, label, label_score, n_markers)])
  label_lu <- setNames(as.character(cell_label_lu$label), cell_label_lu$cell_id)
  label_score_lu <- setNames(cell_label_lu$label_score, cell_label_lu$cell_id)
  n_markers_lu <- setNames(cell_label_lu$n_markers, cell_label_lu$cell_id)
  int_dt[, label := label_lu[cell_id]]
  int_dt[, label_score := label_score_lu[cell_id]]
  int_dt[, n_markers := n_markers_lu[cell_id]]
} else {
  cluster_label_dt <- assign_cluster_labels_from_panel(mkrs_dt, panel_dt, min_cpm = min_cpm_mkr)
  cluster_label_dt <- merge(data.table(cluster = cl_ord), cluster_label_dt, by = "cluster", all.x = TRUE, sort = FALSE)
  label_lu <- setNames(as.character(cluster_label_dt$label), as.character(cluster_label_dt$cluster))
  label_score_lu <- setNames(cluster_label_dt$label_score, as.character(cluster_label_dt$cluster))
  n_markers_lu <- setNames(cluster_label_dt$n_markers, as.character(cluster_label_dt$cluster))
  int_dt[, label := label_lu[as.character(cluster)]]
  int_dt[, label_score := label_score_lu[as.character(cluster)]]
  int_dt[, n_markers := n_markers_lu[as.character(cluster)]]
}
cluster_label_dt <- merge(data.table(cluster = cl_ord), cluster_label_dt, by = "cluster", all.x = TRUE, sort = FALSE)
cluster_label_dt[is.na(label), `:=`(label = "Unassigned", label_score = NA_real_, n_markers = 0L)]
cluster_label_dt[, label := factor(as.character(label), levels = unique(c(panel_label_levels, "Unassigned")))]

int_dt[is.na(label), label := "Unassigned"]
int_dt[, label := factor(as.character(label), levels = unique(c(panel_label_levels, "Unassigned")))]

panel_ls <- setNames(
  lapply(panel_label_levels, function(label_name) panel_dt[as.character(label) == label_name]),
  panel_label_levels
)

cell_exp_dt <- data.table()
```

## Annotation setup

This report assigns labels from the integration clustering column `r cl_col`, then tests marker genes by aggregating counts per sample within each cluster and running one-vs-rest pseudobulk differential expression.

* **Selected cluster resolution used for annotation:** `r sel_res` (`r cl_col`)
* **Clusters carried into annotation:** `r length(cl_ord)`
* **Cells used for annotation:** `r format(nrow(int_dt), big.mark = ",")`
* **Provided marker genes matched in the data:** `r uniqueN(panel_dt$gene_id)`
* **Marker-expression cache available for UMAPs:** `r if (file.exists(marker_expr_rds)) "yes" else "no"`

## Input marker panel

```{r marker_panel_table}
DT::datatable(
  marker_coverage_dt,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", paging = FALSE, searching = FALSE, info = FALSE, ordering = FALSE, autoWidth = TRUE)
)
```

## Heatmaps of marker genes

::: {.panel-tabset}

### Provided markers

```{r plot_heatmap_provided_markers, fig.height = 9, fig.width = 10}
suppressMessages(draw(
  plot_heatmap_of_selected_genes(mkrs_dt, panel_dt),
  heatmap_legend_side = "bottom",
  annotation_legend_side = "bottom",
  merge_legend = TRUE
))
```

### Top markers by cluster

```{r plot_heatmap_top_markers, fig.height = 9, fig.width = 10}
if (nrow(top_heatmap_panel_dt) > 0) {
  suppressMessages(draw(
    plot_heatmap_of_selected_genes(mkrs_dt, top_heatmap_panel_dt),
    heatmap_legend_side = "bottom",
    annotation_legend_side = "bottom",
    merge_legend = TRUE
  ))
} else {
  cat("No top markers passed the display filters for the heatmap.\n")
}
```

:::

## Top marker genes by cluster at resolution `r sel_res`

```{r plot_top_markers, fig.height = 9, fig.width = 10, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (sel_cl in cl_ord) {
  cat('###', sel_cl, '\n\n')
  if (sel_cl %in% top_min_dt$cluster) {
    print(plot_top_marker_genes(sel_cl, top_min_dt, cpms_dt, cl_order = cl_ord))
  } else {
    cat('No marker genes passed the display filters for this cluster.\n\n')
  }
  cat('\n\n')
}
cat(':::\n\n')
```

## Provided marker genes by label

```{r plot_provided_marker_panels, fig.height = 9, fig.width = 10, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (label_name in names(panel_ls)) {
  cat('###', label_name, '\n\n')
  sel_panel_dt <- panel_ls[[label_name]]
  highlight_clusters <- as.character(cluster_label_dt[as.character(label) == label_name, cluster])
  print(plot_selected_marker_panel(
    sel_panel_dt,
    cpms_dt,
    cl_order = cl_ord,
    highlight_clusters = highlight_clusters,
    title = sprintf('%s provided marker genes', label_name)
  ))
  cat('\n\n')
}
cat(':::\n\n')
```

## Annotation labels over UMAP

Cell labels in the panels below are assigned at the cluster level by the provided marker label with the highest mean positive marker log2 fold-change among the matched input markers. Cluster labels are written at each cluster centroid; if two clusters receive the same annotation label, that label is repeated once per cluster.

```{r plot_annotation_umap_overview, fig.height = 7, fig.width = 14}
g_label <- plot_umap_cluster(
  int_dt[, .(cell_id, UMAP1, UMAP2)],
  int_dt[, .(cell_id, cluster = label, centroid_group = as.character(cluster), centroid_label = as.character(label))],
  name = 'cell label'
)
g_cluster <- plot_umap_cluster(
  int_dt[, .(cell_id, UMAP1, UMAP2)],
  int_dt[, .(cell_id, cluster)],
  name = sprintf('cluster (res. %s)', sel_res)
)
print(g_label + g_cluster + patchwork::plot_layout(widths = c(1, 1)))
```

## Top 50 marker tables by cluster at resolution `r sel_res`

The tables below list up to 50 positive markers per cluster after excluding genes that match `r not_ok_re` and requiring expression above the annotation CPM threshold. Each table shows 25 rows per page.

```{r top_marker_tables, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (sel_cl in cl_ord) {
  cat('###', sel_cl, '\n\n')
  sel_tbl <- top50_dt[cluster == sel_cl, .(
    rank = marker_rank,
    symbol,
    gene_type,
    logFC = round(logFC, 2),
    FDR = signif(FDR, 3),
    logcpm_sel = round(logcpm.sel, 2),
    logcpm_other = round(logcpm.other, 2),
    zero_prop = round(prop_zero, 3)
  )]
  if (nrow(sel_tbl) > 0) {
    print(DT::datatable(
      sel_tbl,
      rownames = FALSE,
      class = 'compact stripe hover',
      options = list(
        pageLength = 25,
        lengthMenu = list(c(10, 25), c('10', '25')),
        dom = 'tip',
        autoWidth = TRUE,
        scrollX = TRUE
      )
    ))
  } else {
    cat('No marker genes passed the table filters for this cluster.\n\n')
  }
  cat('\n\n')
}
cat(':::\n\n')
```

## Marker expression over UMAP

Marker-expression panels below use the cached annotation marker-expression file and colour cells by uncapped log CPM values rather than per-gene values rescaled to 1.

```{r plot_marker_umaps, fig.height = 7, fig.width = 14, results = 'asis'}
cell_exp_dt <- if (file.exists(marker_expr_rds)) readRDS(marker_expr_rds) else data.table()

if (nrow(cell_exp_dt) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (label_name in names(panel_ls)) {
    cat('###', label_name, '\n\n')
    sel_expr_dt <- cell_exp_dt[as.character(label) == label_name]
    if (nrow(sel_expr_dt) == 0) {
      cat('No marker expression values were available for this label.\n\n')
      next
    }

    gene_levels <- unique(as.character(sel_expr_dt$symbol))
    for (gene_name in gene_levels) {
      cat('####', gene_name, '\n\n')
      print(plot_expression_umap_pair(
        sel_expr_dt[as.character(symbol) == gene_name, .(cell_id, expr)],
        int_dt[, .(cell_id, UMAP1, UMAP2, cluster)],
        cluster_col = 'cluster',
        gene_name = gene_name,
        cluster_name = sprintf('cluster (res. %s)', sel_res)
      ))
      cat('\n\n')
    }
  }
  cat(':::\n\n')
} else {
  cat('No filtered H5 inputs were available to render marker expression UMAPs.\n\n')
}
```

## Session info

```{r session_info}
sessionInfo()
```