---
title: "Annotation diagnostics"
format:
  html:
    embed-resources: false
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
source("annotation_utils.R")

sel_res <- Sys.getenv("ANNOTATION_SEL_RES", "0.2")
min_cpm_mkr <- as.numeric(Sys.getenv("ANNOTATION_MIN_CPM_MKR", "50"))
not_ok_re <- Sys.getenv("ANNOTATION_NOT_OK_RE", "(lincRNA|lncRNA|pseudogene|antisense)")
top_n <- as.integer(Sys.getenv("ANNOTATION_TOP_N", "10"))
fdr_cut <- as.numeric(Sys.getenv("ANNOTATION_FDR_CUT", "0.01"))
max_zero_p <- as.numeric(Sys.getenv("ANNOTATION_MAX_ZERO_P", "0.5"))
```

```{r load_data}
int_dt <- fread("integration_dt.csv.gz")
mkrs_dt <- fread("annotation_markers.csv.gz")
cpms_dt <- fread("annotation_logcpms.csv.gz")
panel_dt <- fread("annotation_marker_panel.csv.gz")
h5_files <- Sys.glob("filt_counts_*.h5")

cl_col <- paste0("RNA_snn_res.", sel_res)
stopifnot("Selected annotation resolution not found in integration output" = cl_col %in% names(int_dt))

int_dt <- int_dt[!is.na(get(cl_col)) & sample_id != "", .(
  sample_id,
  cell_id,
  UMAP1,
  UMAP2,
  cluster = get(cl_col)
)]

cl_ord <- order_cluster_labels(unique(int_dt$cluster))
int_dt[, cluster := factor(as.character(cluster), levels = cl_ord)]
panel_label_levels <- unique(as.character(panel_dt$label))
panel_dt[, label := factor(as.character(label), levels = panel_label_levels)]

top_min_dt <- mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][
  logcpm.sel >= log(min_cpm_mkr + 1)
] |> get_top_markers(fdr_cut = fdr_cut, n_top = top_n, max_zero_p = max_zero_p)

top_heatmap_panel_dt <- make_top_marker_heatmap_panel(top_min_dt, cl_order = cl_ord)

marker_coverage_dt <- panel_dt[, .(
  markers = paste(unique(symbol), collapse = ", ")
), by = label]

cluster_label_dt <- assign_cluster_labels_from_panel(mkrs_dt, panel_dt, min_cpm = min_cpm_mkr)
cluster_label_dt <- merge(data.table(cluster = cl_ord), cluster_label_dt, by = "cluster", all.x = TRUE, sort = FALSE)
cluster_label_dt[is.na(label), `:=`(label = "Unassigned", label_score = NA_real_, n_markers = 0L)]
cluster_label_dt[, label := factor(as.character(label), levels = unique(c(panel_label_levels, "Unassigned")))]

int_dt <- merge(int_dt, cluster_label_dt[, .(cluster, label)], by = "cluster", all.x = TRUE, sort = FALSE)
int_dt[is.na(label), label := "Unassigned"]
int_dt[, label := factor(as.character(label), levels = unique(c(panel_label_levels, "Unassigned")))]

panel_ls <- setNames(
  lapply(panel_label_levels, function(label_name) panel_dt[as.character(label) == label_name]),
  panel_label_levels
)

cell_exp_dt <- data.table()
if (length(h5_files) > 0 && nrow(panel_dt) > 0) {
  cell_exp_dt <- load_h5_marker_expression(
    h5_files,
    panel_dt,
    int_dt[, .(sample_id, cell_id, UMAP1, UMAP2)],
    pseudo_count = 10
  )
}
```

Marker genes were tested by aggregating counts per sample within each cluster at resolution `r sel_res`, then running one-vs-rest pseudobulk differential expression across clusters.

* **Clusters carried into annotation:** `r length(cl_ord)`
* **Cells used for annotation:** `r format(nrow(int_dt), big.mark = ",")`
* **Provided marker genes matched in the data:** `r uniqueN(panel_dt$gene_id)`
* **Filtered H5 files available for marker UMAPs:** `r length(h5_files)`

## Input marker panel

```{r marker_panel_table}
DT::datatable(
  marker_coverage_dt,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", paging = FALSE, searching = FALSE, info = FALSE, ordering = FALSE, autoWidth = TRUE)
)
```

## Heatmaps of marker genes {.tabset}

### Provided markers

```{r plot_heatmap_provided_markers, fig.height = 9, fig.width = 10}
suppressMessages(draw(
  plot_heatmap_of_selected_genes(mkrs_dt, panel_dt),
  heatmap_legend_side = "bottom",
  annotation_legend_side = "bottom",
  merge_legend = TRUE
))
```

### Top markers by cluster

```{r plot_heatmap_top_markers, fig.height = 9, fig.width = 10}
if (nrow(top_heatmap_panel_dt) > 0) {
  suppressMessages(draw(
    plot_heatmap_of_selected_genes(mkrs_dt, top_heatmap_panel_dt),
    heatmap_legend_side = "bottom",
    annotation_legend_side = "bottom",
    merge_legend = TRUE
  ))
} else {
  cat("No top markers passed the display filters for the heatmap.\n")
}
```

## Top marker genes by cluster

```{r plot_top_markers, fig.height = 9, fig.width = 10, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (sel_cl in cl_ord) {
  cat('###', sel_cl, '\n\n')
  if (sel_cl %in% top_min_dt$cluster) {
    print(plot_top_marker_genes(sel_cl, top_min_dt, cpms_dt, cl_order = cl_ord))
  } else {
    cat('No marker genes passed the display filters for this cluster.\n\n')
  }
  cat('\n\n')
}
cat(':::\n\n')
```

## Provided marker genes by label

```{r plot_provided_marker_panels, fig.height = 9, fig.width = 10, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (label_name in names(panel_ls)) {
  cat('###', label_name, '\n\n')
  sel_panel_dt <- panel_ls[[label_name]]
  highlight_clusters <- as.character(cluster_label_dt[as.character(label) == label_name, cluster])
  print(plot_selected_marker_panel(
    sel_panel_dt,
    cpms_dt,
    cl_order = cl_ord,
    highlight_clusters = highlight_clusters,
    title = sprintf('%s provided marker genes', label_name)
  ))
  cat('\n\n')
}
cat(':::\n\n')
```

## Annotation labels over UMAP

Cell labels in the panels below are assigned at the cluster level by the provided marker label with the highest mean positive marker log2 fold-change among the matched input markers.

```{r plot_annotation_umap_overview, fig.height = 10, fig.width = 18}
g_label <- plot_umap_cluster(
  int_dt[, .(cell_id, UMAP1, UMAP2)],
  int_dt[, .(cell_id, cluster = label)],
  name = 'cell label'
)
g_cluster <- plot_umap_cluster(
  int_dt[, .(cell_id, UMAP1, UMAP2)],
  int_dt[, .(cell_id, cluster)],
  name = 'cluster'
)
print(g_label + g_cluster)
```

## Marker panels over UMAP

```{r plot_marker_umaps, fig.height = 10, fig.width = 18, results = 'asis'}
if (nrow(cell_exp_dt) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (label_name in names(panel_ls)) {
    cat('###', label_name, '\n\n')
    sel_expr_dt <- cell_exp_dt[as.character(label) == label_name]
    if (nrow(sel_expr_dt) == 0) {
      cat('No marker expression values were available for this label.\n\n')
      next
    }

    g_label <- plot_umap_cluster(
      int_dt[, .(cell_id, UMAP1, UMAP2)],
      int_dt[, .(cell_id, cluster = label)],
      name = 'cell label'
    )
    g_expr <- plot_selected_genes_umap(sel_expr_dt)
    print(g_label + g_expr + patchwork::plot_layout(widths = c(1, 1.8)))
    cat('\n\n')
  }
  cat(':::\n\n')
} else {
  cat('No filtered H5 inputs were available to render marker expression UMAPs.\n\n')
}
```

## Session info

```{r session_info}
sessionInfo()
```