---
title: "Annotation diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
  # Memory optimization: render plots to PNG immediately (rasterize large ComplexHeatmap objects)
  fig.keep: 'high'
  # Garbage collect aggressively in the report
  include: true
---

```{r setup}
source("annotation_utils.R")
source("integration_plots.R")

# Helper: load logCPMs for specific genes only (memory-efficient)
load_logcpms_subset <- function(cpms_path, genes_of_interest, verbose = FALSE) {
  genes_of_interest <- unique(genes_of_interest[!is.na(genes_of_interest) & nzchar(genes_of_interest)])
  if (length(genes_of_interest) == 0) {
    return(data.table())
  }

  if (verbose) cat("Loading logCPMs for", length(genes_of_interest), "genes...\n")

  genes_f <- tempfile(fileext = ".txt")
  on.exit(unlink(genes_f), add = TRUE)
  fwrite(data.table(gene_id = genes_of_interest), genes_f, col.names = FALSE)

  awk_script <- paste(
    "NR==FNR { keep[$1]=1; next }",
    "FNR==1 || ($1 in keep)"
  )
  cmd_inner <- sprintf(
    "set -euo pipefail; gzip -dc %s | awk -F',' %s %s -",
    shQuote(cpms_path),
    shQuote(awk_script),
    shQuote(genes_f)
  )
  cmd <- sprintf("bash -lc %s", shQuote(cmd_inner))

  fread(cmd = cmd)
}

infer_annotation_sel_res <- function(int_dt, mkrs_dt, cell_labels_f) {
  cluster_cols <- grep("^RNA_snn_res\\.", names(int_dt), value = TRUE)
  if (length(cluster_cols) == 0) {
    stop("No RNA_snn_res.* columns found in integration output")
  }

  artifact_sets <- list(
    marker_stats = order_cluster_labels(unique(mkrs_dt$cluster))
  )
  if (file.exists(cell_labels_f)) {
    artifact_sets$cell_labels <- order_cluster_labels(unique(fread(cell_labels_f, select = "cluster")$cluster))
  }
  artifact_sets <- artifact_sets[vapply(artifact_sets, length, integer(1)) > 0]

  if (length(artifact_sets) == 0) {
    return(data.table(sel_res = character(), matched_sources = character()))
  }

  match_dt <- rbindlist(lapply(cluster_cols, function(cluster_col) {
    int_clusters <- order_cluster_labels(unique(int_dt[[cluster_col]]))
    matched_sources <- names(artifact_sets)[vapply(artifact_sets, function(x) identical(x, int_clusters), logical(1))]
    if (length(matched_sources) == 0) {
      return(NULL)
    }
    data.table(
      sel_res = sub("^RNA_snn_res\\.", "", cluster_col),
      matched_sources = paste(matched_sources, collapse = ", ")
    )
  }), use.names = TRUE, fill = TRUE)

  unique(match_dt, by = "sel_res")
}

annotation_sel_res_env <- trimws(Sys.getenv("ANNOTATION_SEL_RES", ""))
sel_res <- annotation_sel_res_env
annotation_sel_res_source <- if (nzchar(annotation_sel_res_env)) "environment" else "unset"
min_cpm_mkr <- as.numeric(Sys.getenv("ANNOTATION_MIN_CPM_MKR", "50"))
not_ok_re <- Sys.getenv("ANNOTATION_NOT_OK_RE", "(lincRNA|lncRNA|pseudogene|antisense)")
top_n <- as.integer(Sys.getenv("ANNOTATION_TOP_N", "10"))
top_n_plot <- min(top_n, 5L)
fdr_cut <- as.numeric(Sys.getenv("ANNOTATION_FDR_CUT", "0.01"))
max_zero_p <- as.numeric(Sys.getenv("ANNOTATION_MAX_ZERO_P", "0.5"))

mem_log_f <- "annotation_memory_usage.txt"
writeLines("timestamp\tstage\tr_used_mb", con = mem_log_f)
log_mem <- function(stage) {
  capture.output(gc_tab <- gc(verbose = FALSE))
  used_mb <- sum(gc_tab[, 2])
  cat(sprintf("%s\t%s\t%.2f\n", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), stage, used_mb),
      file = mem_log_f, append = TRUE)
}
```

```{r load_data}
int_dt <- fread("integration_dt.csv.gz")
mkrs_dt <- fread("annotation_markers.csv.gz")
# Don't load full logCPMs yet; load on-demand per section
cpms_path <- "annotation_logcpms.csv.gz"
cpms_dt <- NULL
panel_dt <- fread("annotation_marker_panel.csv.gz")
marker_expr_rds <- "annotation_marker_expression.rds"
top_marker_expr_rds <- "annotation_top_marker_expression.rds"
cell_labels_f <- "annotation_cell_labels.csv.gz"

annotation_res_matches <- infer_annotation_sel_res(int_dt, mkrs_dt, cell_labels_f)
if (nzchar(annotation_sel_res_env)) {
  if (nrow(annotation_res_matches) > 0 && !annotation_sel_res_env %in% annotation_res_matches$sel_res) {
    stop(sprintf(
      paste(
        "ANNOTATION_SEL_RES=%s, but the annotation report artifacts match resolution(s) %s.",
        "This report would mix clusterings; re-render with the correct environment or regenerate the annotation report inputs."
      ),
      annotation_sel_res_env,
      paste(annotation_res_matches$sel_res, collapse = ", ")
    ))
  }
} else {
  if (nrow(annotation_res_matches) == 1) {
    sel_res <- annotation_res_matches$sel_res[[1]]
    annotation_sel_res_source <- sprintf("inferred from %s", annotation_res_matches$matched_sources[[1]])
  } else if (nrow(annotation_res_matches) > 1) {
    stop(sprintf(
      paste(
        "ANNOTATION_SEL_RES is unset and the annotation report artifacts match multiple resolutions: %s.",
        "Set ANNOTATION_SEL_RES explicitly before rendering this report."
      ),
      paste(annotation_res_matches$sel_res, collapse = ", ")
    ))
  } else {
    stop(
      paste(
        "ANNOTATION_SEL_RES is unset and the report could not infer a unique annotation resolution from annotation_markers.csv.gz",
        "and annotation_cell_labels.csv.gz. Set ANNOTATION_SEL_RES explicitly before rendering this report."
      )
    )
  }
}

cl_col <- paste0("RNA_snn_res.", sel_res)
stopifnot("Selected annotation resolution not found in integration output" = cl_col %in% names(int_dt))

# Keep only cells that were eligible for annotation (exclude doublets / doublet-enriched clusters)
keep_ann_idx <- !is.na(int_dt[[cl_col]]) & int_dt$sample_id != ""
if ("is_dbl" %in% names(int_dt)) {
  keep_ann_idx <- keep_ann_idx & !as.logical(int_dt$is_dbl)
}
if ("in_dbl_cl" %in% names(int_dt)) {
  keep_ann_idx <- keep_ann_idx & !as.logical(int_dt$in_dbl_cl)
}

# Save sample-level metadata columns before subsetting (used for composition barcharts)
core_int_re <- "^(cell_id|UMAP1|UMAP2|embedding|dbl_|is_dbl|is_doublet|RNA_snn_res\\.|leiden_)"
meta_var_cols <- setdiff(names(int_dt)[!grepl(core_int_re, names(int_dt))], "sample_id")
sample_meta_dt <- if (length(meta_var_cols) > 0) {
  unique(int_dt[keep_ann_idx, c("sample_id", meta_var_cols), with = FALSE], by = "sample_id")
} else {
  data.table(sample_id = unique(int_dt[keep_ann_idx, sample_id]))
}

int_dt <- int_dt[keep_ann_idx, .(
  sample_id,
  cell_id,
  UMAP1,
  UMAP2,
  cluster = get(cl_col)
)]

cl_ord <- order_cluster_labels(unique(int_dt$cluster))
int_dt[, cluster := factor(as.character(cluster), levels = cl_ord)]
panel_label_levels <- unique(as.character(panel_dt$label))
panel_dt[, label := factor(as.character(label), levels = panel_label_levels)]

top_min_dt <- mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][
  logcpm.sel >= log(min_cpm_mkr + 1)
] |> get_top_markers(fdr_cut = fdr_cut, n_top = top_n_plot, max_zero_p = max_zero_p)

top_heatmap_panel_dt <- make_top_marker_heatmap_panel(top_min_dt, cl_order = cl_ord)

top50_dt <- mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][
  logFC > 0 & logcpm.sel >= log(min_cpm_mkr + 1)
][order(cluster, FDR, -logFC, -logcpm.sel, symbol)]
top50_dt[, prop_zero := n_zero / n_cl]
top50_dt[, marker_rank := seq_len(.N), by = cluster]
top50_dt <- top50_dt[marker_rank <= 50]

marker_coverage_dt <- panel_dt[, .(
  markers = paste(unique(symbol), collapse = ", ")
), by = label]

if (file.exists(cell_labels_f)) {
  cell_labels_dt <- fread(cell_labels_f)
  cluster_label_dt <- unique(cell_labels_dt[, .(cluster, label, label_score, n_markers)])
  cell_label_lu <- unique(cell_labels_dt[, .(cell_id, label, label_score, n_markers)])
  label_lu <- setNames(as.character(cell_label_lu$label), cell_label_lu$cell_id)
  label_score_lu <- setNames(cell_label_lu$label_score, cell_label_lu$cell_id)
  n_markers_lu <- setNames(cell_label_lu$n_markers, cell_label_lu$cell_id)
  int_dt[, label := label_lu[cell_id]]
  int_dt[, label_score := label_score_lu[cell_id]]
  int_dt[, n_markers := n_markers_lu[cell_id]]
} else {
  cluster_label_dt <- assign_cluster_labels_from_panel(mkrs_dt, panel_dt, min_cpm = min_cpm_mkr)
  cluster_label_dt <- merge(data.table(cluster = cl_ord), cluster_label_dt, by = "cluster", all.x = TRUE, sort = FALSE)
  label_lu <- setNames(as.character(cluster_label_dt$label), as.character(cluster_label_dt$cluster))
  label_score_lu <- setNames(cluster_label_dt$label_score, as.character(cluster_label_dt$cluster))
  n_markers_lu <- setNames(cluster_label_dt$n_markers, as.character(cluster_label_dt$cluster))
  int_dt[, label := label_lu[as.character(cluster)]]
  int_dt[, label_score := label_score_lu[as.character(cluster)]]
  int_dt[, n_markers := n_markers_lu[as.character(cluster)]]
}
cluster_label_dt <- merge(data.table(cluster = cl_ord), cluster_label_dt, by = "cluster", all.x = TRUE, sort = FALSE)
cluster_label_dt[is.na(label), `:=`(label = "Unassigned", label_score = NA_real_, n_markers = 0L)]
cluster_label_dt[, label := factor(as.character(label), levels = unique(c(panel_label_levels, "Unassigned")))]

int_dt[is.na(label), label := "Unassigned"]
int_dt[, label := factor(as.character(label), levels = unique(c(panel_label_levels, "Unassigned")))]

panel_ls <- setNames(
  lapply(panel_label_levels, function(label_name) panel_dt[as.character(label) == label_name]),
  panel_label_levels
)

provided_marker_fig_height <- max(
  2.7,
  0.35 * max(1L, max(vapply(panel_ls, function(dt) data.table::uniqueN(dt$symbol), integer(1))))
)

# Determine which genes will actually be used in plots
genes_for_plots <- unique(c(
  panel_dt$gene_id,
  mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][logcpm.sel >= log(min_cpm_mkr + 1)]$gene_id,
  top_min_dt$gene_id
))

# Load logCPMs only for genes that will be displayed
cpms_dt <- load_logcpms_subset(cpms_path, genes_for_plots, verbose = FALSE)

cell_exp_dt <- data.table()

# Clean up after initial data preparation
invisible(gc(verbose = FALSE))
```

## Annotation setup

This report assigns labels from the integration clustering column `r cl_col`, then tests marker genes by aggregating counts per sample within each cluster and running one-vs-rest pseudobulk differential expression.

* **Selected cluster resolution used for annotation:** `r sel_res` (`r cl_col`)
* **Resolution source:** `r annotation_sel_res_source`
* **Clusters carried into annotation:** `r length(cl_ord)`
* **Cells used for annotation (doublets and doublet-enriched clusters removed):** `r format(nrow(int_dt), big.mark = ",")`
* **Provided marker genes matched in the data:** `r uniqueN(panel_dt$gene_id)`
* **Marker-expression cache available for UMAPs:** `r if (file.exists(marker_expr_rds)) "yes" else "no"`

## Input marker panel

```{r marker_panel_table}
DT::datatable(
  marker_coverage_dt,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", paging = FALSE, searching = FALSE, info = FALSE, ordering = FALSE, autoWidth = TRUE)
)
```

## Heatmaps of marker genes


Heatmap stars indicate positive-marker significance from the differential expression step: `***` FDR <= 0.001, `**` FDR <= 0.01, and `*` FDR <= 0.05. Blank cells are either not significant, negative logFC, or below the minimum CPM threshold.


::: {.panel-tabset}

### Provided markers

```{r plot_heatmap_provided_markers, fig.height = 7, fig.width = 7}
suppressMessages(draw(
  plot_heatmap_of_selected_genes(mkrs_dt, panel_dt),
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legend = TRUE
))
rm(panel_dt); invisible(gc(verbose = FALSE))
```

### Top markers by cluster

```{r plot_heatmap_top_markers, fig.height = 7, fig.width = 14}
if (nrow(top_heatmap_panel_dt) > 0) {
  suppressMessages(draw(
    plot_heatmap_of_selected_genes(mkrs_dt, top_heatmap_panel_dt),
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    merge_legend = TRUE
  ))
  rm(top_heatmap_panel_dt); invisible(gc(verbose = FALSE))
} else {
  cat("No top markers passed the display filters for the heatmap.\n")
}
```

:::

## Top marker genes by cluster at resolution `r sel_res`


```{r plot_top_markers, fig.height = 5, fig.width = 7, results = 'asis', cache = FALSE}
top_marker_expr_dt <- if (file.exists(top_marker_expr_rds)) readRDS(top_marker_expr_rds) else data.table()

cat('::: {.panel-tabset}\n\n')
for (sel_cl in cl_ord) {
  cat(sprintf("### %s\n\n", sel_cl))
  has_markers <- sel_cl %in% as.character(top_min_dt$cluster)
  if (!has_markers) {
    cat("No marker genes passed the display filters for this cluster.\n\n")
    next
  }

  p_dot <- plot_top_marker_genes(sel_cl, top_min_dt, cpms_dt, cl_order = cl_ord) +
    theme(
      axis.text.x = element_text(size = 9),
      axis.text.y = element_text(size = 9),
      strip.text = element_text(size = 9),
      plot.title = element_text(size = 12),legend.position="none"
    )
  g_umap <- NULL
  if (nrow(top_marker_expr_dt) > 0) {
    g_umap <- plot_top_marker_umap_facet(
      sel_cl, top_min_dt, top_marker_expr_dt,
      int_dt[, .(cell_id, UMAP1, UMAP2)]
    ) +
      theme(
        strip.text = element_text(size = 9, margin = margin(b = 4)),
        plot.title = element_text(size = 11)
      )
  }

  print(p_dot)

  if (!is.null(g_umap)) {
    print(g_umap)
  }
  cat("\n\n")
  invisible(gc(verbose = FALSE))
}
rm(top_marker_expr_dt)
cat(':::\n\n')
```

## Top 50 marker tables by cluster at resolution `r sel_res`

The tables below list up to 50 positive markers per cluster after excluding genes that match `r not_ok_re` and requiring expression above the annotation CPM threshold. Each table shows 25 rows per page.

```{r top_marker_tables, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (sel_cl in cl_ord) {
  cat(sprintf("### %s\n\n", sel_cl))
  sel_tbl <- top50_dt[as.character(cluster) == as.character(sel_cl), .(
    rank = marker_rank,
    symbol,
    gene_type,
    logFC = round(logFC, 2),
    FDR = formatC(FDR, format = 'e', digits = 2),
    logcpm_sel = round(logcpm.sel, 2),
    logcpm_other = round(logcpm.other, 2),
    zero_prop = round(prop_zero, 3)
  )]
  if (nrow(sel_tbl) > 0) {
    print(htmltools::tagList(DT::datatable(
      sel_tbl,
      rownames = FALSE,
      class = 'compact stripe hover',
      options = list(
        pageLength = 25,
        lengthMenu = list(c(10, 25), c('10', '25')),
        dom = 'tip',
        autoWidth = TRUE,
        scrollX = TRUE
      )
    )))
  } else {
    cat('No marker genes passed the table filters for this cluster.\n\n')
  }
  cat('\n\n')
}
cat(':::\n\n')
```

## Provided marker genes by label

```{r plot_provided_marker_panels, fig.height = provided_marker_fig_height, fig.width = 7, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (i_panel in seq_along(names(panel_ls))) {
  label_name <- names(panel_ls)[i_panel]
  cat('###', label_name, '\n\n')
  sel_panel_dt <- panel_ls[[label_name]]
  highlight_clusters <- as.character(cluster_label_dt[as.character(label) == label_name, cluster])
  print(plot_selected_marker_panel(
    sel_panel_dt,
    cpms_dt,
    cl_order = cl_ord,
    highlight_clusters = highlight_clusters,
    title = sprintf('%s provided marker genes', label_name)
  )+theme(legend.position = "none"))
  cat('\n\n')
  # Clean up every 3 panels
  if (i_panel %% 3 == 0) invisible(gc(verbose = FALSE))
}
cat(':::\n\n')
```


## Marker expression over UMAP

Marker-expression panels below use the cached annotation marker-expression file and colour cells by uncapped log CPM values rather than per-gene values rescaled to 1.

```{r plot_marker_umaps, fig.height = 4.5, fig.width = 7, results = 'asis'}
cell_exp_dt <- if (file.exists(marker_expr_rds)) readRDS(marker_expr_rds) else data.table()

if (nrow(cell_exp_dt) > 0) {
  cat('::: {.panel-tabset}\n\n')
  int_umap_subset <- int_dt[, .(cell_id, UMAP1, UMAP2, cluster)]  # Subset once
  
  for (i_label in seq_along(names(panel_ls))) {
    label_name <- names(panel_ls)[i_label]
    cat('###', label_name, '\n\n')
    sel_expr_dt <- cell_exp_dt[as.character(label) == label_name]
    if (nrow(sel_expr_dt) == 0) {
      cat('No marker expression values were available for this label.\n\n')
      next
    }

    gene_levels <- unique(as.character(sel_expr_dt$symbol))
    for (i_gene in seq_along(gene_levels)) {
      gene_name <- gene_levels[i_gene]
      print(plot_expression_umap_pair(
        sel_expr_dt[as.character(symbol) == gene_name, .(cell_id, expr)],
        int_umap_subset,
        cluster_col = 'cluster',
        gene_name = gene_name,
        cluster_name = sprintf('cluster (res. %s)', sel_res)
      ))
      cat("\n\n")
      # Aggressive cleanup for large label/gene combinations
      if (i_gene %% 2 == 0) invisible(gc(verbose = FALSE))
    }
    if (i_label %% 2 == 0) invisible(gc(verbose = FALSE))
  }
  rm(int_umap_subset, cell_exp_dt); invisible(gc(verbose = FALSE))
  cat(':::\n\n')
} else {
  cat('No filtered H5 inputs were available to render marker expression UMAPs.\n\n')
}
```


## Annotation labels over UMAP

Cell labels in the panels below are assigned at the cluster level by the provided marker label with the highest mean positive marker log2 fold-change among the matched input markers. Cluster labels are written at each cluster centroid; if two clusters receive the same annotation label, that label is repeated once per cluster.

```{r plot_annotation_umap_overview, fig.height = 5, fig.width = 7}
g_label <- plot_umap_cluster(
  int_dt[, .(cell_id, UMAP1, UMAP2)],
  int_dt[, .(cell_id, cluster = label, centroid_group = as.character(label), centroid_label = as.character(label))],
  name = 'cell label'
)+theme(legend.position = "none")
g_cluster <- plot_umap_cluster(
  int_dt[, .(cell_id, UMAP1, UMAP2)],
  int_dt[, .(cell_id, cluster)],
  name = sprintf('cluster (res. %s)', sel_res)
)+theme(legend.position = "none")
print(g_label + g_cluster)
```

```{r celltype_sample_data, include = FALSE}
comp_dt <- int_dt[, .N, by = .(sample_id, label)]
comp_dt[, prop := N / sum(N), by = sample_id]

cat_meta_cols <- meta_var_cols[sapply(meta_var_cols, function(col) {
  v <- sample_meta_dt[[col]]
  is.character(v) || is.factor(v)
})]
cat_meta_cols <- setdiff(cat_meta_cols, "_harmony_batch")

label_levels <- levels(int_dt$label)
label_cols <- setNames(
  rep(nice_cols, length.out = length(label_levels)),
  label_levels
)
sample_levels <- unique(comp_dt$sample_id)
comp_dt[, sample_id := factor(sample_id, levels = sample_levels)]

add_comp_bar_theme <- function(p, label_cols) {
  p +
    geom_col(width = 0.8, colour = "black", linewidth = 0.2) +
    scale_fill_manual(values = label_cols, drop = FALSE) +
    scale_y_continuous(labels = label_percent(accuracy = 1), expand = expansion(mult = c(0, 0.02))) +
    scale_x_discrete(drop = TRUE) +
    theme_bw() +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor    = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      strip.background = element_rect(fill = "white"),
      legend.position  = "right",
      legend.title.position = "top",
      legend.text = element_text(size = 7),
      legend.key.size = unit(0.25, "cm")
    ) +
    guides(fill = guide_legend(ncol = 1, byrow = TRUE)) +
    labs(x = "sample", y = "proportion", fill = "cell type")
}

add_count_bar_theme <- function(p, label_cols) {
  p +
    geom_col(width = 0.8, colour = "black", linewidth = 0.2, show.legend = FALSE) +
    scale_fill_manual(values = label_cols, drop = FALSE) +
    scale_x_discrete(drop = TRUE) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    theme_bw() +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor    = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      strip.background = element_rect(fill = "white")
    ) +
    labs(x = "sample", y = "# cells")
}

get_meta_plot_dt <- function(comp_dt, sample_meta_dt, meta_col) {
  plot_dt <- merge(
    comp_dt,
    sample_meta_dt[, c("sample_id", meta_col), with = FALSE],
    by = "sample_id"
  )
  plot_dt[, sample_id := factor(as.character(sample_id), levels = sample_levels)]
  plot_dt[, (meta_col) := factor(as.character(get(meta_col)))]
  plot_dt
}

make_celltype_count_table <- function(comp_dt, sample_meta_dt, sample_levels, meta_cols) {
  counts_wide_dt <- dcast(
    copy(comp_dt)[, sample_id := as.character(sample_id)],
    sample_id ~ label,
    value.var = "N",
    fill = 0
  )

  props_wide_dt <- dcast(
    copy(comp_dt)[, sample_id := as.character(sample_id)],
    sample_id ~ label,
    value.var = "prop",
    fill = 0
  )

  count_cols <- setdiff(names(counts_wide_dt), "sample_id")
  if (length(count_cols) > 0) {
    clean_count_cols <- paste0("n_", gsub("[^A-Za-z0-9]+", "_", count_cols))
    setnames(counts_wide_dt, count_cols, clean_count_cols)

    prop_cols <- setdiff(names(props_wide_dt), "sample_id")
    clean_prop_cols <- paste0("pct_", gsub("[^A-Za-z0-9]+", "_", prop_cols))
    setnames(props_wide_dt, prop_cols, clean_prop_cols)
    props_wide_dt[, (clean_prop_cols) := lapply(.SD, function(x) round(100 * x, 2)), .SDcols = clean_prop_cols]
  }

  out_dt <- data.table(sample_id = sample_levels, sample_order = seq_along(sample_levels))
  if (length(meta_cols) > 0) {
    meta_dt <- unique(sample_meta_dt[, c("sample_id", meta_cols), with = FALSE], by = "sample_id")
    out_dt <- merge(out_dt, meta_dt, by = "sample_id", all.x = TRUE, sort = FALSE)
  }
  out_dt <- merge(out_dt, counts_wide_dt, by = "sample_id", all.x = TRUE, sort = FALSE)
  out_dt <- merge(out_dt, props_wide_dt, by = "sample_id", all.x = TRUE, sort = FALSE)
  setorder(out_dt, sample_order)
  out_dt[, sample_order := NULL]
  out_dt
}
```

## Cell-type numbers by sample

Bar charts show raw cell counts per sample, faceted by cell type with a fixed y-axis so absolute differences can be compared directly.

```{r celltype_numbers, fig.height = 6.5, fig.width = 7, results = 'asis'}
if (length(cat_meta_cols) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (meta_col in cat_meta_cols) {
    cat(sprintf("### %s\n\n", meta_col))
    plot_counts <- get_meta_plot_dt(comp_dt, sample_meta_dt, meta_col)
    meta_levels <- levels(plot_counts[[meta_col]])
    meta_levels <- meta_levels[!is.na(meta_levels) & nzchar(meta_levels)]
    for (meta_level in meta_levels) {
      level_dt <- plot_counts[get(meta_col) == meta_level]
      if (nrow(level_dt) == 0) {
        next
      }
      cat(sprintf("#### %s\n\n", meta_level))
      g <- add_count_bar_theme(
        ggplot(level_dt, aes(x = sample_id, y = N, fill = label)) +
          facet_wrap(~ label, ncol = 2, scales = "fixed"),
        label_cols
      )
      print(g)
      cat("\n\n")
    }
  }
  cat(':::\n\n')
} else {
  g <- add_count_bar_theme(
    ggplot(comp_dt, aes(x = sample_id, y = N, fill = label)) +
      facet_wrap(~ label, ncol = 2, scales = "fixed"),
    label_cols
  )
  print(g)
}
```

## Cell-type composition by sample

Stacked bar charts show cell-type proportions per sample. Each bar is normalised within a sample to 100%. When metadata covariates are available, each covariate gets its own tab and each covariate level is plotted separately to avoid squeezing many conditions into a single figure.

```{r celltype_composition, fig.height = 4.5, fig.width = 7, results = 'asis'}
if (length(cat_meta_cols) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (meta_col in cat_meta_cols) {
    cat(sprintf("### %s\n\n", meta_col))
    plot_comp <- get_meta_plot_dt(comp_dt, sample_meta_dt, meta_col)
    meta_levels <- levels(plot_comp[[meta_col]])
    meta_levels <- meta_levels[!is.na(meta_levels) & nzchar(meta_levels)]
    for (meta_level in meta_levels) {
      level_dt <- plot_comp[get(meta_col) == meta_level]
      if (nrow(level_dt) == 0) {
        next
      }
      cat(sprintf("#### %s\n\n", meta_level))
      g <- add_comp_bar_theme(
        ggplot(level_dt, aes(x = sample_id, y = prop, fill = label)),
        label_cols
      )
      print(g)
      cat("\n\n")
    }
  }
  cat(':::\n\n')
} else {
  g <- add_comp_bar_theme(
    ggplot(comp_dt, aes(x = sample_id, y = prop, fill = label)),
    label_cols
  )
  print(g)
}
```

## Cell-type counts table by sample

The table below lists raw cell counts per sample for each annotated cell type, alongside the metadata columns used in the composition bar charts.

```{r celltype_count_table}
count_table_dt <- make_celltype_count_table(comp_dt, sample_meta_dt, sample_levels, cat_meta_cols)

DT::datatable(
  count_table_dt,
  rownames = FALSE,
  class = 'compact stripe hover',
  options = list(
    pageLength = 25,
    lengthMenu = list(c(10, 25, 50), c('10', '25', '50')),
    autoWidth = TRUE,
    scrollX = TRUE
  )
)
```


## Session info

```{r session_info}
sessionInfo()
```