---
title: "Annotation predictions"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(scales)
  library(ggrastr)
  library(ggrepel)
  library(patchwork)
})

nice_cols <- c(
  '#DC050C', '#FB8072', '#1965B0', '#7BAFDE', '#882E72', '#B17BA6',
  '#FF7F00', '#FDB462', '#E7298A', '#E78AC3', '#33A02C', '#B2DF8A'
)

make_plot_palette <- function(levels_vec) {
  levels_vec <- unique(as.character(levels_vec))
  if (length(levels_vec) == 0) {
    return(setNames(character(), character()))
  }
  palette_vals <- rep(nice_cols, length.out = length(levels_vec))
  names(palette_vals) <- levels_vec
  palette_vals
}

order_cluster_ids <- function(cluster_ids) {
  cluster_ids <- unique(as.character(cluster_ids))
  if (length(cluster_ids) == 0) {
    return(character())
  }
  if (all(grepl('^cl[0-9]+$', cluster_ids))) {
    return(cluster_ids[order(as.integer(sub('^cl', '', cluster_ids)))])
  }
  sort(cluster_ids)
}

cluster_resolution_label <- function(cluster_col) {
  res_val <- sub('^RNA_snn_res\\.', '', cluster_col)
  sub('^leiden_', '', res_val)
}

coerce_cluster_summary <- function(dt) {
  if (nrow(dt) == 0) {
    return(dt)
  }
  char_cols <- intersect(c('cluster_col', 'cluster_resolution', 'cluster', 'top_label'), names(dt))
  if (length(char_cols) > 0) {
    dt[, (char_cols) := lapply(.SD, as.character), .SDcols = char_cols]
  }
  dt
}

build_cluster_summary <- function(cells_dt, cluster_cols) {
  if (length(cluster_cols) == 0) {
    return(data.table(
      cluster_col = character(),
      cluster_resolution = character(),
      cluster = character(),
      n_cells = integer(),
      top_label = character(),
      cs = numeric(),
      mean_score = numeric()
    ))
  }

  rbindlist(lapply(cluster_cols, function(cluster_col) {
    dt <- cells_dt[
      !is.na(get(cluster_col)) & nzchar(as.character(get(cluster_col))),
      .(
        cluster = as.character(get(cluster_col)),
        label = as.character(label),
        score = as.numeric(score)
      )
    ]
    if (nrow(dt) == 0) {
      return(NULL)
    }

    counts_dt <- dt[, .N, by = .(cluster, label)]
    totals_dt <- dt[, .(
      n_cells = .N,
      mean_score = mean(score, na.rm = TRUE)
    ), by = cluster]

    summary_dt <- merge(counts_dt, totals_dt, by = 'cluster', all.x = TRUE, sort = FALSE)
    summary_dt[, cs := N / n_cells]
    setorder(summary_dt, cluster, -cs, -N, label)
    summary_dt <- summary_dt[, .SD[1], by = cluster]
    summary_dt[, `:=`(
      cluster_col = cluster_col,
      cluster_resolution = cluster_resolution_label(cluster_col),
      top_label = label
    )]
    summary_dt[, .(cluster_col, cluster_resolution, cluster, n_cells, top_label, cs, mean_score)]
  }), use.names = TRUE, fill = TRUE)
}

plot_umap_cluster_consistent <- function(umap_dt, clust_dt, name) {
  stopifnot(
    all(c('UMAP1', 'UMAP2') %in% names(umap_dt)),
    'cell_id' %in% names(umap_dt),
    'cell_id' %in% names(clust_dt),
    'cluster' %in% names(clust_dt)
  )

  plot_dt <- merge(umap_dt, clust_dt, by = 'cell_id', all.x = TRUE)[, .(
    UMAP1 = rescale(UMAP1, to = c(0.05, 0.95)),
    UMAP2 = rescale(UMAP2, to = c(0.05, 0.95)),
    cluster,
    centroid_group = if ('centroid_group' %in% names(clust_dt)) centroid_group else cluster,
    centroid_label = if ('centroid_label' %in% names(clust_dt)) centroid_label else cluster
  )]

  if (is.factor(clust_dt$cluster)) {
    cluster_levels <- levels(clust_dt$cluster)
  } else {
    cluster_levels <- order_cluster_ids(plot_dt$cluster)
  }
  plot_dt[, cluster := factor(as.character(cluster), levels = cluster_levels)]
  plot_dt[, centroid_group := as.character(centroid_group)]
  plot_dt[, centroid_label := as.character(centroid_label)]
  plot_dt <- plot_dt[sample(.N, .N)]

  label_dt <- plot_dt[
    !is.na(centroid_group) & nzchar(centroid_group),
    .(
      UMAP1 = median(UMAP1),
      UMAP2 = median(UMAP2),
      centroid_label = centroid_label[1]
    ),
    by = centroid_group
  ]

  cl_cols <- make_plot_palette(levels(plot_dt$cluster))
  n_col <- 4
  n_rows_lgd <- ceiling(length(cl_cols) / n_col)

  ggplot(plot_dt) +
    aes(x = UMAP1, y = UMAP2, colour = cluster) +
    geom_point(size = 0.1) +
    geom_text_repel(
      data = label_dt,
      aes(x = UMAP1, y = UMAP2, label = centroid_label),
      inherit.aes = FALSE,
      size = 3,
      seed = 1,
      min.segment.length = 0,
      max.overlaps = Inf,
      box.padding = 0.3,
      point.padding = 0.15,
      segment.color = 'grey55'
    ) +
    scale_colour_manual(
      values = cl_cols,
      guide = guide_legend(override.aes = list(size = 3), nrow = n_rows_lgd)
    ) +
    scale_x_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    scale_y_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      aspect.ratio = 1,
      legend.title.position = 'left',
      legend.position = 'bottom',
      axis.ticks = element_blank(),
      axis.text = element_blank()
    ) +
    labs(colour = name)
}

plot_umap_score_consistent <- function(umap_dt, score_dt, name = 'cs') {
  stopifnot(
    all(c('UMAP1', 'UMAP2') %in% names(umap_dt)),
    'cell_id' %in% names(umap_dt),
    'cell_id' %in% names(score_dt),
    'score' %in% names(score_dt)
  )

  plot_dt <- merge(umap_dt, score_dt, by = 'cell_id', all.x = TRUE)[, .(
    UMAP1 = rescale(UMAP1, to = c(0.05, 0.95)),
    UMAP2 = rescale(UMAP2, to = c(0.05, 0.95)),
    score,
    centroid_group = if ('centroid_group' %in% names(score_dt)) centroid_group else NA_character_,
    centroid_label = if ('centroid_label' %in% names(score_dt)) centroid_label else NA_character_
  )]
  plot_dt <- plot_dt[sample(.N, .N)]

  label_dt <- plot_dt[
    !is.na(centroid_group) & nzchar(centroid_group),
    .(
      UMAP1 = median(UMAP1),
      UMAP2 = median(UMAP2),
      centroid_label = centroid_label[1]
    ),
    by = centroid_group
  ]

  ggplot(plot_dt) +
    aes(x = UMAP1, y = UMAP2, colour = score) +
    geom_point(size = 0.1) +
    geom_text_repel(
      data = label_dt,
      aes(x = UMAP1, y = UMAP2, label = centroid_label),
      inherit.aes = FALSE,
      size = 3,
      seed = 1,
      min.segment.length = 0,
      max.overlaps = Inf,
      box.padding = 0.3,
      point.padding = 0.15,
      segment.color = 'grey55'
    ) +
    scale_colour_viridis_c(option = 'inferno', na.value = 'grey80', limits = c(0.5, 1)) +
    scale_x_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    scale_y_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      aspect.ratio = 1,
      legend.title.position = 'left',
      legend.position = 'bottom',
      axis.ticks = element_blank(),
      axis.text = element_blank()
    ) +
    labs(colour = name)
}

method_id <- Sys.getenv("ANNOTATION_METHOD_ID", "annotation_method")
engine <- Sys.getenv("ANNOTATION_ENGINE", "annotation")
reference_name <- Sys.getenv("ANNOTATION_REFERENCE_NAME", "reference")
reference_label_col <- Sys.getenv("ANNOTATION_REFERENCE_LABEL_COL", "")
cells_csv <- Sys.getenv("ANNOTATION_CELLS_CSV", "annotation_cells.csv.gz")
cluster_csv <- Sys.getenv("ANNOTATION_CLUSTER_CSV", "annotation_cluster_summary.csv.gz")

cells_dt <- fread(cells_csv)
cluster_dt_input <- fread(cluster_csv)

cells_dt[, label := as.character(label)]
cells_dt[is.na(label) | !nzchar(label), label := "Unassigned"]

cluster_cols <- grep("^(RNA_snn_res\\.|leiden_)", names(cells_dt), value = TRUE)
cluster_cols <- cluster_cols[order(as.numeric(cluster_resolution_label(cluster_cols)))]

cluster_dt_input <- coerce_cluster_summary(cluster_dt_input)
cluster_dt <- build_cluster_summary(cells_dt, cluster_cols)

label_counts_dt <- cells_dt[, .N, by = label][order(-N, label)]
label_levels <- label_counts_dt$label
label_cols <- make_plot_palette(label_levels)
cells_dt[, label := factor(label, levels = label_levels)]

cluster_summary_missing_cols <- if (
  nrow(cluster_dt_input) > 0 &&
  all(c('cluster_col', 'cluster_resolution', 'cluster', 'top_label', 'cs') %in% names(cluster_dt_input))
) {
  setdiff(cluster_cols, unique(cluster_dt_input$cluster_col))
} else {
  cluster_cols
}
```

This report summarises one annotation run.

* **Method id:** `r method_id`
* **Engine:** `r engine`
* **Reference:** `r reference_name`
* **Reference label column:** `r if (nzchar(reference_label_col)) reference_label_col else "not specified"`
* **Annotated cells:** `r scales::comma(nrow(cells_dt))`
* **Assigned labels:** `r data.table::uniqueN(cells_dt$label)`
* **Cluster score (`cs`):** within-cluster fraction assigned to the top predicted label
* **Cluster summary source:** `r if (length(cluster_summary_missing_cols) > 0) "recomputed from per-cell predictions to cover all clustering resolutions" else "recomputed from per-cell predictions for report consistency"`

## Label counts

```{r label_counts, fig.width=7, fig.height=5}
label_counts_plot_dt <- copy(label_counts_dt)
label_counts_plot_dt[, label := factor(label, levels = rev(label_levels))]

ggplot(label_counts_plot_dt, aes(x = label, y = N, fill = label)) +
  geom_col(width = 0.8, colour = "black", linewidth = 0.2, show.legend = FALSE) +
  geom_text(
    aes(label = comma(N)),
    hjust = -0.05,
    size = 3,
    colour = 'black',
    show.legend = FALSE
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = label_cols, drop = FALSE) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    plot.margin = margin(5.5, 30, 5.5, 5.5)
  ) +
  labs(x = NULL, y = "# cells", title = "Predicted label counts")
```

## Prediction score distribution

```{r score_density, fig.width=8, fig.height=4}
if ("score" %in% names(cells_dt) && any(!is.na(cells_dt$score))) {
  ggplot(cells_dt[!is.na(score)], aes(x = score)) +
    geom_histogram(bins = 40, fill = "#7FCDBB", color = "white") +
    theme_bw() +
    labs(x = "prediction score", y = "cells", title = "Prediction score distribution")
}
```

## UMAP by predicted label and score

Predicted label (left) and prediction score (right) over the same UMAP.

```{r umap_label_score, fig.width=9, fig.height=7}
if (all(c("UMAP1", "UMAP2") %in% names(cells_dt))) {
  g_lab <- plot_umap_cluster_consistent(
      cells_dt[, .(cell_id, UMAP1, UMAP2)],
      cells_dt[, .(
        cell_id,
        cluster = label,
        centroid_group = as.character(label),
        centroid_label = as.character(label)
      )],
      name = 'cell label'
    ) + theme(legend.position = 'none') + ggtitle('predicted label')

  if ("score" %in% names(cells_dt)) {
    g_score <- plot_umap_score_consistent(
      cells_dt[, .(cell_id, UMAP1, UMAP2)],
      cells_dt[, .(cell_id, score)],
      name = 'score'
    ) + ggtitle('prediction score')
    print(g_lab | g_score)
  } else {
    print(g_lab)
  }
}
```

## Cluster Prediction Summary {.tabset}

For each clustering resolution, the confidence score (`cs`) is calculated per cluster as the fraction of cells assigned to that cluster's most frequent predicted label: $cs = \frac{n(\text{top label in cluster})}{n(\text{cells in cluster})}$. A value of `1` means every cell in the cluster shares the same predicted label, while values closer to `0.5` indicate a more mixed or uncertain cluster. Each panel below is derived directly from the per-cell annotation table, which avoids stale or partial cluster-summary inputs and ensures that every clustering resolution in the annotation data is represented.

```{r cluster_umaps, fig.width=7, fig.height=5, results='asis'}
if (length(cluster_cols) > 0 && all(c("UMAP1", "UMAP2") %in% names(cells_dt)) && nrow(cluster_dt) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cluster_col in cluster_cols) {
    res_val <- cluster_resolution_label(cluster_col)
    this_cluster_col <- cluster_col
    cat(sprintf('### %s\n\n', res_val))

    plot_dt <- cells_dt[
      !is.na(UMAP1) & !is.na(UMAP2) & !is.na(get(cluster_col)) & nzchar(as.character(get(cluster_col))),
      .(
        UMAP1,
        UMAP2,
        cluster = as.character(get(cluster_col))
      )
    ]
    if (nrow(plot_dt) == 0) {
      cat('No cells had non-missing cluster assignments at this resolution.\n\n')
      next
    }

    cluster_levels <- order_cluster_ids(plot_dt$cluster)
    cluster_labels_dt <- cluster_dt[cluster_dt$cluster_col == this_cluster_col]
    cluster_labels_dt[, `:=`(
      cluster = factor(cluster, levels = cluster_levels),
      centroid_group = as.character(cluster),
      centroid_label = sprintf('%s: %.2f', top_label, cs)
    )]

    print(
      plot_umap_cluster_consistent(
        cells_dt[, .(cell_id, UMAP1, UMAP2)],
        cells_dt[
          !is.na(get(cluster_col)) & nzchar(as.character(get(cluster_col))),
          .(cell_id, cluster = factor(as.character(get(cluster_col)), levels = cluster_levels))
        ][cluster_labels_dt[, .(cluster, centroid_group, centroid_label)], on = 'cluster'],
        name = sprintf('cluster (res. %s)', res_val)
      ) + theme(legend.position = 'none') + ggtitle(sprintf('res = %s', res_val))
    )
    cat('\n\n')
  }
  cat(':::\n\n')
}
```

## Cluster Confidence by Resolution {.tabset}

This section shows the same clustering resolutions, but colours each cluster by its within-cluster top-label confidence score (`cs`).

```{r cluster_confidence_umaps, fig.width=7, fig.height=5, results='asis'}
if (length(cluster_cols) > 0 && all(c("UMAP1", "UMAP2") %in% names(cells_dt)) && nrow(cluster_dt) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cluster_col in cluster_cols) {
    res_val <- cluster_resolution_label(cluster_col)
    this_cluster_col <- cluster_col
    cat(sprintf('### %s\n\n', res_val))

    plot_dt <- cells_dt[
      !is.na(UMAP1) & !is.na(UMAP2) & !is.na(get(cluster_col)) & nzchar(as.character(get(cluster_col))),
      .(
        UMAP1,
        UMAP2,
        cluster = as.character(get(cluster_col))
      )
    ]
    if (nrow(plot_dt) == 0) {
      cat('No cells had non-missing cluster assignments at this resolution.\n\n')
      next
    }

    cluster_levels <- order_cluster_ids(plot_dt$cluster)
    cluster_scores_dt <- cluster_dt[cluster_dt$cluster_col == this_cluster_col]
    cluster_scores_dt[, `:=`(
      cluster = as.character(cluster),
      centroid_group = as.character(cluster),
      centroid_label = sprintf('%s: %.2f', top_label, cs),
      score = cs
    )]

    print(
      plot_umap_score_consistent(
        cells_dt[, .(cell_id, UMAP1, UMAP2)],
        cells_dt[
          !is.na(get(cluster_col)) & nzchar(as.character(get(cluster_col))),
          .(cell_id, cluster = as.character(get(cluster_col)))
        ][cluster_scores_dt[, .(cluster, score, centroid_group, centroid_label)], on = 'cluster'],
        name = 'cs'
      ) + ggtitle(sprintf('res = %s', res_val))
    )
    cat('\n\n')
  }
  cat(':::\n\n')
}
```

## Cluster Summary Table

```{r cluster_table}
cluster_table_dt <- data.table::copy(cluster_dt)
cluster_table_dt[, cs := signif(cs, 3)]
cluster_table_dt[, mean_score := signif(mean_score, 3)]

DT::datatable(
  cluster_table_dt[order(as.numeric(cluster_resolution), cluster)],
  rownames = FALSE,
  class = 'compact stripe hover',
  options = list(
    pageLength = 25,
    lengthMenu = list(c(10, 25, 50), c('10', '25', '50')),
    autoWidth = TRUE,
    scrollX = TRUE
  )
)
```