---
title: "Annotation diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
  # Memory optimization: render plots to PNG immediately (rasterize large ComplexHeatmap objects)
  fig.keep: 'high'
  # Garbage collect aggressively in the report
  include: true
---

```{r setup}
source("annotation_utils.R")
source("integration_plots.R")

# Helper: load logCPMs for specific genes only (memory-efficient)
load_logcpms_subset <- function(cpms_path, genes_of_interest, verbose = FALSE) {
  genes_of_interest <- unique(genes_of_interest[!is.na(genes_of_interest) & nzchar(genes_of_interest)])
  if (length(genes_of_interest) == 0) {
    return(data.table())
  }

  if (verbose) cat("Loading logCPMs for", length(genes_of_interest), "genes...\n")

  genes_f <- tempfile(fileext = ".txt")
  on.exit(unlink(genes_f), add = TRUE)
  fwrite(data.table(gene_id = genes_of_interest), genes_f, col.names = FALSE)

  awk_script <- paste(
    "NR==FNR { keep[$1]=1; next }",
    "FNR==1 || ($1 in keep)"
  )
  cmd_inner <- sprintf(
    "set -euo pipefail; gzip -dc %s | awk -F',' %s %s -",
    shQuote(cpms_path),
    shQuote(awk_script),
    shQuote(genes_f)
  )
  cmd <- sprintf("bash -lc %s", shQuote(cmd_inner))

  fread(cmd = cmd)
}

sel_res <- Sys.getenv("ANNOTATION_SEL_RES", "0.2")
min_cpm_mkr <- as.numeric(Sys.getenv("ANNOTATION_MIN_CPM_MKR", "50"))
not_ok_re <- Sys.getenv("ANNOTATION_NOT_OK_RE", "(lincRNA|lncRNA|pseudogene|antisense)")
top_n <- as.integer(Sys.getenv("ANNOTATION_TOP_N", "10"))
fdr_cut <- as.numeric(Sys.getenv("ANNOTATION_FDR_CUT", "0.01"))
max_zero_p <- as.numeric(Sys.getenv("ANNOTATION_MAX_ZERO_P", "0.5"))
```

```{r load_data}
int_dt <- fread("integration_dt.csv.gz")
mkrs_dt <- fread("annotation_markers.csv.gz")
# Don't load full logCPMs yet; load on-demand per section
cpms_path <- "annotation_logcpms.csv.gz"
cpms_dt <- NULL
panel_dt <- fread("annotation_marker_panel.csv.gz")
marker_expr_rds <- "annotation_marker_expression.rds"
h5_files <- Sys.glob("filt_counts_*.h5")
cell_labels_f <- "annotation_cell_labels.csv.gz"

cl_col <- paste0("RNA_snn_res.", sel_res)
stopifnot("Selected annotation resolution not found in integration output" = cl_col %in% names(int_dt))

# Save sample-level metadata columns before subsetting (used for composition barcharts)
core_int_re <- "^(cell_id|UMAP1|UMAP2|embedding|dbl_|is_dbl|is_doublet|RNA_snn_res\\.|leiden_)"
meta_var_cols <- setdiff(names(int_dt)[!grepl(core_int_re, names(int_dt))], "sample_id")
sample_meta_dt <- if (length(meta_var_cols) > 0) {
  unique(int_dt[get(cl_col) != "" | is.na(get(cl_col)), c("sample_id", meta_var_cols), with = FALSE], by = "sample_id")
} else {
  data.table(sample_id = unique(int_dt$sample_id))
}

umap_vars <- unique(c("sample_id", meta_var_cols))
umap_vars <- umap_vars[umap_vars %in% names(int_dt)]
int_umap_for_meta <- int_dt[!is.na(UMAP1), c("cell_id", "UMAP1", "UMAP2", umap_vars), with = FALSE]

int_dt <- int_dt[!is.na(get(cl_col)) & sample_id != "", .(
  sample_id,
  cell_id,
  UMAP1,
  UMAP2,
  cluster = get(cl_col)
)]

cl_ord <- order_cluster_labels(unique(int_dt$cluster))
int_dt[, cluster := factor(as.character(cluster), levels = cl_ord)]
panel_label_levels <- unique(as.character(panel_dt$label))
panel_dt[, label := factor(as.character(label), levels = panel_label_levels)]

top_min_dt <- mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][
  logcpm.sel >= log(min_cpm_mkr + 1)
] |> get_top_markers(fdr_cut = fdr_cut, n_top = top_n, max_zero_p = max_zero_p)

top_heatmap_panel_dt <- make_top_marker_heatmap_panel(top_min_dt, cl_order = cl_ord)

top50_dt <- mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][
  logFC > 0 & logcpm.sel >= log(min_cpm_mkr + 1)
][order(cluster, FDR, -logFC, -logcpm.sel, symbol)]
top50_dt[, prop_zero := n_zero / n_cl]
top50_dt[, marker_rank := seq_len(.N), by = cluster]
top50_dt <- top50_dt[marker_rank <= 50]

marker_coverage_dt <- panel_dt[, .(
  markers = paste(unique(symbol), collapse = ", ")
), by = label]

if (file.exists(cell_labels_f)) {
  cell_labels_dt <- fread(cell_labels_f)
  cluster_label_dt <- unique(cell_labels_dt[, .(cluster, label, label_score, n_markers)])
  cell_label_lu <- unique(cell_labels_dt[, .(cell_id, label, label_score, n_markers)])
  label_lu <- setNames(as.character(cell_label_lu$label), cell_label_lu$cell_id)
  label_score_lu <- setNames(cell_label_lu$label_score, cell_label_lu$cell_id)
  n_markers_lu <- setNames(cell_label_lu$n_markers, cell_label_lu$cell_id)
  int_dt[, label := label_lu[cell_id]]
  int_dt[, label_score := label_score_lu[cell_id]]
  int_dt[, n_markers := n_markers_lu[cell_id]]
} else {
  cluster_label_dt <- assign_cluster_labels_from_panel(mkrs_dt, panel_dt, min_cpm = min_cpm_mkr)
  cluster_label_dt <- merge(data.table(cluster = cl_ord), cluster_label_dt, by = "cluster", all.x = TRUE, sort = FALSE)
  label_lu <- setNames(as.character(cluster_label_dt$label), as.character(cluster_label_dt$cluster))
  label_score_lu <- setNames(cluster_label_dt$label_score, as.character(cluster_label_dt$cluster))
  n_markers_lu <- setNames(cluster_label_dt$n_markers, as.character(cluster_label_dt$cluster))
  int_dt[, label := label_lu[as.character(cluster)]]
  int_dt[, label_score := label_score_lu[as.character(cluster)]]
  int_dt[, n_markers := n_markers_lu[as.character(cluster)]]
}
cluster_label_dt <- merge(data.table(cluster = cl_ord), cluster_label_dt, by = "cluster", all.x = TRUE, sort = FALSE)
cluster_label_dt[is.na(label), `:=`(label = "Unassigned", label_score = NA_real_, n_markers = 0L)]
cluster_label_dt[, label := factor(as.character(label), levels = unique(c(panel_label_levels, "Unassigned")))]

int_dt[is.na(label), label := "Unassigned"]
int_dt[, label := factor(as.character(label), levels = unique(c(panel_label_levels, "Unassigned")))]

panel_ls <- setNames(
  lapply(panel_label_levels, function(label_name) panel_dt[as.character(label) == label_name]),
  panel_label_levels
)

# Determine which genes will actually be used in plots
genes_for_plots <- unique(c(
  panel_dt$gene_id,
  mkrs_dt[str_detect(gene_type, not_ok_re, negate = TRUE)][logcpm.sel >= log(min_cpm_mkr + 1)]$gene_id,
  top_min_dt$gene_id
))

# Load logCPMs only for genes that will be displayed
cpms_dt <- load_logcpms_subset(cpms_path, genes_for_plots, verbose = TRUE)

cell_exp_dt <- data.table()

# Clean up after initial data preparation
gc(verbose = FALSE)
```

## Annotation setup

This report assigns labels from the integration clustering column `r cl_col`, then tests marker genes by aggregating counts per sample within each cluster and running one-vs-rest pseudobulk differential expression.

* **Selected cluster resolution used for annotation:** `r sel_res` (`r cl_col`)
* **Clusters carried into annotation:** `r length(cl_ord)`
* **Cells used for annotation:** `r format(nrow(int_dt), big.mark = ",")`
* **Provided marker genes matched in the data:** `r uniqueN(panel_dt$gene_id)`
* **Marker-expression cache available for UMAPs:** `r if (file.exists(marker_expr_rds)) "yes" else "no"`

## Input marker panel

```{r marker_panel_table}
DT::datatable(
  marker_coverage_dt,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", paging = FALSE, searching = FALSE, info = FALSE, ordering = FALSE, autoWidth = TRUE)
)
```

## Heatmaps of marker genes

::: {.panel-tabset}

### Provided markers

```{r plot_heatmap_provided_markers, fig.height = 9, fig.width = 10}
suppressMessages(draw(
  plot_heatmap_of_selected_genes(mkrs_dt, panel_dt),
  heatmap_legend_side = "bottom",
  annotation_legend_side = "bottom",
  merge_legend = TRUE
))
rm(panel_dt); gc(verbose = FALSE)
```

### Top markers by cluster

```{r plot_heatmap_top_markers, fig.height = 9, fig.width = 10}
if (nrow(top_heatmap_panel_dt) > 0) {
  suppressMessages(draw(
    plot_heatmap_of_selected_genes(mkrs_dt, top_heatmap_panel_dt),
    heatmap_legend_side = "bottom",
    annotation_legend_side = "bottom",
    merge_legend = TRUE
  ))
  rm(top_heatmap_panel_dt); gc(verbose = FALSE)
} else {
  cat("No top markers passed the display filters for the heatmap.\n")
}
```

:::

## Top marker genes by cluster at resolution `r sel_res`

```{r plot_top_markers, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (sel_cl in cl_ord) {
  cat(sprintf("### %s\n\n", sel_cl))
  has_markers  <- sel_cl %in% as.character(top_min_dt$cluster)
  top_sel_dt <- unique(top_min_dt[cluster == sel_cl, .(label = as.character(cluster), symbol, gene_id)])
  has_umap_expr <- has_markers && nrow(top_sel_dt) > 0 && length(h5_files) > 0
  if (!has_markers) {
    cat("No marker genes passed the display filters for this cluster.\n\n")
  } else if (!has_umap_expr) {
    print(plot_top_marker_genes(sel_cl, top_min_dt, cpms_dt, cl_order = cl_ord))
    cat("\n\n")
  } else {
    top_marker_expr_dt <- load_h5_marker_expression_multi(
      h5_files,
      sel_dt_list = list(top = top_sel_dt),
      int_dt = int_dt[, .(sample_id, cell_id, UMAP1, UMAP2)]
    )$top
    cat("::: {.panel-tabset}\n\n")
    cat("#### Expression dotplot\n\n")
    print(plot_top_marker_genes(sel_cl, top_min_dt, cpms_dt, cl_order = cl_ord))
    cat("\n\n")
    cat("#### UMAP feature plots\n\n")
    g_umap <- plot_top_marker_umap_facet(
      sel_cl, top_min_dt, top_marker_expr_dt,
      int_dt[, .(cell_id, UMAP1, UMAP2)]
    )
    if (!is.null(g_umap)) print(g_umap)
    cat("\n\n")
    cat(":::\n\n")
    rm(top_marker_expr_dt)
  }
  # Clean up after each cluster to avoid memory accumulation
  gc(verbose = FALSE)
}
cat(':::\n\n')
```

## Provided marker genes by label

```{r plot_provided_marker_panels, fig.height = 9, fig.width = 10, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (i_panel in seq_along(names(panel_ls))) {
  label_name <- names(panel_ls)[i_panel]
  cat('###', label_name, '\n\n')
  sel_panel_dt <- panel_ls[[label_name]]
  highlight_clusters <- as.character(cluster_label_dt[as.character(label) == label_name, cluster])
  print(plot_selected_marker_panel(
    sel_panel_dt,
    cpms_dt,
    cl_order = cl_ord,
    highlight_clusters = highlight_clusters,
    title = sprintf('%s provided marker genes', label_name)
  ))
  cat('\n\n')
  # Clean up every 3 panels
  if (i_panel %% 3 == 0) gc(verbose = FALSE)
}
cat(':::\n\n')
```

## Annotation labels over UMAP

Cell labels in the panels below are assigned at the cluster level by the provided marker label with the highest mean positive marker log2 fold-change among the matched input markers. Cluster labels are written at each cluster centroid; if two clusters receive the same annotation label, that label is repeated once per cluster.

```{r plot_annotation_umap_overview, fig.height = 7, fig.width = 14}
g_label <- plot_umap_cluster(
  int_dt[, .(cell_id, UMAP1, UMAP2)],
  int_dt[, .(cell_id, cluster = label, centroid_group = as.character(cluster), centroid_label = as.character(label))],
  name = 'cell label'
)
g_cluster <- plot_umap_cluster(
  int_dt[, .(cell_id, UMAP1, UMAP2)],
  int_dt[, .(cell_id, cluster)],
  name = sprintf('cluster (res. %s)', sel_res)
)
print(g_label + g_cluster + patchwork::plot_layout(widths = c(1, 1)))
```

## Cell-type composition by sample

Stacked bar charts show cell-type proportions per sample. Each bar is normalised within a sample to 100%. When metadata covariates are available, samples are faceted by covariate value (one tab per covariate).

```{r celltype_composition, fig.height = 5, fig.width = 10, results = 'asis'}
comp_dt <- int_dt[, .N, by = .(sample_id, label)]
comp_dt[, prop := N / sum(N), by = sample_id]

cat_meta_cols <- meta_var_cols[sapply(meta_var_cols, function(col) {
  v <- sample_meta_dt[[col]]
  is.character(v) || is.factor(v)
})]

label_levels <- levels(int_dt$label)
label_cols <- setNames(
  rep(nice_cols, length.out = length(label_levels)),
  label_levels
)

add_comp_bar_theme <- function(p, label_cols, label_levels) {
  p +
    geom_col(width = 0.8, colour = "black", linewidth = 0.2) +
    scale_fill_manual(values = label_cols, drop = FALSE) +
    scale_y_continuous(labels = label_percent(accuracy = 1), expand = expansion(mult = c(0, 0.02))) +
    theme_bw() +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor    = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      strip.background = element_rect(fill = "white"),
      legend.position  = "bottom",
      legend.title.position = "left"
    ) +
    guides(fill = guide_legend(ncol = max(1, ceiling(length(label_levels) / 10)))) +
    labs(x = "sample", y = "proportion", fill = "cell type")
}

if (length(cat_meta_cols) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (meta_col in cat_meta_cols) {
    cat(sprintf("### %s\n\n", meta_col))
    plot_comp <- merge(
      comp_dt,
      sample_meta_dt[, c("sample_id", meta_col), with = FALSE],
      by = "sample_id"
    )
    plot_comp[, (meta_col) := factor(as.character(get(meta_col)))]
    g <- add_comp_bar_theme(
      ggplot(plot_comp, aes(x = sample_id, y = prop, fill = label)) +
        facet_wrap(reformulate(paste0("`", meta_col, "`")), scales = "free_x"),
      label_cols, label_levels
    )
    print(g)
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  g <- add_comp_bar_theme(
    ggplot(comp_dt, aes(x = sample_id, y = prop, fill = label)),
    label_cols, label_levels
  )
  print(g)
}
```

## Top 50 marker tables by cluster at resolution `r sel_res`

The tables below list up to 50 positive markers per cluster after excluding genes that match `r not_ok_re` and requiring expression above the annotation CPM threshold. Each table shows 25 rows per page.

```{r top_marker_tables, results = 'asis'}
cat('::: {.panel-tabset}\n\n')
for (sel_cl in cl_ord) {
  cat(sprintf("### %s\n\n", sel_cl))
  sel_tbl <- top50_dt[as.character(cluster) == as.character(sel_cl), .(
    rank = marker_rank,
    symbol,
    gene_type,
    logFC = round(logFC, 2),
    FDR = signif(FDR, 3),
    logcpm_sel = round(logcpm.sel, 2),
    logcpm_other = round(logcpm.other, 2),
    zero_prop = round(prop_zero, 3)
  )]
  if (nrow(sel_tbl) > 0) {
    print(htmltools::tagList(DT::datatable(
      sel_tbl,
      rownames = FALSE,
      class = 'compact stripe hover',
      options = list(
        pageLength = 25,
        lengthMenu = list(c(10, 25), c('10', '25')),
        dom = 'tip',
        autoWidth = TRUE,
        scrollX = TRUE
      )
    )))
  } else {
    cat('No marker genes passed the table filters for this cluster.\n\n')
  }
  cat('\n\n')
}
cat(':::\n\n')
```

## Marker expression over UMAP

Marker-expression panels below use the cached annotation marker-expression file and colour cells by uncapped log CPM values rather than per-gene values rescaled to 1.

```{r plot_marker_umaps, fig.height = 7, fig.width = 14, results = 'asis'}
cell_exp_dt <- if (file.exists(marker_expr_rds)) readRDS(marker_expr_rds) else data.table()

if (nrow(cell_exp_dt) > 0) {
  cat('::: {.panel-tabset}\n\n')
  int_umap_subset <- int_dt[, .(cell_id, UMAP1, UMAP2, cluster)]  # Subset once
  
  for (i_label in seq_along(names(panel_ls))) {
    label_name <- names(panel_ls)[i_label]
    cat('###', label_name, '\n\n')
    sel_expr_dt <- cell_exp_dt[as.character(label) == label_name]
    if (nrow(sel_expr_dt) == 0) {
      cat('No marker expression values were available for this label.\n\n')
      next
    }

    gene_levels <- unique(as.character(sel_expr_dt$symbol))
    for (i_gene in seq_along(gene_levels)) {
      gene_name <- gene_levels[i_gene]
      cat('####', gene_name, '\n\n')
      print(plot_expression_umap_pair(
        sel_expr_dt[as.character(symbol) == gene_name, .(cell_id, expr)],
        int_umap_subset,
        cluster_col = 'cluster',
        gene_name = gene_name,
        cluster_name = sprintf('cluster (res. %s)', sel_res)
      ))
      cat('\n\n')
      # Aggressive cleanup for large label/gene combinations
      if (i_gene %% 2 == 0) gc(verbose = FALSE)
    }
    if (i_label %% 2 == 0) gc(verbose = FALSE)
  }
  rm(int_umap_subset, cell_exp_dt); gc(verbose = FALSE)
  cat(':::\n\n')
} else {
  cat('No filtered H5 inputs were available to render marker expression UMAPs.\n\n')
}
```

## Metadata across UMAP space

The plots below show how metadata variables are distributed across UMAP space using binned hexagonal plots. Each hexagon is coloured by the proportion of cells with each metadata value, allowing identification of spatial patterns and associations with cell types or clusters.

```{r plot_metadata_umaps, fig.height = 7, fig.width = 14, results='asis'}
if (length(umap_vars) > 0 && nrow(int_umap_for_meta) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in umap_vars) {
    cat('###', var_name, '\n\n')
    var_vals <- int_umap_for_meta[[var_name]]
    var_vals <- var_vals[!is.na(var_vals)]
    
    if (length(var_vals) == 0) {
      cat('No values available for this variable.\n\n')
      next
    }
    
    # For categorical variables, show each category in a separate plot
    if (!is.numeric(var_vals) || uniqueN(var_vals) <= 12) {
      unique_vals <- unique(as.character(var_vals))
      unique_vals <- sort(unique_vals)
      n_cols <- min(3, length(unique_vals))
      
      plot_list <- lapply(unique_vals, function(focus_val) {
        plot_umap_metadata_binned(int_umap_for_meta, var_name, focus_value = focus_val)
      })
      
      combined <- wrap_plots(plot_list, ncol = n_cols, guides = 'auto')
      print(combined)
    } else {
      # For continuous variables, show overall density
      print(plot_umap_metadata_binned(int_umap_for_meta, var_name, focus_value = NULL))
    }
    cat('\n\n')
  }
  cat(':::\n\n')
  rm(int_umap_for_meta); gc(verbose = FALSE)
} else {
  cat('No metadata variables available for visualization.\n\n')
}
```

## Session info

```{r session_info}
sessionInfo()
```