---
title: "Annotation predictions"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(scales)
  library(ggrastr)
})

method_id <- Sys.getenv("ANNOTATION_METHOD_ID", "annotation_method")
engine <- Sys.getenv("ANNOTATION_ENGINE", "annotation")
reference_name <- Sys.getenv("ANNOTATION_REFERENCE_NAME", "reference")
reference_label_col <- Sys.getenv("ANNOTATION_REFERENCE_LABEL_COL", "")
cells_csv <- Sys.getenv("ANNOTATION_CELLS_CSV", "annotation_cells.csv.gz")
cluster_csv <- Sys.getenv("ANNOTATION_CLUSTER_CSV", "annotation_cluster_summary.csv.gz")

cells_dt <- fread(cells_csv)
cluster_dt <- fread(cluster_csv)

cells_dt[, label := as.character(label)]
cells_dt[is.na(label) | !nzchar(label), label := "Unassigned"]

if (nrow(cluster_dt) > 0) {
  char_cols <- intersect(c("cluster_col", "cluster_resolution", "cluster", "top_label"), names(cluster_dt))
  if (length(char_cols) > 0) {
    cluster_dt[, (char_cols) := lapply(.SD, as.character), .SDcols = char_cols]
  }
}

cluster_cols <- grep("^(RNA_snn_res\\.|leiden_)", names(cells_dt), value = TRUE)
cluster_cols <- cluster_cols[order(as.numeric(sub("^(RNA_snn_res\\.|leiden_)", "", cluster_cols)))]

if (!all(c("cluster_col", "cluster_resolution", "cluster", "top_label", "cs") %in% names(cluster_dt)) && length(cluster_cols) > 0) {
  cluster_dt <- rbindlist(lapply(cluster_cols, function(cluster_col) {
    dt <- cells_dt[!is.na(get(cluster_col)) & nzchar(as.character(get(cluster_col))), .(
      cluster = as.character(get(cluster_col)),
      label = as.character(label),
      score = as.numeric(score)
    )]
    if (nrow(dt) == 0) {
      return(NULL)
    }
    counts_dt <- dt[, .N, by = .(cluster, label)]
    totals_dt <- dt[, .(
      n_cells = .N,
      mean_score = mean(score, na.rm = TRUE)
    ), by = cluster]
    summary_dt <- merge(counts_dt, totals_dt, by = "cluster", all.x = TRUE, sort = FALSE)
    summary_dt[, cs := N / n_cells]
    setorder(summary_dt, cluster, -cs, -N, label)
    summary_dt <- summary_dt[, .SD[1], by = cluster]
    summary_dt[, `:=`(
      cluster_col = cluster_col,
      cluster_resolution = sub("^(RNA_snn_res\\.|leiden_)", "", cluster_col),
      top_label = label
    )]
    summary_dt[, .(cluster_col, cluster_resolution, cluster, n_cells, top_label, cs, mean_score)]
  }), use.names = TRUE, fill = TRUE)
}

if (nrow(cluster_dt) > 0) {
  cluster_dt[, `:=`(
    cluster_col = as.character(cluster_col),
    cluster_resolution = as.character(cluster_resolution),
    cluster = as.character(cluster),
    top_label = as.character(top_label)
  )]
}
```

This report summarises one annotation run.

* **Method id:** `r method_id`
* **Engine:** `r engine`
* **Reference:** `r reference_name`
* **Reference label column:** `r if (nzchar(reference_label_col)) reference_label_col else "not specified"`
* **Annotated cells:** `r scales::comma(nrow(cells_dt))`
* **Assigned labels:** `r data.table::uniqueN(cells_dt$label)`
* **Cluster score (`cs`):** within-cluster fraction assigned to the top predicted label

## Label counts

```{r label_counts, fig.width=8, fig.height=5}
label_counts_dt <- cells_dt[, .N, by = label][order(-N)]

ggplot(label_counts_dt, aes(x = reorder(label, N), y = N)) +
  geom_col(fill = "#2C7FB8", color = "black", linewidth = 0.2) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank()) +
  labs(x = NULL, y = "cells", title = "Predicted label counts")
```

## Prediction score distribution

```{r score_density, fig.width=8, fig.height=4}
if ("score" %in% names(cells_dt) && any(!is.na(cells_dt$score))) {
  ggplot(cells_dt[!is.na(score)], aes(x = score)) +
    geom_histogram(bins = 40, fill = "#7FCDBB", color = "white") +
    theme_bw() +
    labs(x = "prediction score", y = "cells", title = "Prediction score distribution")
}
```

## UMAP by predicted label

```{r umap_labels, fig.width=9, fig.height=7}
if (all(c("UMAP1", "UMAP2") %in% names(cells_dt))) {
  plot_dt <- copy(cells_dt)[!is.na(UMAP1) & !is.na(UMAP2)]
  label_levels <- plot_dt[, .N, by = label][order(-N)]$label
  plot_dt[, label := factor(label, levels = label_levels)]

  ggplot(plot_dt, aes(x = UMAP1, y = UMAP2, colour = label)) +
    ggrastr::geom_point_rast(size = 0.12, raster.dpi = 300) +
    theme_bw() +
    theme(panel.grid = element_blank(), aspect.ratio = 1) +
    labs(colour = "label", title = "UMAP coloured by predicted label")
}
```

## UMAP by prediction score

```{r umap_score, fig.width=7, fig.height=6}
if (all(c("UMAP1", "UMAP2", "score") %in% names(cells_dt))) {
  plot_dt <- copy(cells_dt)[!is.na(UMAP1) & !is.na(UMAP2)]

  ggplot(plot_dt, aes(x = UMAP1, y = UMAP2, colour = score)) +
    ggrastr::geom_point_rast(size = 0.12, raster.dpi = 300) +
    scale_colour_viridis_c(option = "C", na.value = "grey80") +
    theme_bw() +
    theme(panel.grid = element_blank(), aspect.ratio = 1) +
    labs(colour = "score", title = "UMAP coloured by prediction score")
}
```

## Cluster Prediction Summary

```{r cluster_umaps, fig.width=11, fig.height=max(5, ceiling(length(cluster_cols) / 2) * 4)}
if (length(cluster_cols) > 0 && all(c("UMAP1", "UMAP2") %in% names(cells_dt)) && nrow(cluster_dt) > 0) {
  plot_long_dt <- rbindlist(lapply(cluster_cols, function(cluster_col) {
    dt <- cells_dt[!is.na(UMAP1) & !is.na(UMAP2) & !is.na(get(cluster_col)) & nzchar(as.character(get(cluster_col))), .(
      UMAP1,
      UMAP2,
      cluster = as.character(get(cluster_col)),
      cluster_col = cluster_col,
      cluster_resolution = sub("^(RNA_snn_res\\.|leiden_)", "", cluster_col)
    )]
    dt
  }), use.names = TRUE, fill = TRUE)

  centroid_dt <- plot_long_dt[, .(
    UMAP1 = mean(UMAP1, na.rm = TRUE),
    UMAP2 = mean(UMAP2, na.rm = TRUE)
  ), by = .(cluster_col, cluster_resolution, cluster)]
  centroid_dt <- merge(centroid_dt, cluster_dt, by = c("cluster_col", "cluster_resolution", "cluster"), all.x = TRUE, sort = FALSE)
  centroid_dt[, label_text := sprintf("%s: cs = %.2f", top_label, cs)]

  ggplot(plot_long_dt, aes(x = UMAP1, y = UMAP2, colour = cluster)) +
    ggrastr::geom_point_rast(size = 0.09, raster.dpi = 300, show.legend = FALSE) +
    geom_text(
      data = centroid_dt,
      aes(label = label_text),
      inherit.aes = FALSE,
      size = 2.6,
      colour = "black",
      check_overlap = TRUE
    ) +
    facet_wrap(~ cluster_resolution, ncol = 2) +
    theme_bw() +
    theme(panel.grid = element_blank(), aspect.ratio = 1) +
    labs(title = "UMAPs by clustering resolution with top predicted cell type per cluster")
}
```

## Cluster Summary Table

```{r cluster_table}
cluster_dt[order(as.numeric(cluster_resolution), cluster)]
```