---
title: "Barcode_estimation_v2 diagnostics"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: false
execute:
  warning: false
  message: false
  cache: false
---

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(knitr)
})
source("barcode_v2_plots.R")

splice_context <- Sys.getenv("BARCODE_V2_SPLICE_CONTEXT", "snrna")
ed_fdr <- as.numeric(Sys.getenv("BARCODE_V2_ED_FDR", "0.001"))
```

```{r load_data}
audit_fs <- list.files(".", pattern = "^barcode_audit_v2_.+\\.csv\\.gz$", full.names = TRUE)
summary_fs <- list.files(".", pattern = "^barcode_summary_v2_.+\\.csv$", full.names = TRUE)

if (length(audit_fs) == 0) stop("No barcode_audit_v2_*.csv.gz files found")
if (length(summary_fs) == 0) stop("No barcode_summary_v2_*.csv files found")

audit_dt <- rbindlist(lapply(audit_fs, fread), fill = TRUE)
summary_dt <- rbindlist(lapply(summary_fs, fread), fill = TRUE)
setorder(summary_dt, sample_id)
sample_ids <- summary_dt$sample_id

splice_txt <- switch(
  splice_context,
  snrna = "lower spliced percentages are generally more consistent with nuclear RNA profiles; unusually high spliced fractions among called nuclei are potential concerns",
  scrna = "higher spliced percentages are generally more consistent with whole-cell RNA profiles; unusually low spliced fractions among called cells are potential concerns",
  "splice percentage is shown as a diagnostic only; no context-specific direction was applied"
)

overview_dt <- summary_dt[, .(
  sample_id,
  total_barcodes_input,
  zero_total_removed,
  total_droplets,
  retained_upper_knee,
  ambiguous_tested,
  called_by_emptydrops,
  excluded_low_count,
  excluded_emptydrops,
  final_cells,
  splice_context_outliers
)]
```

This report summarizes the standalone `Barcode_estimation_v2` caller. The caller separates barcodes into three zones before final calling:

- obvious good droplets retained above the upper knee;
- obvious bad droplets excluded below the low-count threshold; and
- an ambiguous middle population evaluated with `EmptyDrops`.

Spliced percentage is not used to make the call, but it is surfaced as a QC diagnostic. For the current run, `r splice_txt`.

## Audit table

```{r overview_table}
kable(overview_dt, caption = "Per-sample barcode calling audit summary")
```

## Threshold summary

```{r threshold_table}
kable(
  summary_dt[, .(
    sample_id,
    knee1,
    shin1,
    knee2,
    shin2,
    low_count_threshold,
    total_droplets_included,
    ed_fdr_threshold,
    splice_context
  )],
  caption = "Per-sample thresholds and calling parameters"
)
```

## Barcode rank classification

These plots show the three decision regions directly on the barcode-rank curve. The dashed blue line marks the upper-knee retain threshold, the dashed orange line marks the low-count exclusion threshold, and the dotted vertical line indicates the midpoint used to define the empty plateau.

```{r barcode_rank_panels, fig.height = 5, fig.width = 8, results='asis'}
cat('::: {.panel-tabset}\n\n')
for (sid in sample_ids) {
  cat("### ", sid, "\n\n")
  print(plot_barcode_rank_regions_v2(audit_dt[sample_id == sid]))
  cat("\n\n")
}
cat(':::\n\n')
```

## EmptyDrops on ambiguous droplets

Only the ambiguous middle region is interpreted through `EmptyDrops`. Droplets above the blue retain threshold are auto-retained; droplets below the orange threshold are excluded without testing.

```{r emptydrops_panels, fig.height = 5, fig.width = 7, results='asis'}
cat('::: {.panel-tabset}\n\n')
for (sid in sample_ids) {
  cat("### ", sid, "\n\n")
  print(plot_emptydrops_ambiguous_v2(audit_dt[sample_id == sid], ed_fdr = ed_fdr))
  cat("\n\n")
}
cat(':::\n\n')
```

## Spliced percentage diagnostics

These distributions are shown by final call reason. They are intended to help identify samples where the called population looks inconsistent with the expected assay context.

```{r splice_panels, fig.height = 5, fig.width = 7, results='asis'}
cat('::: {.panel-tabset}\n\n')
for (sid in sample_ids) {
  cat("### ", sid, "\n\n")
  print(plot_splice_diagnostics_v2(audit_dt[sample_id == sid]))
  cat("\n\n")
}
cat(':::\n\n')
```

## Session info

<details>
```{r session_info, echo=TRUE}
sessionInfo()
```
</details>