---
title: "HVG selection diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(SummarizedExperiment)
})
source("hvg_plots.R")

n_hvg <- as.integer(Sys.getenv("N_HVG", "4000"))
```

```{r load_data}
hvg_f <- "hvg_stats.csv.gz"
de_f  <- "edger_dt.csv.gz"
pb_f  <- "pb_empties.rds"

stopifnot("hvg_stats.csv.gz not found" = file.exists(hvg_f))
hvg_dt <- fread(hvg_f)
hvg_dt[, highly_variable := as.logical(highly_variable)]

# Try to load DE results and pseudobulk
has_de    <- file.exists(de_f)
has_pb    <- file.exists(pb_f)

if (has_de) {
    de_dt <- fread(de_f)
} else {
    de_dt <- NULL
}

if (has_pb) {
    pb_empty_se <- readRDS(pb_f)
} else {
    pb_empty_se <- NULL
}

n_total    <- nrow(hvg_dt)
n_selected <- sum(hvg_dt$highly_variable, na.rm = TRUE)
has_nbatches <- "highly_variable_nbatches" %in% names(hvg_dt)
```

## Overview

The selection of highly variable genes (HVGs) is a crucial step in integration and clustering.
This report evaluates HVG properties and their relationship to ambient RNA contamination.

- **Total genes**: `r n_total`
- **Selected HVGs**: `r n_selected` (top `r n_hvg`)
- **Ambient DE available**: `r ifelse(has_de, "Yes", "No")`
- **Pseudobulk empties available**: `r ifelse(has_pb, "Yes", "No")`

## Are highly variable genes "ambient"?

HVG selection was performed using scanpy's `seurat_v3` method with `batch_key = sample_id`,
selecting the top **`r n_hvg`** highly variable genes.

The plot below shows the mean-variance relationship used for HVG selection. Each point represents
a gene, colored by HVG selection status.

```{r plot_mean_var, fig.height=6, fig.width=8}
if (all(c("means", "variances_norm") %in% names(hvg_dt))) {
    plot_mean_variance(hvg_dt)
} else {
    cat("Columns 'means' or 'variances_norm' not present — skipping mean-variance plot.\n")
}
```

## Ambient Differential Expression {.tabset}

Differential expression analysis was performed on pseudobulk profiles:
- **Empty droplets**: aggregated counts from low-rank (ambient) barcodes
- **Cells**: aggregated counts from decontX-filtered cell barcodes

Ambient genes are defined as those significantly upregulated in empty droplets (FDR < 0.05, logFC > 0).

### Volcano plot

```{r plot_volcano, fig.height=6, fig.width=8}
if (!is.null(de_dt)) {
    de_dt[, is_ambient := FDR < 0.05 & logFC > 0]
    de_dt[, is_cell := FDR < 0.05 & logFC < 0]
    
    ggplot(de_dt, aes(x = logFC, y = -log10(PValue))) +
        geom_point(aes(color = is_ambient), size = 2, alpha = 0.6) +
        scale_color_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "#95A5A6"),
                           labels = c("TRUE" = "Ambient", "FALSE" = "Not significant")) +
        geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
        geom_vline(xintercept = 0, linetype = "solid", color = "gray50") +
        theme_minimal(base_size = 12) +
        theme(plot.title = element_text(face = "bold", size = 14)) +
        labs(title = "Empty droplets vs. Cells differential expression",
             x = "log2(FC)",
             y = "-log10(p-value)",
             color = "Gene type")
} else {
    cat("Ambient DE results not available.\n")
}
```

### Ambient genes summary

```{r ambient_genes_summary}
if (!is.null(de_dt)) {
    ambient_genes <- de_dt[FDR < 0.05 & logFC > 0, ]
    n_ambient <- nrow(ambient_genes)
    
    cat(sprintf("**Ambient genes (FDR<0.05, logFC>0)**: %d\n", n_ambient))
    
    if (n_ambient > 0) {
        # Check overlap with HVGs
        hvg_ids <- hvg_dt[highly_variable == TRUE, gene_id]
        ambient_in_hvg <- sum(ambient_genes$gene_id %in% hvg_ids)
        cat(sprintf("**Ambient genes that are HVGs**: %d (%.1f%%)\n",
                    ambient_in_hvg, 100 * ambient_in_hvg / n_ambient))
        
        # Top ambient genes
        top_ambient <- ambient_genes[order(-logFC)][seq_len(min(10, .N))]
        cat("\n**Top 10 ambient genes**:\n\n")
        knitr::kable(top_ambient[, .(symbol_ensembl, logFC, PValue, FDR)],
                     digits = 4,
                     col.names = c("Gene", "log2FC", "p-value", "FDR"))
    }
} else {
    cat("Ambient DE results not available.\n")
}
```

### Ambient genes ranked by logFC

```{r plot_ambient_logfc, fig.height=8, fig.width=8}
if (!is.null(de_dt)) {
    ambient_genes <- de_dt[FDR < 0.05 & logFC > 0, ]
    if (nrow(ambient_genes) > 0) {
        top_genes <- ambient_genes[order(-logFC)][seq_len(min(40, .N))]
        
        ggplot(top_genes, aes(x = reorder(symbol_ensembl, logFC), y = logFC)) +
            geom_bar(stat = "identity", fill = "#E74C3C", alpha = 0.7) +
            coord_flip() +
            theme_minimal(base_size = 11) +
            theme(plot.title = element_text(face = "bold", size = 12)) +
            labs(title = "Top ambient genes (ranked by log2FC)",
                 x = "Gene",
                 y = "log2(FC) — empty droplets vs. cells")
    } else {
        cat("No significant ambient genes found.\n")
    }
} else {
    cat("Ambient DE results not available.\n")
}
```

## HVG quality metrics {.tabset}

### HVG rank distribution

Distribution of scanpy HVG ranks among the **`r n_selected`** selected genes.
Genes with lower ranks were more consistently variable across batches.

```{r plot_rank, fig.height=5, fig.width=6}
if ("hvg_rank" %in% names(hvg_dt)) {
    plot_hvg_ranks(hvg_dt)
}
```

### Batch prevalence

For each selected HVG, the number of batches (samples) in which it was independently called as
highly variable. Genes consistently called in many batches are robust HVGs.

```{r plot_nbatches, fig.height=5, fig.width=6}
if (has_nbatches) {
    plot_hvg_nbatches(hvg_dt)
} else {
    cat("Batch-level HVG prevalence not available.\n")
}
```

### Top 50 HVGs

```{r top_hvgs_table}
top_hvg <- hvg_dt[highly_variable == TRUE][order(hvg_rank)][seq_len(min(50, .N))]
if (nrow(top_hvg) > 0) {
    cols <- intersect(c("gene_id", "hvg_rank", "highly_variable_nbatches",
                        "means", "variances", "variances_norm"),
                      names(top_hvg))
    knitr::kable(top_hvg[, ..cols], digits = 4,
                 caption = "Top 50 highly variable genes (ranked by HVG rank)")
}
```

### Highest variance genes

Top 40 genes by normalised variance, colored by HVG selection status.

```{r plot_top_var, fig.height=8, fig.width=7}
if ("variances_norm" %in% names(hvg_dt)) {
    plot_top_genes_bar(hvg_dt, rank_by = "variances_norm", n_top = 40)
}
```

### Highest mean expression genes

Top 40 genes by mean expression, colored by HVG selection status.

```{r plot_top_mean, fig.height=8, fig.width=7}
if ("means" %in% names(hvg_dt)) {
    plot_top_genes_bar(hvg_dt, rank_by = "means", n_top = 40)
}
```

## Pseudobulk ambient profiles

```{r pb_summary}
if (!is.null(pb_empty_se)) {
    cat(sprintf("**Pseudobulk empty counts**: %d genes × %d samples\n",
                nrow(pb_empty_se), ncol(pb_empty_se)))
    cat("**Samples**:", paste(colnames(pb_empty_se), collapse = ", "))
} else {
    cat("Pseudobulk empty droplet SE not available.\n")
}
```

## Session info

<details>
```{r session_info, echo=TRUE}
sessionInfo()
```
</details>

