---
title: "Integration"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
source("integration_plots.R")

metadata_vars_str <- Sys.getenv("METADATA_VARS", "")
leiden_res_str    <- Sys.getenv("LEIDEN_RES",    "0.3 0.5 1.0")

metadata_vars <- trimws(strsplit(metadata_vars_str, "\\s+")[[1]])
metadata_vars <- metadata_vars[nzchar(metadata_vars)]

leiden_res  <- trimws(strsplit(leiden_res_str, "\\s+")[[1]])
leiden_res  <- leiden_res[nzchar(leiden_res)]
leiden_cols <- paste0("leiden_", leiden_res)
```

```{r load_data}
int_f <- "integration_dt.csv.gz"
stopifnot("integration_dt.csv.gz not found" = file.exists(int_f))
int_dt  <- fread(int_f)
n_cells <- nrow(int_dt)

qc_files <- Sys.glob("qc_metrics_*.csv.gz")
qc_dt <- if (length(qc_files) > 0) rbindlist(lapply(qc_files, fread)) else data.table()

qc_names <- c("log_counts", "log_feats", "logit_mito", "logit_spliced")
qc_lu    <- c("no. of UMIs", "no. of genes", "mito pct.", "spliced pct.")
names(qc_lu) <- qc_names

present_leiden <- paste0("RNA_snn_res.", leiden_res)
present_leiden <- present_leiden[present_leiden %in% names(int_dt)]
if (length(present_leiden) == 0) {
  present_leiden <- paste0("leiden_", leiden_res)
  present_leiden <- present_leiden[present_leiden %in% names(int_dt)]
}
doublet_cols_present <- all(c("dbl_UMAP1", "dbl_UMAP2", "is_dbl", "dbl_cluster") %in% names(int_dt))
```

Cells that passed quality control filtering were integrated with Harmony correction where requested, then clustered at multiple resolutions and embedded with UMAP.

This report follows the scprocess integration diagnostics layout. The current scQC-flow integration path removes doublets before HVG selection and integration, so doublet-on-UMAP diagnostics are only shown if those fields are present in the integration output.

## Doublets over UMAP

* **Cells:** `r format(n_cells, big.mark = ",")`
* **Samples:** `r int_dt[, uniqueN(sample_id)]`
* **Metadata vars used for Harmony:** `r if (length(metadata_vars) > 0) paste(metadata_vars, collapse = ", ") else "none"`
* **Leiden resolutions:** `r paste(leiden_res, collapse = ", ")`

```{r doublet_umap, fig.height=5, fig.width=11}
if (doublet_cols_present) {
  dbl_dt <- int_dt[, .(cell_id, UMAP1 = dbl_UMAP1, UMAP2 = dbl_UMAP2, is_dbl, dbl_cluster)]
  plot_umap_doublets(dbl_dt) + plot_umap_density(dbl_dt[, .(cell_id, UMAP1, UMAP2)])
} else {
  cat("Doublet UMAP diagnostics are not available because the current scQC-flow integration path excludes doublets before integration and does not emit dbl_UMAP / dbl_cluster columns.\n")
}
```

## Doublet proportions in clusters

```{r doublet_clusters, fig.height=4, fig.width=6}
if (doublet_cols_present) {
  dbl_dt <- int_dt[, .(cell_id, UMAP1 = dbl_UMAP1, UMAP2 = dbl_UMAP2, is_dbl, dbl_cluster)]
  print(plot_doublet_clusters(dbl_dt, 0.5))
} else {
  cat("Doublet-enriched cluster diagnostics are not available for the current integration output.\n")
}
```

## Clusters over UMAP {.tabset}

```{r umap_clusters, fig.height=7, fig.width=9, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    print(plot_umap_cluster(int_dt, cl_col) + plot_umap_density(int_dt[, .(cell_id, UMAP1, UMAP2)]))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## Evaluating cluster distribution across samples {.tabset}

```{r cluster_entropy, fig.height=5, fig.width=7, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    clust_dt <- int_dt[, .(cell_id, sample_id, cluster = get(cl_col))]
    if (uniqueN(clust_dt$cluster) <= 1) next
    cat("### ", res_val, "\n\n")
    print(plot_cluster_entropies(clust_dt, batch_var = 'sample_id', what = 'norm'))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## Check QC metrics of clusters {.tabset}

```{r cluster_qc, fig.height=8, fig.width=10, results='asis'}
if (length(present_leiden) > 0 && nrow(qc_dt) > 0) {
  qc_melt <- melt(qc_dt[, c('cell_id', qc_names), with = FALSE], id.vars = 'cell_id',
                  variable.name = 'qc_var', value.name = 'qc_val')
  qc_melt[, qc_full := qc_lu[qc_var]]
  qc_melt[, qc_var := factor(qc_var, levels = qc_names)]
  qc_melt[, qc_full := fct_reorder(qc_full, as.integer(qc_var))]

  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    clust_dt <- int_dt[, .(cell_id, cluster = get(cl_col))]
    suppressWarnings(print(plot_cluster_qc_distns(qc_melt, clust_dt, name = sprintf('res = %s', res_val))))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## UMAP by sample

```{r umap_sample, fig.height=5, fig.width=7}
plot_umap_sample(int_dt)
```

## UMAP by metadata variables {.tabset}

```{r umap_metadata, fig.height=5, fig.width=7, results='asis'}
if (length(metadata_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (v in metadata_vars) {
    if (!v %in% names(int_dt)) next
    cat("### ", v, "\n\n")
    print(plot_umap_by_var(int_dt, v))
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  cat("*No metadata variables were provided.*\n\n")
}
```

## Cell counts per cluster {.tabset}

```{r cluster_counts, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    tbl <- cluster_counts_tbl(int_dt, cl_col)
    print(knitr::kable(tbl, caption = sprintf('Cell counts — res %s', res_val)))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## UMAP density

```{r umap_density, fig.height=5, fig.width=6}
plot_umap_density(int_dt)
```

## Session info

<details>
```{r session_info, echo=TRUE}
sessionInfo()
```
</details>
