---
title: "Integration diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
source("integration_plots.R")

leiden_res_str    <- Sys.getenv("LEIDEN_RES",    "0.2 0.5 1.0")

leiden_res  <- trimws(strsplit(leiden_res_str, "\\s+")[[1]])
leiden_res  <- leiden_res[nzchar(leiden_res)]

batch_var   <- "sample_id"
dbl_cl_prop <- 0.5
```

```{r load_data}
int_f <- "integration_dt.csv.gz"
stopifnot("integration_dt.csv.gz not found" = file.exists(int_f))
int_dt  <- fread(int_f)
n_cells <- nrow(int_dt)

qc_files <- Sys.glob("qc_metrics_*.csv.gz")
qc_dt <- if (length(qc_files) > 0) rbindlist(lapply(qc_files, fread)) else data.table()

if (nrow(qc_dt) > 0 && all(c("sample_id", "cell_id") %in% names(qc_dt))) {
  # Integration outputs globally unique cell IDs as <sample_id>_<barcode>.
  # Normalise QC IDs to the same format before joining for cluster QC plots.
  qc_dt[, cell_id := paste(sample_id, cell_id, sep = "_")]
}

qc_names <- c("log_counts", "log_feats", "logit_mito", "logit_spliced")
qc_lu    <- c("no. of UMIs", "no. of genes", "mito pct.", "spliced pct.")
names(qc_lu) <- qc_names

present_leiden <- paste0("RNA_snn_res.", leiden_res)
present_leiden <- present_leiden[present_leiden %in% names(int_dt)]
if (length(present_leiden) == 0) {
  present_leiden <- paste0("leiden_", leiden_res)
  present_leiden <- present_leiden[present_leiden %in% names(int_dt)]
}

# Pass-2 cells only (non-NA UMAP1): doublets removed in two-pass integration
# have NaN UMAP1/2 and NaN cluster columns in the outer-join output
int_dt_pass2 <- int_dt[!is.na(UMAP1)]

core_cols <- c(
  "cell_id", "sample_id", "embedding", "UMAP1", "UMAP2",
  "dbl_UMAP1", "dbl_UMAP2", "dbl_cluster", "dbl_prop", "in_dbl_cl",
  "is_dbl", "is_doublet"
)
cluster_cols <- grep("^(RNA_snn_res\\.|leiden_)", names(int_dt), value = TRUE)
embed_cols <- grep("^(hmny_pca|pca)_", names(int_dt), value = TRUE)
metadata_vars <- setdiff(names(int_dt), c(core_cols, cluster_cols, embed_cols))
metadata_vars <- metadata_vars[!startsWith(metadata_vars, "_")]
metadata_vars <- metadata_vars[vapply(metadata_vars, function(var_name) {
  vals <- int_dt_pass2[[var_name]]
  any(!is.na(vals)) && uniqueN(vals[!is.na(vals)]) > 1
}, logical(1))]

# Doublet annotations — two-pass format uses is_dbl; fall back to is_doublet
has_doublets <- "is_dbl" %in% names(int_dt) || "is_doublet" %in% names(int_dt)
if (has_doublets) {
  if ("is_dbl" %in% names(int_dt)) {
    int_dt[, is_dbl := as.logical(is_dbl)]
    n_doublets <- sum(int_dt$is_dbl, na.rm = TRUE)
  } else {
    int_dt[, is_doublet := as.logical(is_doublet)]
    int_dt[, is_dbl := is_doublet]
    n_doublets <- sum(int_dt$is_dbl)
  }
  n_singlets <- sum(!int_dt$is_dbl, na.rm = TRUE)
}

# Determine embedding type from column names
int_embedding <- if (any(grepl("^hmny_pca", names(int_dt)))) "harmony" else "pca"
int_txt <- if (int_embedding == "harmony") "with `Harmony` " else ""
```

Cells that passed quality control filtering were integrated `r int_txt`together with doublets identified using `scDblFinder`. Clustering of cells at a high resolution is then performed. Cells that appear in clusters which are enriched in doublets are excluded from further analysis.

* **Cells:** `r format(n_cells, big.mark = ",")`
`r if (has_doublets) sprintf("* **Singlets:** %s", format(n_singlets, big.mark = ","))`
`r if (has_doublets) sprintf("* **Doublets:** %s (%.1f%%)", format(n_doublets, big.mark = ","), 100 * n_doublets / (n_doublets + n_singlets))`
* **Samples:** `r int_dt[, uniqueN(sample_id)]`
* **Metadata vars retained in integration output:** `r if (length(metadata_vars) > 0) paste(metadata_vars, collapse = ", ") else "none"`
* **Leiden resolutions:** `r paste(leiden_res, collapse = ", ")`

## Doublets over UMAP

The plot displays a binned UMAP with the proportion of doublets as well as the number of doublets in each bin.

```{r plot_doublet_integration, fig.height = 5, fig.width = 11}
if (has_doublets && n_doublets > 0) {
  g_dbl <- plot_umap_doublets(int_dt)
  # Use pass-1 embedding for density (all cells incl. doublets) when available
  if (all(c("dbl_UMAP1", "dbl_UMAP2") %in% names(int_dt))) {
    dens_dt <- int_dt[!is.na(dbl_UMAP1), .(cell_id, UMAP1 = dbl_UMAP1, UMAP2 = dbl_UMAP2)]
  } else {
    dens_dt <- int_dt[, .(cell_id, UMAP1, UMAP2)]
  }
  g_dens <- plot_umap_density(dens_dt)
  g_dbl + g_dens
} else {
  cat("No doublet annotations available.\n")
}
```

## Doublet proportions in clusters

The plot shows the proportion of doublets for each cluster in relation to the total number of cells in that cluster. Clusters with a doublet proportion exceeding `r dbl_cl_prop*100`% are excluded from further analysis.

```{r plot_doublet_clusters, fig.height = 4, fig.width = 6, results='asis'}
has_dbl_cluster_col <- "dbl_cluster" %in% names(int_dt)
if (has_doublets && n_doublets > 0 && has_dbl_cluster_col) {
  # Two-pass format: dbl_cluster and dbl_prop are precomputed per cell
  plot_doublet_clusters(int_dt, dbl_cl_prop)
} else if (has_doublets && n_doublets > 0 && length(present_leiden) > 0) {
  # Legacy fallback
  plot_doublet_clusters(int_dt, dbl_cl_prop, cl_col = present_leiden[length(present_leiden)])
} else {
  cat("No doublet annotations available for cluster analysis.\n")
}
```

## Clusters over UMAP

Clustering of data is performed at different resolution values. For each value the clusters are displayed over a UMAP together with a plot showing the density of cells.

After removing doublets and doublet-enriched cluster cells, there were **`r if (has_doublets) format(n_singlets, big.mark = ",") else format(n_cells, big.mark = ",")` cells used for pass-2 integration**.

```{r plot_clusters_integration, fig.height = 7, fig.width = 9, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    g_cl   <- plot_umap_cluster(
      umap_dt  = int_dt_pass2[, .(cell_id, UMAP1, UMAP2)],
      clust_dt = int_dt_pass2[, .(cell_id, cluster = get(cl_col))],
      name     = sprintf('res = %s', res_val))
    g_dens <- plot_umap_density(int_dt_pass2[, .(cell_id, UMAP1, UMAP2)])
    print(g_cl + g_dens)
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## Evaluating cluster distribution across samples

This plot visualizes the relationship between the entropy of clusters and the maximum proportion of cells from a single sample within each cluster (higher values suggest that a cluster predominantly contains cells from a single sample).
Entropy measures how evenly distributed cells are across samples within each cluster—higher entropy indicates that cells from different samples are more evenly distributed, while lower entropy suggests that a cluster is dominated by cells from a small number of samples.

```{r plot_entropy_selected_clusters, fig.height = 5, fig.width = 7, results='asis'}
if (length(present_leiden) > 0) {
  n_clusts <- sapply(present_leiden, function(cl_col) {
    int_dt_pass2[, uniqueN(get(cl_col))]
  })
  res_ls_pl <- present_leiden[n_clusts > 1]
  if (length(res_ls_pl) > 0) {
    cat('::: {.panel-tabset}\n\n')
    for (cl_col in res_ls_pl) {
      res_val <- sub('^RNA_snn_res\\.', '', cl_col)
      res_val <- sub('^leiden_', '', res_val)
      cat("### ", res_val, "\n\n")
      input_dt <- int_dt_pass2[, .(batch_var = get(batch_var), cell_id,
        cluster = get(cl_col))]
      input_dt[, cluster := as.factor(cluster)]
      suppressWarnings(print(plot_cluster_entropies(input_dt, 'batch_var', what = "norm")))
      cat("\n\n")
    }
    cat(':::\n\n')
  } else {
    cat("No cluster entropy panels available because no resolution produced more than one cluster.\n\n")
  }
}
```

## Check QC metrics of clusters

Distributions of QC metrics (library size, number of features, mitochondrial proportion, and spliced proportion) are shown for each cluster across different resolution values.

```{r plot_qc_distns, fig.height = 8, fig.width = 10, results='asis'}
if (length(present_leiden) > 0 && nrow(qc_dt) > 0) {
  qc_melt <- melt(qc_dt[, c('cell_id', qc_names), with = FALSE], id.vars = 'cell_id',
                  variable.name = 'qc_var', value.name = 'qc_val')
  qc_melt[, qc_full := qc_lu[qc_var]]
  qc_melt[, qc_var := factor(qc_var, levels = qc_names)]
  qc_melt[, qc_full := fct_reorder(qc_full, as.integer(qc_var))]
  qc_melt <- qc_melt[!is.na(qc_val)]

  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    suppressWarnings(print(plot_cluster_qc_distns(
      qc_melt,
      clust_dt = int_dt_pass2[, .(cell_id, cluster = get(cl_col))],
      name     = as.numeric(res_val))))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## UMAP by sample

```{r umap_sample, fig.height = 5, fig.width = 7}
plot_umap_sample(int_dt_pass2)
```

## UMAP by metadata variables

```{r umap_metadata, fig.height = 5, fig.width = 7, results='asis'}
if (length(metadata_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in metadata_vars) {
    cat("### ", var_name, "\n\n")
    print(plot_umap_by_var(int_dt_pass2, var_name))
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  cat("No metadata variables were retained in the integration output.\n\n")
}
```

## Cell counts per cluster

```{r cluster_counts, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    tbl <- cluster_counts_tbl(int_dt_pass2, cl_col)
    print(knitr::kable(tbl, caption = sprintf('Cell counts - res %s', res_val)))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## Session info

<details>
```{r session_info, echo=TRUE}
sessionInfo()
```
</details>
