---
title: "Integration diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
knitr::opts_chunk$set(fig.path = 'plots/', dev = 'png', dpi = 150)
source("integration_plots.R")

leiden_res_str    <- Sys.getenv("LEIDEN_RES",    "0.2 0.5 1.0")
metadata_vars_str <- Sys.getenv("METADATA_VARS", "")

leiden_res  <- trimws(strsplit(leiden_res_str, "\\s+")[[1]])
leiden_res  <- leiden_res[nzchar(leiden_res)]
requested_metadata_vars <- trimws(strsplit(metadata_vars_str, "\\s+")[[1]])
requested_metadata_vars <- requested_metadata_vars[nzchar(requested_metadata_vars)]

batch_var   <- "sample_id"
dbl_cl_prop <- as.numeric(Sys.getenv("DBL_CL_PROP", "0.5"))
message("New report!")
qc_feature_palette <- c("#fff7ec", "#fdd49e", "#fdbb84", "#ef6548", "#990000")

plot_umap_qc_feature <- function(int_dt, var_name, label) {
  stopifnot(all(c("UMAP1", "UMAP2", var_name) %in% names(int_dt)))

  plot_dt <- copy(int_dt)[!is.na(get(var_name)), c("UMAP1", "UMAP2", var_name), with = FALSE]
  plot_dt[, UMAP1 := rescale(UMAP1, to = c(0.05, 0.95))]
  plot_dt[, UMAP2 := rescale(UMAP2, to = c(0.05, 0.95))]
  plot_dt <- plot_dt[sample(.N, .N)]

  ggplot(plot_dt, aes(x = UMAP1, y = UMAP2, colour = get(var_name))) +
    geom_point(size = 0.1) +
    scale_x_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    scale_y_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    scale_colour_gradientn(colours = qc_feature_palette, na.value = "grey85") +
    theme_bw() +
    theme(panel.grid = element_blank(), aspect.ratio = 1,
      legend.title.position = "left", legend.position = "bottom",
      axis.ticks = element_blank(), axis.text = element_blank()) +
    labs(colour = label)
}
```

```{r load_data}
int_f <- "integration_dt.csv.gz"
stopifnot("integration_dt.csv.gz not found" = file.exists(int_f))
int_dt  <- fread(int_f)
n_cells <- nrow(int_dt)

qc_files <- Sys.glob("qc_metrics_*.csv.gz")
qc_dt <- if (length(qc_files) > 0) rbindlist(lapply(qc_files, fread)) else data.table()

if (nrow(qc_dt) > 0 &&
    !"splice_frac" %in% names(qc_dt) &&
    all(c("total_spliced", "total_unspliced") %in% names(qc_dt))) {
  qc_dt[, splice_frac := (total_spliced + 1) / (total_spliced + total_unspliced + 2)]
} else if (nrow(qc_dt) > 0 &&
           !"splice_frac" %in% names(qc_dt) &&
           "logit_spliced" %in% names(qc_dt)) {
  qc_dt[, splice_frac := plogis(logit_spliced)]
}

if (nrow(qc_dt) > 0 && all(c("sample_id", "cell_id") %in% names(qc_dt))) {
  # Integration outputs globally unique cell IDs as <sample_id>_<barcode>.
  # Normalise QC IDs to the same format before joining for cluster QC plots.
  qc_dt[, cell_id := paste(sample_id, cell_id, sep = "_")]
}

feature_sources <- list(
  list(key = "ncount_rna",   source = "sum",         label = "nCount RNA"),
  list(key = "nfeature_rna", source = "detected",    label = "nFeature RNA"),
  list(key = "splice_frac",  source = "splice_frac", label = "splice fraction"),
  list(key = "mito_pct",     source = "mito_pct",    label = "mitochondrial content")
)

qc_names <- c("log_counts", "log_feats", "logit_mito", "logit_spliced")
qc_lu    <- c("no. of UMIs", "no. of genes", "mito pct.", "spliced pct.")
names(qc_lu) <- qc_names

present_leiden <- paste0("RNA_snn_res.", leiden_res)
present_leiden <- present_leiden[present_leiden %in% names(int_dt)]
if (length(present_leiden) == 0) {
  present_leiden <- paste0("leiden_", leiden_res)
  present_leiden <- present_leiden[present_leiden %in% names(int_dt)]
}

# Pass-2 cells only (non-NA UMAP1): doublets removed in two-pass integration
# have NaN UMAP1/2 and NaN cluster columns in the outer-join output
int_dt_pass2 <- int_dt[!is.na(UMAP1)]

if (nrow(qc_dt) > 0) {
  feature_cols <- unique(unlist(lapply(feature_sources, `[[`, "source")))
  present_feature_cols <- feature_cols[feature_cols %in% names(qc_dt)]
  if (length(present_feature_cols) > 0) {
    feature_dt <- unique(qc_dt[, c("cell_id", present_feature_cols), with = FALSE], by = "cell_id")
    int_dt_pass2 <- merge(int_dt_pass2, feature_dt, by = "cell_id", all.x = TRUE)
  }
}

core_cols <- c(
  "cell_id", "sample_id", "embedding", "UMAP1", "UMAP2",
  "dbl_UMAP1", "dbl_UMAP2", "dbl_cluster", "dbl_prop", "in_dbl_cl",
  "is_dbl", "is_doublet"
)
cluster_cols <- grep("^(RNA_snn_res\\.|leiden_)", names(int_dt), value = TRUE)
embed_cols <- grep("^(hmny_pca|pca)_", names(int_dt), value = TRUE)
metadata_vars <- setdiff(names(int_dt), c(core_cols, cluster_cols, embed_cols))
metadata_vars <- metadata_vars[!startsWith(metadata_vars, "_")]
metadata_vars <- metadata_vars[vapply(metadata_vars, function(var_name) {
  vals <- int_dt_pass2[[var_name]]
  any(!is.na(vals))
}, logical(1))]
if (length(requested_metadata_vars) > 0) {
  metadata_vars <- c(
    requested_metadata_vars[requested_metadata_vars %in% metadata_vars],
    setdiff(metadata_vars, requested_metadata_vars)
  )
}
umap_vars <- unique(c("sample_id", metadata_vars))
umap_vars <- umap_vars[umap_vars %in% names(int_dt_pass2)]
metadata_umap_vars <- setdiff(umap_vars, "sample_id")
bar_vars <- umap_vars[vapply(umap_vars, function(var_name) {
  vals <- int_dt_pass2[[var_name]]
  vals <- vals[!is.na(vals)]
  if (length(vals) == 0) {
    return(FALSE)
  }
  !is.numeric(vals) || uniqueN(vals) <= 12
}, logical(1))]
metadata_dens_vars <- setdiff(bar_vars, "sample_id")
metadata_dens_vars <- metadata_dens_vars[vapply(metadata_dens_vars, function(var_name) {
  vals <- int_dt_pass2[[var_name]]
  vals <- vals[!is.na(vals)]
  data.table::uniqueN(vals) > 1
}, logical(1))]
if (length(metadata_dens_vars) > 0) {
  sel_cl_int_dt <- copy(int_dt_pass2)[, c("cell_id", "UMAP1", "UMAP2", metadata_dens_vars), with = FALSE]
}
feature_vars <- Filter(function(spec) spec$source %in% names(int_dt_pass2), feature_sources)
missing_feature_vars <- Filter(function(spec) !spec$source %in% names(int_dt_pass2), feature_sources)

# Doublet annotations — two-pass format uses is_dbl; fall back to is_doublet
has_doublets <- "is_dbl" %in% names(int_dt) || "is_doublet" %in% names(int_dt)
if (has_doublets) {
  if ("is_dbl" %in% names(int_dt)) {
    int_dt[, is_dbl := as.logical(is_dbl)]
    n_doublets <- sum(int_dt$is_dbl, na.rm = TRUE)
  } else {
    int_dt[, is_doublet := as.logical(is_doublet)]
    int_dt[, is_dbl := is_doublet]
    n_doublets <- sum(int_dt$is_dbl)
  }
  n_singlets <- sum(!int_dt$is_dbl, na.rm = TRUE)
}

# Determine embedding type from column names
int_embedding <- if (any(grepl("^hmny_pca", names(int_dt)))) "harmony" else "pca"
int_txt <- if (int_embedding == "harmony") "with `Harmony` " else ""
```

Cells that passed quality control filtering were integrated `r int_txt`together with doublets identified using `scDblFinder`. Clustering of cells at a high resolution is then performed. Cells that appear in clusters which are enriched in doublets are excluded from further analysis.

* **Cells:** `r format(n_cells, big.mark = ",")`
`r if (has_doublets) sprintf("* **Singlets:** %s", format(n_singlets, big.mark = ","))`
`r if (has_doublets) sprintf("* **Doublets:** %s (%.1f%%)", format(n_doublets, big.mark = ","), 100 * n_doublets / (n_doublets + n_singlets))`
* **Samples:** `r data.table::uniqueN(int_dt$sample_id)`
* **Metadata vars retained in integration output:** `r if (length(metadata_vars) > 0) paste(metadata_vars, collapse = ", ") else "none"`
* **Leiden resolutions:** `r paste(leiden_res, collapse = ", ")`

## Doublets over UMAP

The plot displays a binned UMAP with the proportion of doublets as well as the number of doublets in each bin.

```{r plot_doublet_integration, fig.height = 5, fig.width = 11}
if (has_doublets && n_doublets > 0) {
  g_dbl <- plot_umap_doublets(int_dt)
  # Use pass-1 embedding for density (all cells incl. doublets) when available
  if (all(c("dbl_UMAP1", "dbl_UMAP2") %in% names(int_dt))) {
    dens_dt <- int_dt[!is.na(dbl_UMAP1), .(cell_id, UMAP1 = dbl_UMAP1, UMAP2 = dbl_UMAP2)]
  } else {
    dens_dt <- int_dt[, .(cell_id, UMAP1, UMAP2)]
  }
  g_dens <- plot_umap_density(dens_dt)
  g_dbl + g_dens
} else {
  cat("No doublet annotations available.\n")
}
```

## Doublet proportions in clusters

The plot shows the proportion of doublets for each cluster in relation to the total number of cells in that cluster. Clusters with a doublet proportion exceeding `r dbl_cl_prop*100`% are excluded from further analysis.

```{r plot_doublet_clusters, fig.height = 4, fig.width = 6, results='asis'}
has_dbl_cluster_col <- "dbl_cluster" %in% names(int_dt)
if (has_doublets && n_doublets > 0 && has_dbl_cluster_col) {
  # Two-pass format: dbl_cluster and dbl_prop are precomputed per cell
  plot_doublet_clusters(int_dt, dbl_cl_prop)
} else if (has_doublets && n_doublets > 0 && length(present_leiden) > 0) {
  # Legacy fallback
  plot_doublet_clusters(int_dt, dbl_cl_prop, cl_col = present_leiden[length(present_leiden)])
} else {
  cat("No doublet annotations available for cluster analysis.\n")
}
```

## Doublet proportion sweep

Pass-1 Leiden clustering is repeated at multiple resolutions. The plot below shows the distribution of doublet proportions across all clusters at each resolution, with the removal threshold marked in red. Clusters that are stably elevated above the threshold across resolutions are more likely to represent genuine doublet-enriched populations.

```{r load_dbl_sweep}
sweep_f <- "dbl_sweep.csv.gz"
has_sweep <- file.exists(sweep_f)
if (has_sweep) {
  sweep_dt <- fread(sweep_f)
  sweep_dt[, resolution := factor(resolution)]
}
```

```{r plot_dbl_sweep, fig.height = 5, fig.width = 9}
if (has_sweep && exists("sweep_dt") && nrow(sweep_dt) > 0) {
  ggplot(sweep_dt, aes(x = resolution, y = dbl_prop)) +
    geom_hline(yintercept = dbl_cl_prop, colour = "red", linetype = "dashed", linewidth = 0.7) +
    geom_violin(fill = "steelblue", alpha = 0.45, scale = "width", colour = NA) +
    geom_jitter(aes(size = n_cells), width = 0.18, alpha = 0.55, colour = "steelblue4") +
    scale_size_continuous(name = "cluster size", range = c(1, 5),
                          labels = scales::comma) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       limits = c(0, NA)) +
    labs(
      x = "Leiden resolution (pass 1)",
      y = "Doublet proportion per cluster",
      caption = sprintf("Red dashed line = removal threshold (%.0f%%)", dbl_cl_prop * 100)
    ) +
    theme_bw() +
    theme(panel.grid.minor = element_blank())
} else {
  cat("Doublet proportion sweep data not available.\n")
}
```

## Clusters over UMAP

Clustering of data is performed at different resolution values. For each value the clusters are displayed over a UMAP together with a plot showing the density of cells.

After removing doublets and doublet-enriched cluster cells, there were **`r if (has_doublets) format(n_singlets, big.mark = ",") else format(n_cells, big.mark = ",")` cells used for pass-2 integration**.

```{r plot_clusters_integration, fig.height = 7, fig.width = 9, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    g_cl   <- plot_umap_cluster(
      umap_dt  = int_dt_pass2[, .(cell_id, UMAP1, UMAP2)],
      clust_dt = int_dt_pass2[, .(cell_id, cluster = get(cl_col))],
      name     = sprintf('res = %s', res_val))
    g_dens <- plot_umap_density(int_dt_pass2[, .(cell_id, UMAP1, UMAP2)])
    print(g_cl + g_dens)
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## Evaluating cluster distribution across samples

This plot visualizes the relationship between the entropy of clusters and the maximum proportion of cells from a single sample within each cluster (higher values suggest that a cluster predominantly contains cells from a single sample).
Entropy measures how evenly distributed cells are across samples within each cluster—higher entropy indicates that cells from different samples are more evenly distributed, while lower entropy suggests that a cluster is dominated by cells from a small number of samples.

```{r plot_entropy_selected_clusters, fig.height = 5, fig.width = 7, results='asis'}
if (length(present_leiden) > 0) {
  n_clusts <- sapply(present_leiden, function(cl_col) {
    int_dt_pass2[, data.table::uniqueN(get(cl_col))]
  })
  res_ls_pl <- present_leiden[n_clusts > 1]
  if (length(res_ls_pl) > 0) {
    cat('::: {.panel-tabset}\n\n')
    for (cl_col in res_ls_pl) {
      res_val <- sub('^RNA_snn_res\\.', '', cl_col)
      res_val <- sub('^leiden_', '', res_val)
      cat("### ", res_val, "\n\n")
      input_dt <- int_dt_pass2[, .(batch_var = get(batch_var), cell_id,
        cluster = get(cl_col))]
      input_dt[, cluster := as.factor(cluster)]
      suppressWarnings(print(plot_cluster_entropies(input_dt, 'batch_var', what = "norm")))
      cat("\n\n")
    }
    cat(':::\n\n')
  } else {
    cat("No cluster entropy panels available because no resolution produced more than one cluster.\n\n")
  }
}
```

## Check QC metrics of clusters

Distributions of QC metrics (library size, number of features, mitochondrial proportion, and spliced proportion) are shown for each cluster across different resolution values.

```{r plot_qc_distns, fig.height = 8, fig.width = 10, results='asis'}
if (length(present_leiden) > 0 && nrow(qc_dt) > 0) {
  qc_melt <- melt(qc_dt[, c('cell_id', qc_names), with = FALSE], id.vars = 'cell_id',
                  variable.name = 'qc_var', value.name = 'qc_val')
  qc_melt[, qc_full := qc_lu[qc_var]]
  qc_melt[, qc_var := factor(qc_var, levels = qc_names)]
  qc_melt[, qc_full := fct_reorder(qc_full, as.integer(qc_var))]
  qc_melt <- qc_melt[!is.na(qc_val)]

  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    suppressWarnings(print(plot_cluster_qc_distns(
      qc_melt,
      clust_dt = int_dt_pass2[, .(cell_id, cluster = get(cl_col))],
      name     = as.numeric(res_val))))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## UMAP by sample and metadata variables

```{r umap_metadata, fig.height = 5, fig.width = 7, results='asis'}
if (length(umap_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in umap_vars) {
    cat("### ", var_name, "\n\n")
    print(plot_umap_by_var(int_dt_pass2, var_name))
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  cat("No sample or metadata variables were retained in the integration output.\n\n")
}
```

## UMAP by binned metadata variables

```{r umap_metadata_binned, fig.height = 5, fig.width = 7, results='asis'}
if (length(metadata_dens_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in metadata_dens_vars) {
    cat("### ", var_name, "\n\n")
    print(plot_clusters_annotated_by_densities(sel_cl_int_dt, var_name))
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  cat("No discrete metadata variables were retained in the integration output.\n\n")
}
```

## Barcharts

Stacked bar charts below show the composition of each cluster after pass-2 integration. Each bar is normalised within a cluster, so the bar height always sums to 100%.

```{r cluster_barcharts, fig.height = max(4, 2.3 * length(present_leiden)), fig.width = 12, results='asis'}
if (length(present_leiden) > 0 && length(bar_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in bar_vars) {
    cat("### ", var_name, "\n\n")
    print(plot_cluster_composition_bars(int_dt_pass2, present_leiden, var_name))
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  cat("No categorical sample or metadata variables were available for stacked cluster bar charts.\n\n")
}
```

## UMAP by QC features

```{r umap_features, fig.height = 5, fig.width = 7, results='asis'}
if (length(feature_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (spec in feature_vars) {
    cat("### ", spec$label, "\n\n")
    print(plot_umap_qc_feature(int_dt_pass2, spec$source, spec$label))
    cat("\n\n")
  }
  for (spec in missing_feature_vars) {
    cat("### ", spec$label, "\n\n")
    cat(sprintf("%s is not available in the current integration report inputs.\n\n", spec$label))
  }
  cat(':::\n\n')
} else {
  cat("No QC feature variables were available for UMAP plotting.\n\n")
}
```

## Cell counts per cluster

```{r cluster_counts, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    tbl <- cluster_counts_tbl(int_dt_pass2, cl_col)
    print(knitr::kable(tbl, caption = sprintf('Cell counts - res %s', res_val)))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```


## Session info

<details>
```{r session_info, echo=TRUE}
sessionInfo()
```
</details>
