---
title: "Mapping + cell calling (knee/shin)"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{=html}
<style>.quarto-figure img,.cell-output-display img,img.img-fluid{max-width:min(100%,860px);height:auto}</style>
```

Read mapping (`piscem`/`alevin-fry`) and cell calling for one sample batch,
both reported together since cell calling is only ever interpreted against how
cleanly its input reads mapped. Cells are called by the **classic barcode-rank
(knee/shin)** method that originally fed decontX (decontX itself is not run
here -- the same selection is replicated exactly, matching scprocess's own
`call_cells_and_empties()`, then decontX's correction step is skipped since
CellSweep runs downstream instead). Three buckets:

- **cell** -- top `expected_cells` barcodes by UMI rank
- **ambient** -- `in_empty_plateau` barcodes (from knee/shin estimation); this is
  the background CellSweep uses for decontamination
- **excluded** -- everything else, never used for anything

Splice-blind. See the CellSweep-inputs report (rendered later, right before CellSweep
runs) for how the `cell` bucket further splits into QC-kept vs QC-removed.

```{r setup}
knitr::opts_chunk$set(fig.path = 'plots/', dev = 'png', dpi = 110)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(patchwork)
  library(magrittr)
})
ps <- 10
```

```{r load_data}
stats_fs <- list.files(".", pattern = "^alevinfry_stats_.+\\.csv$", full.names = TRUE)
af_sample_stats <- rbindlist(lapply(stats_fs, fread), fill = TRUE)
af_sample_stats[, `:=`(
  short_id       = sub("_snRNAseq$", "", sub("^LM[0-9]+_", "", sample_id)),
  reads_m        = round(num_reads / 1e6, 1),
  percent_mapped = round(percent_mapped, 1),
  saturation_pct = round(100 * saturation, 1),
  barcodes_k     = round(n_barcodes_above_floor / 1e3, 1),
  runtime_min    = round(runtime_seconds / 60, 1)
)]
setorder(af_sample_stats, -sample_id)
af_sample_stats[, short_id := factor(short_id, levels = short_id)]

sum_fs <- list.files(".", pattern="^knee_summary_.+\\.csv$", full.names=TRUE)
lab_fs <- list.files(".", pattern="^knee_labels_.+\\.csv\\.gz$", full.names=TRUE)
names(lab_fs) <- sub("\\.csv\\.gz$","",sub("^.*/knee_labels_","",lab_fs))
sids <- sort(names(lab_fs))

# Paginate the per-sample tabset into chunks so large cohorts stay navigable
# (nested chunk tabs -> per-sample sub-tabs). render_one(sid) emits its own
# '#### <sid>' header + body. Single-chunk cohorts skip the outer tab layer.
page_size <- 30L
render_sample_chunks <- function(ids, render_one) {
  chunks <- split(ids, ceiling(seq_along(ids) / page_size))
  multi  <- length(chunks) > 1L
  if (multi) cat('::: {.panel-tabset}\n\n')
  for (ci in seq_along(chunks)) {
    if (multi) cat(sprintf("### Samples %d–%d\n\n",
                           (ci - 1L) * page_size + 1L,
                           (ci - 1L) * page_size + length(chunks[[ci]])))
    cat('::: {.panel-tabset}\n\n')
    for (sid in chunks[[ci]]) { render_one(sid); cat("\n\n") }
    cat(':::\n\n')
  }
  if (multi) cat(':::\n\n')
}
summ <- rbindlist(lapply(sum_fs, fread), fill=TRUE)
summ[, short_id := sub("_snRNAseq$", "", sub("^LM[0-9]+_", "", sample_id))]
setorder(summ, -sample_id)
summ[, short_id := factor(short_id, levels = short_id)]

pop_cols <- c(cell="grey80", ambient="#FB8072", excluded="grey30")
LOGB <- log10(ps + c(0,10,30,100,300,1000,3000,1e4,3e4,1e5,3e5))
LOGL <- c("0","10","30","100","300","1k","3k","10k","30k","100k","300k")
PCTB <- qlogis(c(0.001,0.003,0.01,0.03,0.1,0.5,0.9,0.97,0.99,0.997,0.999))
PCTL <- c("0.1%","0.3%","1%","3%","10%","50%","90%","97%","99%","99.7%","99.9%")
```

```{r helpers}
# One boxed, horizontal-bar-with-numbers panel per metric, sharing the sample
# axis (facet_grid + coord_flip).
plot_metric_bars <- function(df, sample_var, metric_cols, metric_labs, fill = "#2c8a3d") {
  plot_dt <- df[, c(sample_var, metric_cols), with = FALSE] %>%
    melt(id.vars = sample_var, variable.name = "metric", value.name = "value") %>%
    .[, metric := factor(metric_labs[metric], levels = metric_labs)]

  ggplot(plot_dt, aes(x = .data[[sample_var]], y = value)) +
    geom_col(fill = fill, width = 0.75) +
    geom_text(aes(label = value), hjust = -0.15, size = 2.6) +
    facet_grid(. ~ metric, scales = "free_x") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.35)), n.breaks = 4) +
    coord_flip() +
    theme_bw(base_size = 9) +
    theme(
      panel.border     = element_rect(colour = "grey35", fill = NA),
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "grey90", colour = "grey35"),
      strip.text       = element_text(face = "bold", size = 7.5)
    ) +
    labs(x = NULL, y = NULL)
}

# Splits samples into <=max_per_page pages, tabbing if there's more than one.
paginate_and_plot <- function(df, sample_var, max_per_page, plot_fn) {
  ids    <- levels(df[[sample_var]])
  groups <- split(ids, ceiling(seq_along(ids) / max_per_page))

  if (length(groups) == 1) {
    print(plot_fn(df))
  } else {
    cat("::: {.panel-tabset}\n\n")
    for (i in seq_along(groups)) {
      g <- groups[[i]]
      cat(sprintf("### Samples %d–%d\n\n",
                  (i - 1) * max_per_page + 1, (i - 1) * max_per_page + length(g)))
      page_dt <- df[get(sample_var) %in% g]
      page_dt[, (sample_var) := droplevels(get(sample_var))]
      print(plot_fn(page_dt))
      cat("\n\n")
    }
    cat(":::\n\n")
  }
}

mapping_metric_cols <- c("reads_m", "percent_mapped", "saturation_pct",
                          "barcodes_k", "median_genes_per_bc", "runtime_min")
mapping_metric_labs <- c(
  reads_m             = "reads (M)",
  percent_mapped      = "mapped (%)",
  saturation_pct      = "saturation (%)",
  barcodes_k          = "barcodes ≥200 reads (k)",
  median_genes_per_bc = "median genes/bc",
  runtime_min         = "runtime (min)"
)

knee_row1_cols <- c("n_total", "n_cell", "n_empty", "n_excluded", "expected_cells")
knee_row1_labs <- c(n_total="total barcodes", n_cell="cell", n_empty="ambient",
                     n_excluded="excluded", expected_cells="expected_cells")
knee_row2_cols <- c("total_droplets_included", "knee1", "shin1", "knee2", "shin2")
knee_row2_labs <- c(total_droplets_included="total_droplets_included", knee1="knee1",
                     shin1="shin1", knee2="knee2", shin2="shin2")

plot_knee_summary_page <- function(df) {
  plot_metric_bars(df, "short_id", knee_row1_cols, knee_row1_labs) /
    plot_metric_bars(df, "short_id", knee_row2_cols, knee_row2_labs)
}
```

## Sequencing & mapping

Read-level and per-barcode stats straight from `piscem`/`alevin-fry`
(`af_map/map_info.json`, `af_quant/featureDump.txt`), independent of cell
calling -- these answer "did this sample sequence and map cleanly", not
"which barcodes are cells" (see below for that).
`median_genes_per_bc` is restricted to barcodes with >=200 deduplicated reads
so it isn't swamped by the empty-droplet background.

```{r af_metrics_panel, fig.width=8, fig.height=40*0.16+0.6, results='asis'}
paginate_and_plot(af_sample_stats, "short_id", 40, function(d)
  plot_metric_bars(d, "short_id", mapping_metric_cols, mapping_metric_labs))
```

## Cells called per sample

Top row: population counts (cell = top `expected_cells` by rank; ambient =
`in_empty_plateau`; excluded = everything else). Bottom row: the underlying
DropletUtils barcode-rank parameters those counts are derived from.

```{r counts, fig.width=9, fig.height=20*0.16*2+1.4, results='asis'}
paginate_and_plot(summ, "short_id", 20, plot_knee_summary_page)
```

## Per-sample: UMI × splice by call

```{r per-sample, results='asis', fig.width=8, fig.height=3.6}
render_sample_chunks(sids, function(sid) {
  cat("#### ", sid, "\n\n")
  d <- fread(lab_fs[[sid]])
  d[, lx := log10(total+ps)]; d[, ly := qlogis((spliced+1)/(total+2))]
  d[, population := factor(population, levels=names(pop_cols))]
  p <- ggplot(d[total>0][order(population)], aes(lx, ly, colour=population)) +
    geom_point(size=0.1) +
    scale_colour_manual(values=pop_cols, drop=FALSE) +
    scale_x_continuous(breaks=LOGB, labels=LOGL) + scale_y_continuous(breaks=PCTB, labels=PCTL) +
    guides(colour=guide_legend(override.aes=list(size=2,alpha=1))) +
    theme_classic(base_size=8) +
    labs(title=paste(sid, "— knee/shin call"), x="total UMIs (log)", y="spliced fraction")
  print(p)
})
```

## Session info

<details>
```{r session_info}
sessionInfo()
```
</details>
