---
title: "scQC-flow"
subtitle: "Pipeline run landing page"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(jsonlite)
  library(knitr)
})

payload <- fromJSON("landing_page_payload.json", simplifyVector = FALSE)
report_files <- sort(list.files(".", pattern = "\\.html$", full.names = FALSE))
report_files <- setdiff(report_files, "index.html")

stage_order <- c("mapping", "ambient", "qc", "hvg", "integration", "annotation", "zoom")
stage_titles <- c(
  mapping = "Mapping",
  ambient = "Ambient",
  qc = "QC",
  hvg = "HVG",
  integration = "Integration",
  annotation = "Annotation",
  zoom = "Zoom"
)

known_reports <- list(
  mapping_report.html = list(title = "Mapping diagnostics", category = "Core", stage = "mapping", blurb = "Barcode knee plots and read-level mapping diagnostics across samples."),
  ambient_report.html = list(title = "Ambient RNA diagnostics", category = "Core", stage = "ambient", blurb = "decontX summaries, contamination estimates, and ambient cleanup diagnostics."),
  cellbender_report.html = list(title = "CellBender diagnostics", category = "Core", stage = "ambient", blurb = "Estimated-vs-called barcode comparisons and CellBender call diagnostics across samples."),
  qc_report.html = list(title = "Cell-level QC", category = "Core", stage = "qc", blurb = "Filtering metrics, threshold summaries, and QC visualizations."),
  hvg_report.html = list(title = "HVG selection", category = "Core", stage = "hvg", blurb = "Highly variable gene selection diagnostics and ambient-gene context."),
  integration_report.html = list(title = "Integration diagnostics", category = "Core", stage = "integration", blurb = "Harmony integration, cluster structure, metadata overlays, and QC-feature UMAPs."),
  annotation_report.html = list(title = "Annotation diagnostics", category = "Core", stage = "annotation", blurb = "Marker-panel matching, cluster labels, and marker expression views."),
  index.html = list(title = "Landing page", category = "Meta", stage = "meta", blurb = "Run overview and links to all generated HTML reports.")
)

pretty_report_name <- function(path) {
  if (!is.null(known_reports[[path]])) {
    return(known_reports[[path]]$title)
  }
  if (startsWith(path, "zoom_") && endsWith(path, "_report.html")) {
    zoom_name <- sub("^zoom_(.+)_report\\.html$", "\\1", path)
    zoom_name <- gsub("_", " ", zoom_name)
    zoom_name <- tools::toTitleCase(zoom_name)
    return(sprintf("Zoom: %s", zoom_name))
  }
  label <- sub("\\.html$", "", path)
  label <- gsub("_", " ", label)
  tools::toTitleCase(label)
}

report_category <- function(path) {
  if (!is.null(known_reports[[path]])) {
    return(known_reports[[path]]$category)
  }
  if (startsWith(path, "zoom_")) {
    return("Zoom")
  }
  "Other"
}

report_stage <- function(path) {
  if (!is.null(known_reports[[path]])) {
    return(known_reports[[path]]$stage)
  }
  if (startsWith(path, "zoom_")) {
    return("zoom")
  }
  "other"
}

report_blurb <- function(path) {
  if (!is.null(known_reports[[path]])) {
    return(known_reports[[path]]$blurb)
  }
  if (startsWith(path, "zoom_")) {
    return("Subset rerun report for a configured zoom, including reintegration and top-marker summaries.")
  }
  "Generated HTML report from this pipeline run."
}

flatten_params <- function(x, section = NULL) {
  if (is.null(x)) {
    return(data.frame(section = character(), parameter = character(), value = character()))
  }

  if (!is.list(x)) {
    return(data.frame(
      section = if (is.null(section)) "general" else section,
      parameter = if (is.null(section)) "value" else sub("^.+\\.", "", section),
      value = paste(as.character(x), collapse = ", "),
      stringsAsFactors = FALSE
    ))
  }

  if (length(x) == 0) {
    return(data.frame(section = character(), parameter = character(), value = character()))
  }

  rows <- lapply(seq_along(x), function(i) {
    value <- x[[i]]
    key <- names(x)[i]
    if (is.null(key) || key == "") {
      key <- paste0("item_", i)
    }
    next_section <- if (is.null(section)) key else paste(section, key, sep = ".")

    if (is.list(value) && length(value) > 0 && !is.null(names(value))) {
      flatten_params(value, next_section)
    } else if (is.list(value)) {
      data.frame(
        section = if (is.null(section)) key else section,
        parameter = key,
        value = toJSON(value, auto_unbox = TRUE),
        stringsAsFactors = FALSE
      )
    } else {
      data.frame(
        section = if (is.null(section)) "general" else section,
        parameter = key,
        value = paste(as.character(value), collapse = ", "),
        stringsAsFactors = FALSE
      )
    }
  })

  do.call(rbind, rows)
}

step_labels <- c(
  mapping = "Mapping",
  ambient = "Ambient",
  qc = "QC",
  hvg = "HVG",
  integration = "Integration",
  annotation = "Annotation",
  zooms = "Zooms"
)

enabled_steps <- payload$pipeline_steps
step_badges <- vapply(names(step_labels), function(step_name) {
  is_enabled <- isTRUE(enabled_steps[[step_name]])
  sprintf(
    '<span class="stage-badge %s">%s</span>',
    if (is_enabled) "on" else "off",
    step_labels[[step_name]]
  )
}, character(1))

zoom_items <- payload$params$zoom$items
zoom_count <- if (is.null(zoom_items)) 0 else length(zoom_items)
params_df <- flatten_params(payload$params)
params_df$section <- gsub("\\.", " / ", params_df$section)
params_df <- params_df[order(params_df$section, params_df$parameter), ]

report_df <- data.frame(
  file = report_files,
  title = vapply(report_files, pretty_report_name, character(1)),
  category = vapply(report_files, report_category, character(1)),
  stage = vapply(report_files, report_stage, character(1)),
  blurb = vapply(report_files, report_blurb, character(1)),
  stringsAsFactors = FALSE
)
report_df$stage_rank <- match(report_df$stage, stage_order)
report_df$stage_rank[is.na(report_df$stage_rank)] <- 999
report_df <- report_df[order(report_df$stage_rank, report_df$title), ]

format_runtime <- function(seconds) {
  if (is.null(seconds) || length(seconds) == 0 || is.na(seconds)) {
    return("Not available")
  }

  seconds <- as.integer(seconds)
  hours <- seconds %/% 3600
  minutes <- (seconds %% 3600) %/% 60
  secs <- seconds %% 60

  parts <- character(0)
  if (hours > 0) {
    parts <- c(parts, sprintf("%dh", hours))
  }
  if (minutes > 0 || hours > 0) {
    parts <- c(parts, sprintf("%dm", minutes))
  }
  parts <- c(parts, sprintf("%ds", secs))
  paste(parts, collapse = " ")
}

get_retained_cells <- function() {
  integration_files <- Sys.glob("integration_dt*.csv.gz")
  integration_files <- integration_files[basename(integration_files) != "NO_FILE"]
  if (length(integration_files) == 0) {
    return(NA_integer_)
  }

  tryCatch({
    integration_dt <- fread(integration_files[[1]], select = "UMAP1", showProgress = FALSE)
    sum(!is.na(integration_dt$UMAP1))
  }, error = function(...) {
    NA_integer_
  })
}

ambient_method <- payload$params$workflow_flags$ambient_method
ambient_method <- if (is.null(ambient_method) || !nzchar(ambient_method)) "Not available" else ambient_method
runtime_text <- format_runtime(payload$runtime_seconds)
retained_cells <- get_retained_cells()
retained_cells_text <- if (is.na(retained_cells)) "Not available" else format(retained_cells, big.mark = ",")

has_report <- function(file_name) {
  file_name %in% report_df$file
}

stage_report_file <- function(stage_name) {
  stage_files <- report_df$file[report_df$stage == stage_name]
  if (length(stage_files) == 0) {
    return(NA_character_)
  }
  stage_files[[1]]
}

render_stage_node <- function(label, file_name = NA_character_, status = "active", extra_class = "") {
  classes <- paste(c("dag-node", status, extra_class), collapse = " ")
  if (!is.na(file_name) && nzchar(file_name)) {
    sprintf('<a class="%s" href="%s">%s</a>', classes, file_name, label)
  } else {
    sprintf('<span class="%s">%s</span>', classes, label)
  }
}

main_flow_nodes <- c("mapping", "ambient", "qc", "hvg", "integration", "annotation")
main_flow_html <- vapply(main_flow_nodes, function(stage_name) {
  status <- if (isTRUE(enabled_steps[[stage_name]])) "active" else "muted"
  render_stage_node(stage_titles[[stage_name]], stage_report_file(stage_name), status)
}, character(1))

branch_source_stage <- if (isTRUE(enabled_steps[["annotation"]])) "annotation" else "integration"
branch_source_index <- match(branch_source_stage, main_flow_nodes)
branch_prefix_html <- character(0)
if (!is.na(branch_source_index) && branch_source_index > 1) {
  for (stage_name in main_flow_nodes[seq_len(branch_source_index - 1)]) {
    branch_prefix_html <- c(
      branch_prefix_html,
      render_stage_node(stage_titles[[stage_name]], NA_character_, "muted", "ghost"),
      '<span class="dag-arrow ghost">&rarr;</span>'
    )
  }
}

zoom_df <- report_df[report_df$stage == "zoom", , drop = FALSE]
zoom_nodes_html <- if (nrow(zoom_df) > 0) {
  vapply(seq_len(nrow(zoom_df)), function(i) {
    render_stage_node(zoom_df$title[[i]], zoom_df$file[[i]], "active", "zoom")
  }, character(1))
} else {
  character(0)
}

report_sections <- split(report_df, factor(report_df$stage, levels = stage_order))
report_sections <- report_sections[vapply(report_sections, nrow, integer(1)) > 0]
```

```{=html}
<style>
#title-block-header {
  display: none;
}
.hero {
  padding: 1.5rem 1.75rem;
  border-radius: 18px;
  background: linear-gradient(135deg, #f7f2e8 0%, #e2eef3 100%);
  border: 1px solid #d7dde3;
  margin-bottom: 1.5rem;
}
.hero h1 {
  margin: 0 0 0.4rem 0;
  font-size: 2rem;
}
.hero p {
  margin: 0;
  max-width: 60rem;
}
.meta-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(180px, 1fr));
  gap: 0.9rem;
  margin: 1.25rem 0;
}
.meta-card {
  background: rgba(255, 255, 255, 0.78);
  border: 1px solid #d8e0e6;
  border-radius: 14px;
  padding: 0.9rem 1rem;
}
.meta-label {
  font-size: 0.75rem;
  text-transform: uppercase;
  letter-spacing: 0.06em;
  color: #5a6973;
}
.meta-value {
  font-size: 1.1rem;
  font-weight: 600;
  margin-top: 0.2rem;
  color: #1f2b33;
}
.badge-row {
  display: flex;
  flex-wrap: wrap;
  gap: 0.55rem;
  margin-top: 0.5rem;
}
.stage-badge {
  display: inline-flex;
  align-items: center;
  border-radius: 999px;
  padding: 0.32rem 0.7rem;
  font-size: 0.86rem;
  font-weight: 600;
  border: 1px solid transparent;
}
.stage-badge.on {
  background: #e7f4ea;
  border-color: #a8c8b0;
  color: #1e5c32;
}
.stage-badge.off {
  background: #f2f4f5;
  border-color: #d8dee2;
  color: #667780;
}
.dag-wrap {
  margin: 1rem 0 1.75rem 0;
  padding: 1.1rem 1.15rem;
  border-radius: 18px;
  background: linear-gradient(180deg, #fbfcfc 0%, #f3f6f7 100%);
  border: 1px solid #d8dee2;
}
.dag-row {
  display: flex;
  flex-wrap: wrap;
  align-items: center;
  gap: 0.7rem;
  margin: 0.65rem 0;
}
.dag-row.branch {
  padding-left: 0;
}
.dag-arrow {
  color: #6c7c86;
  font-weight: 700;
  font-size: 1rem;
}
.dag-node {
  display: inline-flex;
  align-items: center;
  justify-content: center;
  min-height: 3rem;
  padding: 0.7rem 1rem;
  border-radius: 14px;
  border: 1px solid #cfd8dd;
  background: #ffffff;
  color: #1f2b33;
  text-decoration: none;
  font-weight: 600;
  box-shadow: 0 6px 16px rgba(22, 31, 38, 0.05);
}
.dag-node:hover {
  text-decoration: none;
  transform: translateY(-1px);
}
.dag-node.active {
  border-color: #a9c3cd;
  background: linear-gradient(135deg, #fffef8 0%, #edf6f8 100%);
}
.dag-node.muted {
  background: #f3f5f6;
  color: #70808a;
  border-style: dashed;
  box-shadow: none;
}
.dag-node.zoom {
  background: linear-gradient(135deg, #fff5ea 0%, #fffdf8 100%);
  border-color: #e0c6a8;
}
.dag-node.ghost,
.dag-arrow.ghost {
  visibility: hidden;
}
.section-label {
  margin: 1.35rem 0 0.4rem 0;
  font-size: 1rem;
  color: #30414a;
}
</style>
```

```{r hero, results='asis'}
description_steps <- names(Filter(isTRUE, enabled_steps))
description_steps <- step_labels[description_steps]
description_text <- if (length(description_steps) > 0) {
  paste(description_steps, collapse = ", ")
} else {
  "No optional downstream stages"
}

cat(sprintf(
  '<div class="hero"><h1>Run overview</h1><p>Summary of this scQC-flow run, including the effective configuration and direct links to each generated report. Enabled stages: %s.</p><div class="meta-grid"><div class="meta-card"><div class="meta-label">Run name</div><div class="meta-value">%s</div></div><div class="meta-card"><div class="meta-label">Profile</div><div class="meta-value">%s</div></div><div class="meta-card"><div class="meta-label">Ambient method</div><div class="meta-value">%s</div></div><div class="meta-card"><div class="meta-label">Cells retained</div><div class="meta-value">%s</div></div><div class="meta-card"><div class="meta-label">Run time</div><div class="meta-value">%s</div></div></div><div class="badge-row">%s</div></div>',
  description_text,
  payload$run_name,
  payload$profile,
  ambient_method,
  retained_cells_text,
  runtime_text,
  paste(step_badges, collapse = "")
))
```

## Run overview

* **Generated at:** `r payload$generated_at`
* **Run time:** `r runtime_text`
* **Total cells retained:** `r retained_cells_text`
* **Ambient method:** `r ambient_method`
* **Output directory:** `r payload$output_dir`
* **Reports directory:** `r payload$reports_dir`
* **Raw data directory:** `r payload$params$input_output$raw_data_dir`
* **Chemistry:** `r payload$params$references$chemistry`

## Pipeline flow

```{r pipeline_flow, results='asis'}
cat('<div class="dag-wrap">')
cat('<div class="dag-row">')
cat(paste(main_flow_html, collapse = '<span class="dag-arrow">&rarr;</span>'))
cat('</div>')
if (length(zoom_nodes_html) > 0) {
  cat('<div class="dag-row branch">')
  if (length(branch_prefix_html) > 0) {
    cat(paste(branch_prefix_html, collapse = ''))
  }
  cat('<span class="dag-arrow">&boxur;</span>')
  cat(render_stage_node('Zoom subsets', NA_character_, 'active', 'zoom'))
  cat('<span class="dag-arrow">&rarr;</span>')
  cat(paste(zoom_nodes_html, collapse = '<span class="dag-arrow">&rarr;</span>'))
  cat('</div>')
}
cat('</div>')
```

## Report manifest

```{r report_table}
report_manifest_df <- data.frame(
  stage = report_df$stage,
  report = sprintf('<a href="%s">%s</a>', report_df$file, report_df$title),
  file = sprintf('<a href="%s">%s</a>', report_df$file, report_df$file),
  stringsAsFactors = FALSE
)

knitr::kable(
  report_manifest_df,
  col.names = c('Stage', 'Report', 'File'),
  caption = 'Generated HTML reports linked from this landing page',
  escape = FALSE
)
```

## Parameters

```{r params_table}
knitr::kable(
  params_df,
  col.names = c('Section', 'Parameter', 'Value'),
  caption = 'Effective run configuration captured for this pipeline execution'
)
```