---
title: "Read mapping diagnostics — EmptyDrops cell calling"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: false
execute:
  warning: false
  message: false
  cache: false
---

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(ggrepel)
  library(ggbeeswarm)
  library(ggh4x)
  library(magrittr)
  library(BiocParallel)
  library(patchwork)
})
```

```{r load_data}
ed_fs <- list.files(".", pattern = "^ed_plot_data_.+\\.csv$",
                    full.names = TRUE, recursive = FALSE)

if (length(ed_fs) == 0) stop("No ed_plot_data_*.csv files found")

sample_var <- "sample_id"

ids   <- sub("^ed_plot_data_(.+)\\.csv$", "\\1", basename(ed_fs))
ed_fs <- setNames(ed_fs, ids)
```

```{r helpers}
# ---- per-sample summary -------------------------------------------------------
get_ed_summary <- function(ed_f, sample_var) {
  dt <- fread(ed_f)
  sid <- unique(dt[[sample_var]])[1]

  knee1_val      <- unique(dt$knee1)[1]
  total_barcodes <- nrow(dt)
  n_auto_cell    <- sum(!is.na(dt$ED_Total) & dt$ED_Total >= knee1_val,   na.rm = TRUE)
  n_tested       <- sum(!is.na(dt$ED_PValue), na.rm = TRUE)
  n_sig          <- sum(!is.na(dt$ED_FDR) & dt$ED_FDR <= 0.001,          na.rm = TRUE)
  n_total_cells  <- sum(dt$is_cell == TRUE,                               na.rm = TRUE)

  data.table(
    sample_id           = sid,
    total_barcodes      = total_barcodes,
    knee1_retain_thr    = as.integer(knee1_val),
    cells_auto_retained = n_auto_cell,
    barcodes_tested     = n_tested,
    cells_by_test       = n_sig,
    total_cells_ED      = n_total_cells,
    pct_cells           = round(100 * n_total_cells / total_barcodes, 2)
  )
}

# ---- barcode rank plot with cell/empty/plateau coloring ----------------------
plot_barcode_ranks_ED <- function(ed_fs, sample_var) {
  s_ord <- names(ed_fs)

  plot_dt <- lapply(s_ord, function(s) {
    dt <- fread(ed_fs[[s]]) %>%
      .[, .(barcode, rank, total, knee1, shin1, knee2, shin2,
            total_droplets_included, in_empty_plateau, is_cell,
            sample_id = get(sample_var))]
    dt[, status := fcase(
      is_cell == TRUE,           "cell",
      in_empty_plateau == TRUE,  "empty plateau",
      default                    = "other"
    )]
    dt
  }) %>% rbindlist()

  # aggregate to rank curve (one point per unique lib size per sample)
  rank_curve <- plot_dt[, .(n_bc = .N), by = .(sample_id, lib_size = total)] %>%
    .[order(sample_id, -lib_size)] %>%
    .[, bc_rank := cumsum(n_bc), by = sample_id]

  # knee lines
  knee_lines <- lapply(s_ord, function(s) {
    dt <- fread(ed_fs[[s]])
    data.table(
      sample_id = s,
      variable  = c("knee1", "shin1", "knee2", "shin2", "total_droplets_included"),
      value     = c(unique(dt$knee1), unique(dt$shin1), unique(dt$knee2),
                    unique(dt$shin2), unique(dt$total_droplets_included)),
      axis      = c("y", "y", "y", "y", "x")
    )
  }) %>% rbindlist()

  hlines <- knee_lines[axis == "y"]
  vlines <- knee_lines[axis == "x"]

  p_labels <- c("1", "10", "100", "1k", "10k", "100k", "1M")
  p_breaks <- c(1e0, 1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

  rank_curve[[sample_var]] <- factor(rank_curve[[sample_var]], levels = s_ord)
  hlines[[sample_var]]     <- factor(hlines[[sample_var]], levels = s_ord)
  vlines[[sample_var]]     <- factor(vlines[[sample_var]], levels = s_ord)

  ggplot(rank_curve) +
    aes(x = bc_rank, y = lib_size) +
    geom_line(linewidth = 0.3, color = "#283747") +
    geom_hline(data = hlines,
               aes(yintercept = value, color = variable), linewidth = 0.4) +
    geom_vline(data = vlines,
               aes(xintercept = value), color = "#88a0dc", linewidth = 0.4,
               linetype = "dashed") +
    geom_text_repel(data = hlines, aes(y = value, x = 10, label = variable,
                                       color = variable), size = 2.2,
                    direction = "y", nudge_x = -0.2) +
    facet_wrap(~ get(sample_var), ncol = 4) +
    scale_x_log10(labels = p_labels, breaks = p_breaks) +
    scale_y_log10(labels = p_labels, breaks = p_breaks) +
    scale_color_manual(values = c(
      knee1 = "#c0392b", shin1 = "#e67e22",
      knee2 = "#2980b9", shin2 = "#1abc9c"
    )) +
    theme_classic(base_size = 9) +
    theme(legend.position = "none") +
    labs(x = "barcode rank", y = "library size")
}

# ---- p-value histogram --------------------------------------------------------
plot_pvalue_hist <- function(ed_fs, sample_var, ncol = 4) {
  plot_dt <- lapply(names(ed_fs), function(s) {
    fread(ed_fs[[s]])[!is.na(ED_PValue),
                      .(ED_PValue, sample_id = get(sample_var))]
  }) %>% rbindlist()

  plot_dt[[sample_var]] <- factor(plot_dt[[sample_var]], levels = names(ed_fs))

  ggplot(plot_dt, aes(x = ED_PValue)) +
    geom_histogram(binwidth = 0.05, boundary = 0,
                   fill = "#4a90d9", color = "white", linewidth = 0.2) +
    facet_wrap(~ get(sample_var), ncol = ncol, scales = "free_y") +
    theme_classic(base_size = 9) +
    labs(x = "p-value (EmptyDrops)", y = "count",
         title = "P-value distribution for tested barcodes",
         subtitle = "Uniform distribution = good ambient profile; peak at 0 = true cells detected")
}

# ---- LogProb vs total UMI scatter --------------------------------------------
plot_logprob_scatter <- function(ed_f, sample_var) {
  dt <- fread(ed_f) %>%
    .[!is.na(ED_LogProb)] %>%
    .[, status := fifelse(is_cell == TRUE, "cell", "empty/untested")]

  sid <- unique(dt[[sample_var]])[1]

  p_labels <- c("1", "10", "100", "1k", "10k", "100k", "1M")
  p_breaks <- c(1e0, 1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

  ggplot(dt[order(status)], aes(x = total, y = ED_LogProb, color = status)) +
    geom_point(size = 0.3, alpha = 0.5) +
    scale_x_log10(labels = p_labels, breaks = p_breaks) +
    scale_color_manual(values = c(cell = "#1965B0", "empty/untested" = "grey60")) +
    theme_classic(base_size = 9) +
    labs(x = "total UMIs", y = "log-probability (EmptyDrops)",
         color = NULL, title = sid) +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
}

# ---- FDR vs total UMI scatter -------------------------------------------------
plot_fdr_scatter <- function(ed_f, sample_var) {
  dt <- fread(ed_f) %>%
    .[!is.na(ED_FDR)] %>%
    .[, is_sig := ED_FDR <= 0.001]

  sid <- unique(dt[[sample_var]])[1]

  p_labels <- c("1", "10", "100", "1k", "10k", "100k", "1M")
  p_breaks <- c(1e0, 1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

  ggplot(dt[order(is_sig)], aes(x = total, y = ED_FDR, color = is_sig)) +
    geom_point(size = 0.3, alpha = 0.5) +
    geom_hline(yintercept = 0.001, linetype = "dashed", color = "#c0392b",
               linewidth = 0.5) +
    scale_x_log10(labels = p_labels, breaks = p_breaks) +
    scale_y_log10() +
    scale_color_manual(values = c("TRUE" = "#1965B0", "FALSE" = "grey60"),
                       labels = c("TRUE" = "cell (FDR ≤ 0.001)", "FALSE" = "not called")) +
    theme_classic(base_size = 9) +
    labs(x = "total UMIs", y = "FDR", color = NULL, title = sid) +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
}

# ---- UMI / splice violin -------------------------------------------------------
plot_qc_metrics_ED <- function(ed_fs, sample_var = "sample_id") {
  plot_dt <- lapply(ed_fs, function(f) {
    fread(f) %>%
      .[is_cell == TRUE | in_empty_plateau == TRUE] %>%
      .[, `:=`(
        umis       = log10(total),
        splice_pct = qlogis((spliced + 1) / (spliced + unspliced + 2)),
        what       = fifelse(is_cell == TRUE, "cell", "empty")
      )] %>%
      .[, c(sample_var, "barcode", "umis", "splice_pct", "what"), with = FALSE]
  }) %>%
    rbindlist() %>%
    melt(id.vars       = c(sample_var, "barcode", "what"),
         measure.vars  = c("umis", "splice_pct"),
         variable.name = "qc_metric", value.name = "qc_val") %>%
    .[, qc_metric := fcase(
      qc_metric == "umis",       "no. of UMIs",
      qc_metric == "splice_pct", "spliced pct.",
      default = NA_character_
    )]

  umis_brks   <- c(1e0,1e1,3e1,1e2,3e2,1e3,3e3,1e4,3e4,1e5,3e5,1e6) %>% log10()
  umis_labs   <- c("1","10","30","100","300","1k","3k","10k","30k","100k","300k","1M")
  splice_brks <- c(0.01,0.03,0.1,0.3,0.5,0.7,0.9,0.97,0.99) %>% qlogis()
  splice_labs <- c("1%","3%","10%","30%","50%","70%","90%","97%","99%")

  ggplot() +
    geom_violin(
      data   = plot_dt[!is.na(qc_val)],
      aes(x = get(sample_var), y = qc_val, fill = what),
      colour = NA, kernel = "rectangular", adjust = 0.1, scale = "width", width = 0.8
    ) +
    facet_grid(. ~ qc_metric, scales = "free", space = "free_y") +
    facetted_pos_scales(
      y = list(
        qc_metric == "no. of UMIs"  ~ scale_y_continuous(breaks = umis_brks, labels = umis_labs),
        qc_metric == "spliced pct." ~ scale_y_continuous(breaks = splice_brks, labels = splice_labs)
      )
    ) +
    scale_fill_manual(values = c(cell = "#1965B0", empty = "grey")) +
    coord_flip() +
    theme_classic() +
    labs(x = NULL, y = NULL, fill = "barcode\ncontent") +
    theme(panel.spacing = unit(1, "lines"))
}

# ---- layout helpers ------------------------------------------------------------
plot_dims <- function(sample_n) {
  w <- 9; h <- 6
  if (sample_n <= 4)      { h <- ceiling(6 * 1/3); w <- ceiling(9 * sample_n / 4) }
  else if (sample_n <= 8) { h <- ceiling(6 * 2/3) }
  violin_h   <- 0.6
  max_marg_h <- 14
  list(w = w, h = h, marg_h = min((sample_n + 1) * violin_h, max_marg_h))
}
```

```{r compute}
ed_summary <- lapply(names(ed_fs), function(s)
  get_ed_summary(ed_fs[[s]], sample_var)) %>%
  rbindlist() %>% .[order(sample_id)]

dims <- plot_dims(length(ed_fs))
```

## Called cells summary

```{r summary_table}
knitr::kable(ed_summary, digits = 2,
  caption = paste0(
    "EmptyDrops cell calls per sample. ",
    "cells_auto_retained: Total >= knee1 (retain threshold, FDR = NA); ",
    "cells_by_test: FDR ≤ 0.001 among tested barcodes; ",
    "total_cells_ED: union of both."
  )
)
```

## Barcode rank plots

Knee parameters shown as horizontal lines (knee1 = dark red, shin1 = orange,
knee2 = blue, shin2 = teal). Dashed vertical line = `total_droplets_included`.

```{r rank_plots, fig.height=dims$h, fig.width=dims$w}
suppressWarnings(print(
  plot_barcode_ranks_ED(ed_fs, sample_var)
))
```

## EmptyDrops diagnostics {.tabset}

### P-value histograms

For barcodes between `lower` (default 100 UMIs) and `retain` (`knee1`) that were
tested against the ambient profile.  A uniform distribution indicates a well-estimated
ambient profile; enrichment near 0 reflects true cells.

```{r pval_hist, fig.height=dims$h, fig.width=dims$w}
plot_pvalue_hist(ed_fs, sample_var, ncol = 4)
```

### LogProb vs UMI {.tabset}

Log-probability assigned by EmptyDrops vs total UMI count. Cells (blue) should cluster
at lower (more negative) log-probabilities relative to empty droplets of similar size.

```{r logprob_scatter, fig.height=dims$h, fig.width=dims$w, results='asis'}
for (s in names(ed_fs)) {
  cat("####", s, "\n")
  print(plot_logprob_scatter(ed_fs[[s]], sample_var))
  cat("\n\n")
}
```

### FDR vs UMI {.tabset}

FDR for tested barcodes (barcodes above `knee1` are auto-retained and not shown here).
Dashed red line = 0.001 threshold.

```{r fdr_scatter, fig.height=dims$h, fig.width=dims$w, results='asis'}
for (s in names(ed_fs)) {
  cat("####", s, "\n")
  print(plot_fdr_scatter(ed_fs[[s]], sample_var))
  cat("\n\n")
}
```

## Library sizes and spliced proportions

Cells (EmptyDrops calls) vs empty plateau barcodes.

```{r violin, fig.height=dims$marg_h, fig.width=9}
plot_qc_metrics_ED(ed_fs, sample_var)
```

## Session info

<details>
```{r session_info}
devtools::session_info()
```
</details>
