---
title: "Per-sample annotation (workflow_improvement.md step 4)"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{=html}
<style>
.quarto-figure img, .cell-output-display img, p > img, img.img-fluid {
  max-width: min(100%, 820px);
  height: auto;
}
</style>
```

Per-sample normalize -> HVG -> PCA -> Leiden(`r Sys.getenv("ANNOTATION_CLUSTER_COL")`) ->
xgboost cluster-level annotation, run independently per sample (no cross-sample
integration/Harmony). Reuses the exact pipeline scripts (`hvg_selection.py`,
`run_integration.py --singlet_only`, `prepare_annotation_query.R`,
`run_xgboost_reference_annotation.R`) and report styling (`plot_umap_cluster`, `nice_cols`,
`plot_qc_metric_scatter`, `plot_qc_split_violin`) as the rest of the pipeline.

```{r setup}
knitr::opts_chunk$set(fig.path = 'plots/', dev = 'png', dpi = 110)
suppressPackageStartupMessages({library(data.table); library(ggplot2)})
source("integration_plots.R")
source("qc_plots.R")
setDTthreads(1)

method_id <- Sys.getenv("ANNOTATION_METHOD_ID")

cells_fs <- list.files(".", pattern = sprintf("^annotation_cells_%s_.+\\.csv\\.gz$", method_id), full.names = TRUE)
names(cells_fs) <- sub(sprintf("^annotation_cells_%s_(.+)\\.csv\\.gz$", method_id), "\\1", basename(cells_fs))
summ_fs <- list.files(".", pattern = sprintf("^annotation_cluster_summary_%s_.+\\.csv\\.gz$", method_id), full.names = TRUE)
names(summ_fs) <- sub(sprintf("^annotation_cluster_summary_%s_(.+)\\.csv\\.gz$", method_id), "\\1", basename(summ_fs))
qc_fs <- list.files(".", pattern = "^qc_metrics_.+\\.csv\\.gz$", full.names = TRUE)
names(qc_fs) <- sub("\\.csv\\.gz$", "", sub("^qc_metrics_", "", basename(qc_fs)))

sids <- sort(Reduce(intersect, list(names(cells_fs), names(summ_fs), names(qc_fs))))

method_label <- toupper(substr(method_id, 1, 1)) %>% paste0(substr(method_id, 2, nchar(method_id)))

qc_names <- c("n_cells", "log_counts", "log_feats", "logit_mito", "logit_spliced")
qc_lu    <- setNames(c("no. of cells", "no. of UMIs", "no. of genes", "mito. pct.", "spliced pct."), qc_names)
hard_min_counts <- as.numeric(Sys.getenv("HARD_MIN_COUNTS", "200"))
hard_min_feats  <- as.numeric(Sys.getenv("HARD_MIN_FEATS", "100"))
hard_max_mito   <- as.numeric(Sys.getenv("HARD_MAX_MITO", "0.5"))
min_counts <- as.numeric(Sys.getenv("MIN_COUNTS", "500"))
min_feats  <- as.numeric(Sys.getenv("MIN_FEATS", "300"))
max_mito   <- as.numeric(Sys.getenv("MAX_MITO", "0.2"))
min_mito   <- as.numeric(Sys.getenv("MIN_MITO", "0"))
max_splice <- as.numeric(Sys.getenv("MAX_SPLICE", "0.6"))
min_splice <- as.numeric(Sys.getenv("MIN_SPLICE", "0"))

cuts_dt      <- make_cuts_dt(sids, min_counts, min_feats, min_mito, max_mito, min_splice, max_splice)
hard_cuts_dt <- make_hard_cuts_dt(sids, hard_min_counts, hard_min_feats, hard_max_mito)

# Per-sample data, loaded once up front and reused across all sections
load_sample <- function(sid) {
  qc_all_dt <- fread(qc_fs[[sid]])
  setnames(qc_all_dt, "sample_id", "batch_var")
  list(
    int_dt    = fread(cells_fs[[sid]]),
    m_summ    = fread(summ_fs[[sid]]),
    qc_all_dt = qc_all_dt,
    kept_dt   = qc_all_dt[keep == TRUE]
  )
}
sample_data <- setNames(lapply(sids, load_sample), sids)
```

```{r pca_plot_fn}
# Same styling as plot_umap_cluster, applied to PC1/PC2 instead of UMAP1/UMAP2
plot_pca_cluster <- function(dt) {
  pdt <- copy(dt)[, .(cell_id, UMAP1 = pca_01, UMAP2 = pca_02)]
  plot_umap_cluster(pdt, dt[, .(cell_id, cluster)], name = "cluster") +
    labs(x = "PC1", y = "PC2")
}
```

## QC metrics

Per-sample distributions (vertical violin, pre-QC vs post-QC per metric)
shown first so the effect of filtering on each metric is visible directly.
Dashed lines mark the soft thresholds, dotted lines the hard thresholds (see
legend). Same thresholds as the pipeline's `qc_report`.

```{r results='asis', fig.height=4, fig.width=9}
cat("::: {.panel-tabset}\n\n")
for (sid in sids) {
  cat("### ", sid, "\n\n")
  d <- sample_data[[sid]]
  suppressWarnings(print(
    plot_qc_split_violin(d$qc_all_dt, sid, qc_names, qc_lu, cuts_dt, hard_cuts_dt = hard_cuts_dt)
  ))
  cat("\n\n")
}
cat(":::\n\n")
```

Reasons cells were removed (UpSet plot of overlapping exclusion criteria:
doublet, hard threshold, soft threshold per metric, MAD-adaptive filter per
metric). Tabs are omitted for samples with nothing excluded.

```{r results='asis', fig.height=4, fig.width=7}
cat("::: {.panel-tabset}\n\n")
for (sid in sids) {
  d <- sample_data[[sid]]
  qc_tmp <- d$qc_all_dt
  if (nrow(qc_tmp) <= 1 || sum(qc_tmp$keep) == nrow(qc_tmp)) next
  cat("### ", sid, "\n\n")
  suppressMessages(print(plot_upset_of_exclusions(qc_tmp, qc_names, qc_lu, cuts_dt)))
  cat("\n\n")
}
cat(":::\n\n")
```

## QC density plots

Pairwise relationships between QC metrics at individual cell level (grey = soft
threshold pass zone).

```{r results='asis', fig.height=6, fig.width=8}
cat("::: {.panel-tabset}\n\n")
for (sid in sids) {
  cat("### ", sid, "\n\n")
  d <- sample_data[[sid]]
  print(plot_qc_metric_scatter(d$qc_all_dt, qc_names, qc_lu, cuts_dt[batch_var == sid], sid))
  cat("\n\n")
}
cat(":::\n\n")
```

## PCA

```{r results='asis'}
cat("::: {.panel-tabset}\n\n")
for (sid in sids) {
  cat("### ", sid, "\n\n")
  print(plot_pca_cluster(sample_data[[sid]]$int_dt))
  cat("\n\n")
}
cat(":::\n\n")
```

```{r results='asis'}
cat(sprintf("## Annotation (%s)\n\n", method_label))
```

Cell label (left) and cluster (right) over the same per-sample UMAP -- both shown
together since the cluster identity is already visible alongside the label.

```{r results='asis'}
cat("::: {.panel-tabset}\n\n")
for (sid in sids) {
  cat("### ", sid, "\n\n")
  d <- sample_data[[sid]]
  g_lab <- plot_umap_cluster(
    d$int_dt[, .(cell_id, UMAP1, UMAP2)],
    d$int_dt[, .(cell_id, cluster = label, centroid_group = as.character(label), centroid_label = as.character(label))],
    name = "cell label"
  ) + theme(legend.position = "none")
  g_clu <- plot_umap_cluster(
    d$int_dt[, .(cell_id, UMAP1, UMAP2)],
    d$int_dt[, .(cell_id, cluster)],
    name = "cluster"
  ) + theme(legend.position = "none")
  print(g_lab + g_clu)
  cat(sprintf("\n\n%d cells, %d clusters.\n\n", nrow(d$int_dt), uniqueN(d$int_dt$cluster)))
}
cat(":::\n\n")
```

## Cells annotated

Cluster-level annotation summary (top label + mean prediction score per cluster).

```{r results='asis'}
cat("::: {.panel-tabset}\n\n")
for (sid in sids) {
  cat("### ", sid, "\n\n")
  print(knitr::kable(sample_data[[sid]]$m_summ[order(cluster)],
                      caption = sprintf("%s cluster summary", method_label)))
  cat("\n\n")
}
cat(":::\n\n")
```

## Session info

<details>
```{r session_info}
sessionInfo()
```
</details>
