---
title: "Cell-level QC"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
source("qc_plots.R")
setDTthreads(1)

# QC metric definitions
qc_names <- c("n_cells", "log_counts", "log_feats", "logit_mito", "logit_spliced")
qc_lu    <- c("no. of cells", "no. of UMIs", "no. of genes", "mito. pct.", "spliced pct.") %>%
  setNames(qc_names)

# Hard thresholds (for pre-filtering obvious junk before plotting)
hard_min_counts <- as.numeric(Sys.getenv("HARD_MIN_COUNTS", "200"))
hard_min_feats  <- as.numeric(Sys.getenv("HARD_MIN_FEATS",  "100"))
hard_max_mito   <- as.numeric(Sys.getenv("HARD_MAX_MITO",   "0.5"))

# Soft thresholds (for QC ranges shown on plots)
min_counts  <- as.numeric(Sys.getenv("MIN_COUNTS",  "200"))
min_feats   <- as.numeric(Sys.getenv("MIN_FEATS",   "100"))
max_mito    <- as.numeric(Sys.getenv("MAX_MITO",    "0.1"))
min_mito    <- as.numeric(Sys.getenv("MIN_MITO",    "0.0001"))
max_splice  <- as.numeric(Sys.getenv("MAX_SPLICE",  "0.75"))
min_splice  <- as.numeric(Sys.getenv("MIN_SPLICE",  "0.0001"))
```

```{r load_data}
# Load all per-sample QC metrics CSVs (collected by Nextflow)
qc_files <- Sys.glob("qc_metrics_*.csv.gz")
stopifnot("No qc_metrics CSV files found" = length(qc_files) > 0)

qc_dt <- rbindlist(lapply(qc_files, fread))
setnames(qc_dt, "sample_id", "batch_var")

# All cells entering QC — used for pre-QC plots, scatter plots, and N pre-QC in summary
qc_all_dt <- copy(qc_dt)

# Hard-threshold-filtered subset — used internally for upset soft-threshold analysis
qc_dt <- qc_dt[log_counts >= log10(hard_min_counts) &
                log_feats  >= log10(hard_min_feats) &
                logit_mito <  qlogis(hard_max_mito)]

# Cells passing full QC (all thresholds, singlets only)
kept_dt <- qc_all_dt[keep == TRUE]

# Sample ordering from the full input
b_lvls    <- sort(unique(qc_all_dt$batch_var))
n_batches <- length(b_lvls)

# Build cuts data.table from threshold params
cuts_dt <- make_cuts_dt(b_lvls, min_counts, min_feats,
                        min_mito, max_mito, min_splice, max_splice)
hard_cuts_dt <- make_hard_cuts_dt(b_lvls, hard_min_counts, hard_min_feats, hard_max_mito)
```

```{r plot_sizes}
violin_h   <- 0.4
max_marg_h <- 12
marg_h     <- min((n_batches + 1) * violin_h, max_marg_h)
```

## Sample QC metrics {.tabset}

An overview of quality control (QC) metrics for all samples, both before and after filtering. The QC metrics include:

* **no. of cells** : Total number of barcodes called as cells for each sample.
* **no. of UMIs** : Sum of counts across all features for each cell (log-transformed).
* **no. of genes** : Number of unique features with non-zero counts for each cell (log-transformed).
* **mito. pct.** : The proportion of reads mapping to mitochondrial genes. High proportions indicate poor-quality cells or nuclei. Logit-transformed values are displayed.
* **spliced pct.** : The proportion of spliced reads. In single-nuclei RNA-seq, high spliced proportions may indicate inadequate removal of cytoplasmic material. Logit-transformed values are displayed.


### Pre-QC

```{r plot_pre_qc, fig.height=marg_h, fig.width=8}
suppressWarnings(
  plot_qc_ranges_marginals(qc_all_dt, b_lvls, qc_names, qc_lu, cuts_dt,
                           hard_cuts_dt = hard_cuts_dt)
)
```

### After QC

```{r plot_post_qc, fig.height=marg_h, fig.width=8}
suppressWarnings(
  plot_qc_ranges_marginals(kept_dt, b_lvls, qc_names, qc_lu, cuts_dt)
)
```

## QC metrics scatterplots

Pairwise relationships of QC metrics at individual cell level.

```{r plot_scatters, fig.height=6, fig.width=8, results='asis'}
cat('::: {.panel-tabset}\n\n')
for (b in b_lvls) {
  qc_tmp <- qc_dt[batch_var == b]
  if (nrow(qc_tmp) <= 1) next
  cat("### ", b, "\n\n")
  print(plot_qc_metric_scatter(qc_tmp, qc_names, qc_lu, cuts_dt[batch_var == b], b))
  cat("\n\n")
}
cat(':::' )
```

## How many cells and samples retained?

Table summarizing the number of cells before and after QC filtering. The per-metric
exclusion columns are counts of cells failing each criterion and may overlap,
so they are not expected to sum to the total excluded count.

```{r summary_table}
htmltools::tagList(
  htmltools::tags$style(htmltools::HTML("\n+    .qc-summary-table { font-size: 0.85rem; }\n+    .qc-summary-table table.dataTable thead th { font-size: 0.85rem; }\n+    .qc-summary-table .dataTables_wrapper input[type='search'],\n+    .qc-summary-table .dataTables_wrapper input[type='text'] {\n+      font-size: 0.8rem;\n+      padding: 2px 4px;\n+      height: auto;\n+      width: 100% !important;\n+      max-width: 8em;\n+      box-sizing: border-box;\n+    }\n+  ")),
  htmltools::div(
    class = "qc-summary-table",
  DT::datatable(
    calc_qc_summary(qc_all_dt, kept_dt, cuts_dt, qc_lu),
    rownames = FALSE,
    filter = "top",
    class = "compact stripe hover",
    options = list(
      pageLength = 20,
      lengthMenu = list(c(20, 50, 100), c("20", "50", "100")),
      searching = TRUE,
      ordering = TRUE,
      scrollX = TRUE,
      autoWidth = TRUE,
      dom = "frtip"
    )
  )
  )
)
```

## What were the reasons for excluding barcodes?

UpSet plots visualize the overlap between different exclusion criteria per sample. Plots are omitted for samples where no barcodes were excluded.

```{r plot_upsets, fig.height=4, fig.width=6, results='asis'}
cat('::: {.panel-tabset}\n\n')
for (bb in b_lvls) {
  qc_tmp <- qc_all_dt[batch_var == bb]
  if (nrow(qc_tmp) <= 1 || sum(qc_tmp$keep) == nrow(qc_tmp)) next
  cat("### ", bb, "\n\n")
  suppressMessages(print(plot_upset_of_exclusions(qc_tmp, qc_names, qc_lu, cuts_dt)))
  cat("\n\n")
}
cat(':::\n\n')
```

## Session info

<details>
```{r session_info, echo=TRUE}
sessionInfo()
```
</details>
