---
title: "Zoom diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
source("annotation_utils.R")
source("integration_plots.R")

zoom_name <- Sys.getenv("ZOOM_NAME", "zoom")
zoom_source <- Sys.getenv("ZOOM_SOURCE", "cluster")
zoom_values_str <- Sys.getenv("ZOOM_VALUES", "")
zoom_values <- trimws(strsplit(zoom_values_str, ",")[[1]])
zoom_values <- zoom_values[nzchar(zoom_values)]

marker_sel_res <- Sys.getenv("ZOOM_MARKER_SEL_RES", "0.2")
marker_top_n <- as.integer(Sys.getenv("ZOOM_MARKER_TOP_N", "10"))
marker_min_cpm <- as.numeric(Sys.getenv("ZOOM_MARKER_MIN_CPM", "50"))
marker_fdr_cut <- as.numeric(Sys.getenv("ZOOM_MARKER_FDR_CUT", "0.01"))
marker_max_zero_p <- as.numeric(Sys.getenv("ZOOM_MARKER_MAX_ZERO_P", "0.5"))
metadata_vars_str <- Sys.getenv("METADATA_VARS", "")
requested_metadata_vars <- trimws(strsplit(metadata_vars_str, "\\s+")[[1]])
requested_metadata_vars <- requested_metadata_vars[nzchar(requested_metadata_vars)]

qc_feature_palette <- c("#fff7ec", "#fdd49e", "#fdbb84", "#ef6548", "#990000")

plot_umap_qc_feature <- function(int_dt, var_name, label) {
  stopifnot(all(c("UMAP1", "UMAP2", var_name) %in% names(int_dt)))

  plot_dt <- copy(int_dt)[!is.na(get(var_name)), c("UMAP1", "UMAP2", var_name), with = FALSE]
  plot_dt[, UMAP1 := rescale(UMAP1, to = c(0.05, 0.95))]
  plot_dt[, UMAP2 := rescale(UMAP2, to = c(0.05, 0.95))]
  plot_dt <- plot_dt[sample(.N, .N)]

  ggplot(plot_dt, aes(x = UMAP1, y = UMAP2, colour = get(var_name))) +
    geom_point(size = 0.1) +
    scale_x_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    scale_y_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    scale_colour_gradientn(colours = qc_feature_palette, na.value = "grey85") +
    theme_bw() +
    theme(panel.grid = element_blank(), aspect.ratio = 1,
      legend.title.position = "left", legend.position = "bottom",
      axis.ticks = element_blank(), axis.text = element_blank()) +
    labs(colour = label)
}
```

```{r load_data}
int_dt <- fread("zoom_integration_dt.csv.gz")
qc_dt <- fread("zoom_qc_metrics.csv.gz")
marker_dt <- fread("zoom_marker_stats.csv.gz")
cpms_dt <- fread("zoom_marker_logcpms.csv.gz")
selection_dt <- fread("zoom_selection.csv.gz")

if (!"splice_frac" %in% names(qc_dt) && all(c("total_spliced", "total_unspliced") %in% names(qc_dt))) {
  qc_dt[, splice_frac := (total_spliced + 1) / (total_spliced + total_unspliced + 2)]
} else if (!"splice_frac" %in% names(qc_dt) && "logit_spliced" %in% names(qc_dt)) {
  qc_dt[, splice_frac := plogis(logit_spliced)]
}

qc_dt[, cell_id := paste(sample_id, cell_id, sep = "_")]

feature_sources <- list(
  list(key = "ncount_rna", source = "sum", label = "nCount RNA"),
  list(key = "nfeature_rna", source = "detected", label = "nFeature RNA"),
  list(key = "splice_frac", source = "splice_frac", label = "splice fraction"),
  list(key = "mito_pct", source = "mito_pct", label = "mitochondrial content")
)

present_leiden <- grep("^(RNA_snn_res\\.|leiden_)", names(int_dt), value = TRUE)
int_dt_pass2 <- int_dt[!is.na(UMAP1)]

feature_cols <- unique(unlist(lapply(feature_sources, `[[`, "source")))
present_feature_cols <- feature_cols[feature_cols %in% names(qc_dt)]
if (length(present_feature_cols) > 0) {
  feature_dt <- unique(qc_dt[, c("cell_id", present_feature_cols), with = FALSE], by = "cell_id")
  int_dt_pass2 <- merge(int_dt_pass2, feature_dt, by = "cell_id", all.x = TRUE)
}

core_cols <- c(
  "cell_id", "sample_id", "embedding", "UMAP1", "UMAP2",
  "dbl_UMAP1", "dbl_UMAP2", "dbl_cluster", "dbl_prop", "in_dbl_cl",
  "is_dbl", "is_doublet"
)
cluster_cols <- grep("^(RNA_snn_res\\.|leiden_)", names(int_dt), value = TRUE)
embed_cols <- grep("^(hmny_pca|pca)_", names(int_dt), value = TRUE)
metadata_vars <- setdiff(names(int_dt), c(core_cols, cluster_cols, embed_cols))
metadata_vars <- metadata_vars[!startsWith(metadata_vars, "_")]
metadata_vars <- metadata_vars[vapply(metadata_vars, function(var_name) {
  vals <- int_dt_pass2[[var_name]]
  any(!is.na(vals))
}, logical(1))]
if (length(requested_metadata_vars) > 0) {
  metadata_vars <- c(
    requested_metadata_vars[requested_metadata_vars %in% metadata_vars],
    setdiff(metadata_vars, requested_metadata_vars)
  )
}

umap_vars <- unique(c("sample_id", metadata_vars))
umap_vars <- umap_vars[umap_vars %in% names(int_dt_pass2)]
feature_vars <- Filter(function(spec) spec$source %in% names(int_dt_pass2), feature_sources)

marker_cl_col <- resolve_cluster_column(int_dt_pass2, paste0("RNA_snn_res.", marker_sel_res))
marker_top_dt <- marker_dt[str_detect(gene_type, "(lincRNA|lncRNA|pseudogene|antisense)", negate = TRUE)][
  logcpm.sel >= log(marker_min_cpm + 1)
] |> get_top_markers(fdr_cut = marker_fdr_cut, n_top = marker_top_n, max_zero_p = marker_max_zero_p)
marker_cl_ord <- order_cluster_labels(unique(marker_dt$cluster))
```

Zoom `r zoom_name` re-runs HVG selection and integration on a selected subset of cells from the broad analysis.

* **Selection source:** `r zoom_source`
* **Selected values:** `r if (length(zoom_values) > 0) paste(zoom_values, collapse = ", ") else "none"`
* **Cells retained:** `r format(nrow(int_dt_pass2), big.mark = ",")`
* **Samples retained:** `r uniqueN(int_dt_pass2$sample_id)`
* **Marker resolution:** `r marker_sel_res`

## Selected cells per sample

```{r selection_table}
sel_tbl <- selection_dt[, .(n_cells = .N), by = .(sample_id)][order(-n_cells)]
DT::datatable(sel_tbl, rownames = FALSE, class = "compact stripe hover",
  options = list(dom = "t", paging = FALSE, searching = FALSE, info = FALSE, ordering = TRUE, autoWidth = TRUE))
```

## Clusters over UMAP

```{r plot_clusters_integration, fig.height = 7, fig.width = 9, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    g_cl <- plot_umap_cluster(
      umap_dt = int_dt_pass2[, .(cell_id, UMAP1, UMAP2)],
      clust_dt = int_dt_pass2[, .(cell_id, cluster = get(cl_col))],
      name = sprintf('res = %s', res_val)
    )
    g_dens <- plot_umap_density(int_dt_pass2[, .(cell_id, UMAP1, UMAP2)])
    print(g_cl + g_dens)
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## UMAP by sample and metadata variables

```{r umap_metadata, fig.height = 5, fig.width = 7, results='asis'}
if (length(umap_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in umap_vars) {
    cat("### ", var_name, "\n\n")
    print(plot_umap_by_var(int_dt_pass2, var_name))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## UMAP by QC features

```{r umap_features, fig.height = 5, fig.width = 7, results='asis'}
if (length(feature_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (spec in feature_vars) {
    cat("### ", spec$label, "\n\n")
    print(plot_umap_qc_feature(int_dt_pass2, spec$source, spec$label))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## Top marker genes by cluster

```{r top_markers, fig.height = 9, fig.width = 10, results='asis'}
cat('::: {.panel-tabset}\n\n')
for (sel_cl in marker_cl_ord) {
  cat('###', sel_cl, '\n\n')
  if (sel_cl %in% marker_top_dt$cluster) {
    print(plot_top_marker_genes(sel_cl, marker_top_dt, cpms_dt, cl_order = marker_cl_ord))
  } else {
    cat('No marker genes passed the display filters for this cluster.\n\n')
  }
  cat('\n\n')
}
cat(':::\n\n')
```

## Cell counts per cluster

```{r cluster_counts}
tbl <- cluster_counts_tbl(int_dt_pass2, marker_cl_col)
knitr::kable(tbl, caption = sprintf('Cell counts at res %s', marker_sel_res))
```

## Session info

<details>
```{r session_info, echo=TRUE}
sessionInfo()
```
</details>