---
title: "Zoom diagnostics"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
source("annotation_utils.R")
source("integration_plots.R")

zoom_name <- Sys.getenv("ZOOM_NAME", "zoom")
zoom_source <- Sys.getenv("ZOOM_SOURCE", "cluster")
zoom_values_str <- Sys.getenv("ZOOM_VALUES", "")
zoom_values <- trimws(strsplit(zoom_values_str, ",")[[1]])
zoom_values <- zoom_values[nzchar(zoom_values)]

marker_sel_res <- Sys.getenv("ZOOM_MARKER_SEL_RES", "0.5")
marker_top_n <- as.integer(Sys.getenv("ZOOM_MARKER_TOP_N", "10"))
marker_top_n_plot <- min(marker_top_n, 5L)
marker_min_cpm <- as.numeric(Sys.getenv("ZOOM_MARKER_MIN_CPM", "50"))
marker_fdr_cut <- as.numeric(Sys.getenv("ZOOM_MARKER_FDR_CUT", "0.01"))
marker_max_zero_p <- as.numeric(Sys.getenv("ZOOM_MARKER_MAX_ZERO_P", "0.5"))
metadata_vars_str <- Sys.getenv("METADATA_VARS", "")
requested_metadata_vars <- trimws(strsplit(metadata_vars_str, "\\s+")[[1]])
requested_metadata_vars <- requested_metadata_vars[nzchar(requested_metadata_vars)]

qc_feature_palette <- c("#fff7ec", "#fdd49e", "#fdbb84", "#ef6548", "#990000")

plot_umap_qc_feature <- function(int_dt, var_name, label) {
  stopifnot(all(c("UMAP1", "UMAP2", var_name) %in% names(int_dt)))

  plot_dt <- copy(int_dt)[!is.na(get(var_name)), c("UMAP1", "UMAP2", var_name), with = FALSE]
  plot_dt[, UMAP1 := rescale(UMAP1, to = c(0.05, 0.95))]
  plot_dt[, UMAP2 := rescale(UMAP2, to = c(0.05, 0.95))]
  plot_dt <- plot_dt[sample(.N, .N)]

  ggplot(plot_dt, aes(x = UMAP1, y = UMAP2, colour = get(var_name))) +
    geom_point(size = 0.1) +
    scale_x_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    scale_y_continuous(breaks = pretty_breaks(), limits = c(0, 1)) +
    scale_colour_gradientn(colours = qc_feature_palette, na.value = "grey85") +
    theme_bw() +
    theme(panel.grid = element_blank(), aspect.ratio = 1,
      legend.title.position = "left", legend.position = "bottom",
      axis.ticks = element_blank(), axis.text = element_blank()) +
    labs(colour = label)
}
```

```{r load_data}
int_dt <- fread("zoom_integration_dt.csv.gz")
qc_dt <- fread("zoom_qc_metrics.csv.gz")
marker_dt <- fread("zoom_marker_stats.csv.gz")

cpms_dt <- fread("zoom_marker_logcpms.csv.gz")
selection_dt <- fread("zoom_selection.csv.gz")

# Path to cached marker expression RDS for top marker UMAPs
top_marker_expr_rds <- "zoom_marker_expression.rds"

parent_csv_path <- Sys.getenv("PARENT_INTEGRATION_CSV", "")
if (nzchar(parent_csv_path) && file.exists(parent_csv_path)) {
  parent_umap_dt <- fread(parent_csv_path, select = intersect(
    c("cell_id", "sample_id", "UMAP1", "UMAP2"),
    fread(parent_csv_path, nrows = 0L) |> names()
  ))
  parent_umap_dt <- parent_umap_dt[!is.na(UMAP1) & !is.na(UMAP2)]
} else {
  parent_umap_dt <- NULL
}

# Parent cluster-level annotation (cell_id, cluster, label, label_score) at the
# annotation resolution — used to highlight the SELECTED CLUSTERS on the parent UMAP.
parent_ann_path <- Sys.getenv("PARENT_ANNOTATION_CELL_LABELS", "")
if (nzchar(parent_ann_path) && file.exists(parent_ann_path) && basename(parent_ann_path) != "NO_FILE") {
  parent_ann_dt <- fread(parent_ann_path)
  parent_ann_dt[, cell_id := as.character(cell_id)]
  parent_ann_dt[, cluster := as.character(cluster)]
} else {
  parent_ann_dt <- NULL
}
annotation_sel_res <- Sys.getenv("ANNOTATION_SEL_RES", "0.2")

if (!"splice_frac" %in% names(qc_dt) && all(c("total_spliced", "total_unspliced") %in% names(qc_dt))) {
  qc_dt[, splice_frac := (total_spliced + 1) / (total_spliced + total_unspliced + 2)]
} else if (!"splice_frac" %in% names(qc_dt) && "logit_spliced" %in% names(qc_dt)) {
  qc_dt[, splice_frac := plogis(logit_spliced)]
}

qc_names <- c("log_counts", "log_feats", "logit_mito", "logit_spliced")
qc_lu <- c("no. of UMIs", "no. of genes", "mito pct.", "spliced pct.")
names(qc_lu) <- qc_names

qc_dt[, cell_id := paste(sample_id, cell_id, sep = "_")]

feature_sources <- list(
  list(key = "ncount_rna", source = "sum", label = "nCount RNA"),
  list(key = "nfeature_rna", source = "detected", label = "nFeature RNA"),
  list(key = "splice_frac", source = "splice_frac", label = "splice fraction"),
  list(key = "mito_pct", source = "mito_pct", label = "mitochondrial content")
)

present_leiden <- grep("^(RNA_snn_res\\.|leiden_)", names(int_dt), value = TRUE)
present_leiden <- present_leiden[order(as.numeric(sub("^(RNA_snn_res\\.|leiden_)", "", present_leiden)), decreasing = FALSE)]
int_dt_pass2 <- int_dt[!is.na(UMAP1)]

feature_cols <- unique(unlist(lapply(feature_sources, `[[`, "source")))
present_feature_cols <- feature_cols[feature_cols %in% names(qc_dt)]
if (length(present_feature_cols) > 0) {
  feature_dt <- unique(qc_dt[, c("cell_id", present_feature_cols), with = FALSE], by = "cell_id")
  int_dt_pass2 <- merge(int_dt_pass2, feature_dt, by = "cell_id", all.x = TRUE)
}

core_cols <- c(
  "cell_id", "sample_id", "embedding", "UMAP1", "UMAP2",
  "dbl_UMAP1", "dbl_UMAP2", "dbl_cluster", "dbl_prop", "in_dbl_cl",
  "is_dbl", "is_doublet"
)
cluster_cols <- grep("^(RNA_snn_res\\.|leiden_)", names(int_dt), value = TRUE)
embed_cols <- grep("^(hmny_pca|pca)_", names(int_dt), value = TRUE)
metadata_vars <- setdiff(names(int_dt), c(core_cols, cluster_cols, embed_cols))
metadata_vars <- metadata_vars[!startsWith(metadata_vars, "_")]
metadata_vars <- metadata_vars[vapply(metadata_vars, function(var_name) {
  vals <- int_dt_pass2[[var_name]]
  any(!is.na(vals))
}, logical(1))]
if (length(requested_metadata_vars) > 0) {
  metadata_vars <- c(
    requested_metadata_vars[requested_metadata_vars %in% metadata_vars],
    setdiff(metadata_vars, requested_metadata_vars)
  )
}

umap_vars <- unique(c("sample_id", metadata_vars))
umap_vars <- umap_vars[umap_vars %in% names(int_dt_pass2)]
metadata_umap_vars <- setdiff(umap_vars, "sample_id")
feature_vars <- Filter(function(spec) spec$source %in% names(int_dt_pass2), feature_sources)
sample_meta_dt <- if (length(metadata_vars) > 0) {
  unique(int_dt_pass2[, c("sample_id", metadata_vars), with = FALSE], by = "sample_id")
} else {
  unique(int_dt_pass2[, .(sample_id)])
}

# Integration-style binned metadata variables
bar_vars <- umap_vars[vapply(umap_vars, function(var_name) {
  vals <- int_dt_pass2[[var_name]]
  vals <- vals[!is.na(vals)]
  if (length(vals) == 0) {
    return(FALSE)
  }
  !is.numeric(vals) || uniqueN(vals) <= 12
}, logical(1))]
metadata_dens_vars <- setdiff(bar_vars, "sample_id")
metadata_dens_vars <- metadata_dens_vars[vapply(metadata_dens_vars, function(var_name) {
  vals <- int_dt_pass2[[var_name]]
  vals <- vals[!is.na(vals)]
  data.table::uniqueN(vals) > 1
}, logical(1))]
if (length(metadata_dens_vars) > 0) {
  sel_cl_int_dt <- copy(int_dt_pass2)[, c("cell_id", "UMAP1", "UMAP2", metadata_dens_vars), with = FALSE]
}

marker_cl_col <- resolve_cluster_column(int_dt_pass2, paste0("RNA_snn_res.", marker_sel_res))
marker_top_dt <- marker_dt[str_detect(gene_type, "(lincRNA|lncRNA|pseudogene|antisense)", negate = TRUE)][
  logcpm.sel >= log(marker_min_cpm + 1)
] |> get_top_markers(fdr_cut = marker_fdr_cut, n_top = marker_top_n_plot, max_zero_p = marker_max_zero_p)
marker_cl_ord <- order_cluster_labels(unique(marker_dt$cluster))
```

Zoom `r zoom_name` re-runs HVG selection and integration on a selected subset of cells from the broad analysis.

* **Selection source:** `r zoom_source`
* **Selected values:** `r if (length(zoom_values) > 0) paste(zoom_values, collapse = ", ") else "none"`
* **Cells retained:** `r format(nrow(int_dt_pass2), big.mark = ",")`
* **Samples retained:** `r uniqueN(int_dt_pass2$sample_id)`
* **Marker resolution:** `r marker_sel_res`

## Selected cells

The zoom subset is defined at the **cluster** level on the full-dataset analysis
(resolution `r annotation_sel_res`): whole clusters whose annotation matched the
target are taken, not individual cells. Below the **selected clusters** are highlighted
on the full-dataset UMAP (coloured, labelled with the cluster's annotation and mean
confidence score); every other cell is greyed out.

```{r selection_umap, fig.height = 5.5, fig.width = 6}
p1 <- NULL
if (!is.null(parent_umap_dt) && !is.null(parent_ann_dt)) {
  pdt <- merge(parent_umap_dt,
               parent_ann_dt[, .(cell_id, cluster, label, label_score)],
               by = "cell_id", all.x = TRUE)
  sel_cells    <- unique(selection_dt$cell_id)
  sel_clusters <- sort(unique(pdt[cell_id %in% sel_cells & !is.na(cluster), cluster]))

  pdt[, selected := !is.na(cluster) & cluster %in% sel_clusters]
  cl_cols <- setNames(rep(nice_cols, length.out = length(sel_clusters)), sel_clusters)

  cl_lab <- pdt[selected == TRUE, .(
    UMAP1 = median(UMAP1), UMAP2 = median(UMAP2),
    label = label[1], score = mean(label_score, na.rm = TRUE)
  ), by = cluster]
  cl_lab[, txt := sprintf("%s: %s (%.2f)", cluster, label, score)]

  p1 <- ggplot() +
    ggrastr::geom_point_rast(data = pdt[selected == FALSE],
                             aes(UMAP1, UMAP2), colour = "grey88", size = 0.08, raster.dpi = 300) +
    ggrastr::geom_point_rast(data = pdt[selected == TRUE],
                             aes(UMAP1, UMAP2, colour = cluster), size = 0.18, raster.dpi = 300) +
    scale_colour_manual(values = cl_cols, drop = FALSE, guide = "none") +
    ggrepel::geom_text_repel(data = cl_lab, aes(UMAP1, UMAP2, label = txt),
                             size = 3, seed = 1, min.segment.length = 0, max.overlaps = Inf,
                             box.padding = 0.4, point.padding = 0.2, segment.color = "grey55") +
    theme_bw() +
    theme(panel.grid = element_blank(), axis.ticks = element_blank(),
          axis.text = element_blank(), aspect.ratio = 1) +
    labs(title = sprintf("Selected clusters (res %s) over full-dataset UMAP", annotation_sel_res),
         x = "UMAP 1", y = "UMAP 2")
} else if (!is.null(parent_umap_dt)) {
  # Fallback (no parent annotation staged): selected cells over the parent UMAP
  ov <- merge(parent_umap_dt, unique(selection_dt[, .(cell_id, zoom_value)]), by = "cell_id", all.x = TRUE)
  ov[, sel := !is.na(zoom_value)]
  p1 <- ggplot() +
    ggrastr::geom_point_rast(data = ov[sel == FALSE], aes(UMAP1, UMAP2), colour = "grey88", size = 0.08, raster.dpi = 300) +
    ggrastr::geom_point_rast(data = ov[sel == TRUE],  aes(UMAP1, UMAP2), colour = "#c0392b", size = 0.15, raster.dpi = 300) +
    theme_bw() +
    theme(panel.grid = element_blank(), axis.ticks = element_blank(),
          axis.text = element_blank(), aspect.ratio = 1) +
    labs(title = "Selected cells over full-dataset UMAP", x = "UMAP 1", y = "UMAP 2")
}
if (!is.null(p1)) print(p1)
```

## Selected cells per sample


```{r selection_bars, fig.height = 14, fig.width = 7}
if ("condition" %in% names(sample_meta_dt)) {
  sel_tbl <- merge(sel_tbl, unique(sample_meta_dt[, .(sample_id, condition)]), by = "sample_id", all.x = TRUE, sort = FALSE)
  sel_tbl[, condition := factor(as.character(condition))]

  p2 <- ggplot(sel_tbl, aes(x = sample_id, y = n_cells, fill = condition)) +
    geom_col(width = 0.8, colour = "black", linewidth = 0.25) +
    geom_text(aes(label = scales::comma(n_cells)), angle = 90, hjust = -0.1, size = 2.7) +
    facet_wrap(~ condition, ncol = 1, scales = "free_x") +
    coord_cartesian(clip = "off") +
    scale_y_continuous(
      limits = function(x) c(0, max(x, na.rm = TRUE) * 1.2),
      expand = expansion(mult = c(0, 0.02))
    ) +
    theme_bw() +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      plot.margin = margin(5.5, 20, 5.5, 5.5),
      legend.position = "none"
    ) +
    labs(x = "sample", y = "selected cells", title = "Selected cells per sample")
} else {
  p2 <- ggplot(sel_tbl, aes(x = sample_id, y = n_cells)) +
    geom_col(width = 0.8, fill = "#4C78A8", colour = "black", linewidth = 0.25) +
    geom_text(aes(label = scales::comma(n_cells)), angle = 90, hjust = -0.1, size = 2.7) +
    coord_cartesian(clip = "off") +
    scale_y_continuous(
      limits = function(x) c(0, max(x, na.rm = TRUE) * 1.2),
      expand = expansion(mult = c(0, 0.02))
    ) +
    theme_bw() +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      plot.margin = margin(5.5, 20, 5.5, 5.5)
    ) +
    labs(x = "sample", y = "selected cells", title = "Selected cells per sample")
}

print(p2)
```

## Clusters over UMAP

```{r plot_clusters_integration, fig.height = 5, fig.width = 7, results='asis'}
if (length(present_leiden) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    g_cl <- plot_umap_cluster(
      umap_dt = int_dt_pass2[, .(cell_id, UMAP1, UMAP2)],
      clust_dt = int_dt_pass2[, .(cell_id, cluster = get(cl_col))],
      name = sprintf('res = %s', res_val)
    )+theme(legend.position = "none")
    g_dens <- plot_umap_density(int_dt_pass2[, .(cell_id, UMAP1, UMAP2)])
    print(g_cl + g_dens)
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## UMAP by sample and metadata variables

```{r umap_metadata, fig.height = 5, fig.width = 7, results='asis'}
if (length(umap_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in umap_vars) {
    cat("### ", var_name, "\n\n")
    print(plot_umap_by_var(int_dt_pass2, var_name))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## UMAP by QC features

```{r umap_features, fig.height = 5, fig.width = 7, results='asis'}
if (length(feature_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (spec in feature_vars) {
    cat("### ", spec$label, "\n\n")
    print(plot_umap_qc_feature(int_dt_pass2, spec$source, spec$label))
    cat("\n\n")
  }
  cat(':::\n\n')
}
```

## Check QC metrics of clusters

Distributions of QC metrics (library size, number of features, mitochondrial proportion, and spliced proportion) are shown for each cluster across different resolution values.

```{r plot_qc_distns, fig.height = 8, fig.width = 10, results='asis'}
present_qc_names <- qc_names[qc_names %in% names(qc_dt)]
if (length(present_leiden) > 0 && length(present_qc_names) > 0 && nrow(qc_dt) > 0) {
  qc_melt <- melt(qc_dt[, c('cell_id', present_qc_names), with = FALSE], id.vars = 'cell_id',
                  variable.name = 'qc_var', value.name = 'qc_val')
  qc_melt[, qc_full := qc_lu[qc_var]]
  qc_melt[, qc_var := factor(qc_var, levels = present_qc_names)]
  qc_melt[, qc_full := fct_reorder(qc_full, as.integer(qc_var))]
  qc_melt <- qc_melt[!is.na(qc_val)]

  cat('::: {.panel-tabset}\n\n')
  for (cl_col in present_leiden) {
    res_val <- sub('^RNA_snn_res\\.', '', cl_col)
    res_val <- sub('^leiden_', '', res_val)
    cat("### ", res_val, "\n\n")
    suppressWarnings(print(plot_cluster_qc_distns(
      qc_melt,
      clust_dt = int_dt_pass2[, .(cell_id, cluster = get(cl_col))],
      name = as.numeric(res_val)
    )))
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  cat("No QC metric distributions were available for cluster plots.\n\n")
}
```

## Barcharts

Stacked bar charts below show the composition of each cluster in the zoom integration. Each bar is normalised within a cluster, so the bar height always sums to 100%.

```{r cluster_barcharts, fig.height = max(4, 2.3 * length(present_leiden)), fig.width = 12, results='asis'}
if (length(present_leiden) > 0 && length(bar_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in bar_vars) {
    cat("### ", var_name, "\n\n")
    print(plot_cluster_composition_bars(int_dt_pass2, present_leiden, var_name))
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  cat("No categorical sample or metadata variables were available for stacked cluster bar charts.\n\n")
}
```


## Top marker genes by cluster at resolution `r marker_sel_res`

```{r top_markers, fig.height = 5, fig.width = 7, results = 'asis', cache = FALSE}
# Match annotation report: dotplot and UMAP facet per cluster
top_marker_expr_dt <- if (file.exists(top_marker_expr_rds)) readRDS(top_marker_expr_rds) else data.table()

cat('::: {.panel-tabset}\n\n')
for (sel_cl in marker_cl_ord) {
  cat(sprintf("### %s\n\n", sel_cl))
  has_markers <- sel_cl %in% as.character(marker_top_dt$cluster)
  if (!has_markers) {
    cat("No marker genes passed the display filters for this cluster.\n\n")
    next
  }

  p_dot <- plot_top_marker_genes(sel_cl, marker_top_dt, cpms_dt, cl_order = marker_cl_ord) +
    theme(
      axis.text.x = element_text(size = 9),
      axis.text.y = element_text(size = 9),
      strip.text = element_text(size = 9),
      plot.title = element_text(size = 12),legend.position="none"
    )
  g_umap <- NULL
  if (nrow(top_marker_expr_dt) > 0) {
    g_umap <- plot_top_marker_umap_facet(
      sel_cl, marker_top_dt, top_marker_expr_dt,
      int_dt_pass2[, .(cell_id, UMAP1, UMAP2)]
    ) +
      theme(
        strip.text = element_text(size = 9, margin = margin(b = 4)),
        plot.title = element_text(size = 11)
      )
  }

  print(p_dot)

  if (!is.null(g_umap)) {
    print(g_umap)
  }
  cat("\n\n")
  invisible(gc(verbose = FALSE))
}
rm(top_marker_expr_dt)
cat(':::\n\n')
```

## Cell counts per cluster

```{r cluster_counts}
tbl <- cluster_counts_tbl(int_dt_pass2, marker_cl_col)
knitr::kable(tbl, caption = sprintf('Cell counts at res %s', marker_sel_res))
```


## UMAP by binned metadata variables

```{r umap_metadata_binned, fig.height = 5, fig.width = 7, results='asis'}
if (length(metadata_dens_vars) > 0) {
  cat('::: {.panel-tabset}\n\n')
  for (var_name in metadata_dens_vars) {
    cat("### ", var_name, "\n\n")
    print(plot_clusters_annotated_by_densities(sel_cl_int_dt, var_name))
    cat("\n\n")
  }
  cat(':::\n\n')
} else {
  cat("No discrete metadata variables were retained in the integration output.\n\n")
}
```

## Session info

<details>
```{r session_info, echo=TRUE}
sessionInfo()
```
</details>